---
title: "Corn Spatial Project"
author: "Badri Khanal"
date: "9/16/2020"
output: html_document
---

---
title: "corn_spatial_1"
author: "Badri Khanal"
date: "8/26/2020"
output: 
  html_document:
    number_sections : yes
    theme: flatly
    toc: yes
  word document:
    toc : yes
editor_options: 
  chunk_output_type: inline
---
## Including Plots
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. packages loading
``` {r eval=TRUE, echo=FALSE}
library(sf) # to encode spatial vector data
library(sp) # Classes and methods for spatial data
library(GWmodel) # geographically-weighted (GW) models           
library(here) # Constructs paths to your project's files
library(data.table) # Fast aggregation of large data
library(ggplot2) # creating graphics, based on "The Grammar of Graphics"
library(viridis) # color scales  to make plots
library(dplyr) #A fast, consistent tool for working with data frame like objects
library("RColorBrewer")
library(microbenchmark)
```

# 2. Load data
```{r  , echo=FALSE}
mydata <- st_read("/Users/badrikhanal/Box/1. Analysis_folder_Dissertation/Spatial_project/OFPE/Bohnhoff_Tims_2019 Processed Data.gpkg") %>%
  mutate(S = tgts) %>% 
  mutate(N = n)%>%
  mutate(S2 = (S**2), N2 = (N**2),SN = (S*N),SN2 = (S*N^2),S2N = (S**2*N),
        S2N2 = (S**2 * N**2) ) %>% 
  cbind(., st_coordinates(st_centroid(.)))
  
#summary(mydata)
#mydata$S2N2
mydata2<- na.omit(mydata)
mydata_sp <- as(mydata2, 'Spatial')
mydata_sp_not_cleaned <- as(mydata, 'Spatial')

```

# 3. visualization 
```{r, echo=FALSE}
library("RColorBrewer")
map.na = list("SpatialPolygonsRescale", layout.north.arrow(),
              offset = c(329000, 261500), scale = 4000, col = 1)
map.scale.1 = list("SpatialPolygonsRescale", layout.scale.bar(),
                   offset = c(326500, 217000), scale = 5000, col = 1,
                   fill = c("transparent", "blue"))
map.scale.2 = list("sp.text", c(326500, 217900), "0", cex = 0.9, col = 1) 
map.scale.3 = list("sp.text", c(331500, 217900), "5km", cex = 0.9, col = 1) 
map.layout <- list(map.na, map.scale.1, map.scale.2, map.scale.3)
mypalette.1 <- brewer.pal(9, "Reds")
mypalette.2 <- brewer.pal(9, "Blues")
mypalette.3 <- brewer.pal(9, "Greens")
mypalette.4 <- brewer.pal(9, "Oranges")
mypalette.5 <- brewer.pal(9, "Purples")
mypalette.6 <- brewer.pal(9, "Spectral")
spplot(mydata_sp, "yield", key.space = "right",
       col.regions = mypalette.3, cuts = 5, #at = c( 4,6,8),
       sp.layout = map.layout,
       main = "yield selected plotting (basic)") 
spplot(mydata_sp_not_cleaned, "yield", key.space = "right",
       col.regions = mypalette.3, cuts = 5, #at = c( 4,6,8),
       sp.layout = map.layout,
       main = "yield all plotting (basic)") 

```

# 4. basic lm model
```{r}
# library(mgcv)
# #gam_result <- gam(yield~s(seed,k=3)+s(n,k=6)+s(X,k=6)+s(Y,k=6)+te(X,Y),data=mydata_sp)
# gam_result <- gam(yield~s(seed,k=3)+s(n,k=6)+s(X,k=6)+s(Y,k=6),data=mydata_sp)
# plot(gam_result)
# summary(gam_result)
lm_model1 <- lm("yield ~ S +  N ",data = mydata_sp)
summary(lm_model1)
lm_model2 <- lm("yield ~ S + S2 + N + N2 ",data = mydata_sp)
summary(lm_model2)
lm_model3 <- lm("yield ~ S + S2 + N + N2 +SN+ SN2+S2N +S2N2 ",data = mydata_sp)
summary(lm_model3)

```
# 5. Finding optimal bandwidth
```{r}

bw.gwr.1 <- bw.gwr(yield ~ S + N +S2 + N2 +
                  SN+ S2N + SN2 +S2N2, data = mydata_sp,
                  approach = "AICc", kernel = "bisquare", adaptive = TRUE)
bw.gwr.1 #bw =70, AIC =15618.68

bw.gwr.2 <- bw.gwr(yield ~ S + N +S2 + N2 +
                  SN+ S2N + SN2 +S2N2, data = mydata_sp,
                  approach = "AICc", kernel = "gaussian", adaptive = TRUE)
bw.gwr.2 #bw =16, AIC =15656.27

bw.gwr.3 <- bw.gwr(yield ~ S + N +S2 + N2 +
                  SN+ S2N + SN2 +S2N2, data = mydata_sp,
                  approach = "AICc", kernel = "exponential", adaptive = TRUE)
bw.gwr.3 #bw =16, AIC =15837.64

bw.gwr.4 <- bw.gwr(yield ~ S + N +S2 + N2 +
                  SN+ S2N + SN2 +S2N2, data = mydata_sp,
                  approach = "AICc", kernel = "tricube", adaptive = TRUE)
bw.gwr.4 #bw =70, AIC =15612.51

bw.gwr.5 <- bw.gwr(yield ~ S + N +S2 + N2 +
                  SN+ S2N + SN2 +S2N2, data = mydata_sp,
                  approach = "AICc", kernel = "boxcar", adaptive = TRUE)
bw.gwr.5 #bw =29, AIC =15733.19
```


# 6. model selection procedure
```{r, echo=FALSE}

DeVar <- "yield"
InDeVars <- c("S","N", "S2", "N2","SN",
             "SN2","S2N","S2N2" )

model.sel1 <- model.selection.gwr(DeVar, InDeVars, data = mydata_sp,
                                kernel = "bisquare", adaptive = TRUE, bw = 70)
# model.sel2 <- model.selection.gwr(DeVar, InDeVars, data = mydata_sp,
#                                  kernel = "gaussian", adaptive = TRUE, bw = 16)
# model.sel <- model.selection.gwr(DeVar, InDeVars, data = mydata_sp,
#                                  kernel = "exponential", adaptive = TRUE, bw = 16)
sorted.models <- model.sort.gwr(model.sel1, numVars = length(InDeVars),
                                ruler.vector = model.sel1[[2]][,2])
model.list <- sorted.models[[1]]
model.view.gwr(DeVar, InDeVars, model.list = model.list)
# circular view of 3 regressions from stepwise process,
#
plot(sorted.models[[2]][,2], col = "blue", pch = 20, lty = 5,
     main = "Alternative view of GWR model selection procedure",
     ylab = "AICc", xlab = "Model number", type = "b")
# AIC from model, AIC value continue to fall until both variables are included, this suggest
# that containing with bοth independent variable is worthwhile ( for specified bandwidth)
```



# 7. Run the GWR and MGWR models
```{r}

M1_corn_gwr <- gwr.basic( yield ~ S + N +S2+ N2 +SN+S2N +SN2+S2N2 , data = mydata_sp,
                            bw = bw.gwr.1, kernel = "bisquare", adaptive = TRUE) 

print(M1_corn_gwr)

M2_corn_mgwr <- gwr.mixed("yield ~ S + N +S2+ N2 +SN+S2N +SN2+S2N2", 
              data = mydata_sp, fixed.vars = c("S2","N2", "SN", "SN2", "S2N", "S2N2"),
              intercept.fixed = FALSE,
                           bw = bw.gwr.1, 
                           diagnostic = T,
                           kernel = "bisquare", 
                           adaptive = T,
                           longlat = F) 
print(M2_corn_mgwr)
# hist(M2_corn_mgwr$SDF$S_L, freq = FALSE, breaks = 100 )
# M2_corn_mgwr$global
# M2_corn_mgwr$local

save(M1_corn_gwr, file="M1_corn_gwr.rdata")
save(M2_corn_mgwr, file="M2_corn_mgwr.rdata")
load("M2_corn_mgwr.rdata")
load("M1_corn_gwr.rdata")
print(M1_corn_gwr)
print(M2_corn_mgwr)
```
#### ############ ######
# 8. optimization #######
#### ################### ###
```{r}
##########

#===================================
# Preparation
#===================================
#--------------------------
# packages and dataset
#--------------------------
#--- packages ---#
load("~/Dropbox/STRIPS_WTPAnalysis/Data/OFPE/M2_corn_mgwr.rdata")
library(sf)
library(sp)
library(data.table)
library(tidyverse)

#--- dataset ---#
coef_tbl <- st_as_sf(M2_corn_mgwr$SDF) %>% 
  mutate(id = 1:nrow(.)) 

coef_dt <- data.table(coef_tbl)

n_cells <- nrow(coef_dt)

#--------------------------
# Set parameters
#--------------------------
seed_range <- seq(20, 40, 0.2) ## SR
nitrogen_range <- seq(100, 270, 1) ## NR
c_price <- 3.5
n_price <- 0.4
s_price <- 1.5

#===================================
# Optimization (all the combinations)
#===================================
profit_data <- expand.grid(id = 1:nrow(coef_dt), S = seed_range, N = nitrogen_range) %>%
  data.table() %>%
  #--- merge the coef data to the evaluation points data ---#
  coef_dt[., on = "id"] %>% 
  .[order(id),] %>% 
  mutate(
    yield = Intercept_L + S_L * S + N_L * N + S2_F * S^2 + 
      N2_F * N^2 + SN_F * S * N + SN2_F * S * N^2 + S2N_F * S^2 * N + 
      S2N2_F * S^2 * N^2,
    revenue = yield * c_price,
    cost = S * s_price + N * n_price,
    profit = revenue - cost
  ) 

object.size(profit_data)

ggplot(profit_data[id == 1, ]) +
  geom_line(aes(y = yield, x = N))

optimal_SN <- profit_data %>%  
  setkey(id, profit) %>% 
  group_by(id) %>% 
  slice(n()) %>% 
  select(id, yield, revenue, cost, profit, S, N)

#--- merge the opt_NS data to the sf data ---#
sf_with_opt_NS <- left_join(coef_tbl, optimal_SN, by = "id")

#===================================
# Optimization (by id)
#===================================

get_opt_NS <- function(i) {

  profit_data <- expand.grid(S = seed_range, N = nitrogen_range) %>%
    data.table() %>%
    .[, id := i] %>% 
    #--- merge the coef data to the evaluation points data ---#
    coef_dt[., on = "id"] %>% 
    mutate(
      yield = Intercept_L + S_L * S + N_L * N + S2_F * S^2 + 
        N2_F * N^2 + SN_F * S * N + SN2_F * S * N^2 + S2N_F * S^2 * N + 
        S2N2_F * S^2 * N^2,
      revenue = yield * c_price,
      cost = S * s_price + N * n_price,
      profit = revenue - cost
    ) %>% 
    .[order(profit),] %>% 
    .[.N, ] %>% 
    .[, .(id, yield, revenue, cost, profit, S, N)]

  return(profit_data)

} 

library(parallel)

opt_NS_all <- mclapply(1:n_cells, get_opt_NS, mc.cores = 12) %>% 
  rbindlist()

#--- merge the opt_NS data to the sf data ---#
sf_with_opt_NS <- left_join(coef_tbl, opt_NS_all, by = "id")

#===================================
# Visualize
#===================================
ggplot(sf_with_opt_NS) +
  geom_sf(aes(fill = N)) +
  scale_fill_viridis_c()

ggplot(sf_with_opt_NS) +
  geom_sf(aes(fill = S)) +
  scale_fill_viridis_c()

ggplot(sf_with_opt_NS) +
  geom_sf(aes(fill = profit)) +
  scale_fill_viridis_c()

```

##### #################
# 9. Find profit maximizing  seed and nitrogen rates ()
#### ##################
```{r}
#
library(data.table)
ID <- beta_data1$ID
SR <- beta_data1$seed
NR <- beta_data1$nitrogen
Profit_values <- beta_data1$profit

group <- data.frame(Subject =ID,max_profit=Profit_values,seed_star =SR, n_star=NR )
group <- as.data.table(group)
group[group[, .I[max_profit ==max(max_profit)],by=Subject]$V1]

grid_profit_values <- group[group[, .I[max_profit ==max(max_profit)],by=Subject]$V1] %>% as.data.frame()

class(grid_profit_values)

print(grid_profit_values)

```
##### ################# ########
# 10. calculate optimal yield farm profit
##### ######### ###########

```{r}

#combine two dataframe
new_data2 <- cbind(Subject =mydata_sp$ID,beta_I,beta_S,beta_N,beta_SN,
                   beta_S2,beta_N2,beta_S2N,beta_SN2,beta_S2N2, 
                   corn_price,n_price,seed_price) %>% as.data.frame()

merged_data <- merge.data.frame(grid_profit_values,new_data2) %>% 
  mutate(yield_star = beta_I +
           beta_S* seed_star+ beta_N * n_star + 
           beta_SN* I(seed_star * n_star) +
           beta_S2* I(seed_star^2) + beta_N2 * I(n_star^2) +
           beta_S2N* I(seed_star^2 *n_star) + 
           beta_SN2 *I(seed_star *n_star^2) + 
           beta_S2N2 * I(seed_star^2 *n_star^2))
summary(merged_data)
# calculate farm profit
```

#### ###### #############
### 11. Randomize praire strips 
#### ############### #########
```{r}

```

#### ###### #############
### 12. Gaussian model All 
#### ############### #########

```{r}
M1_corn_gwr_gauss <- gwr.basic( yield ~ S + N +S2+ N2 +SN+S2N +SN2+S2N2 , data = mydata_sp,
                            bw = bw.gwr.2, kernel = "gaussian", adaptive = TRUE) 
M2_corn_mgwr_gauss <- gwr.mixed("yield ~ S + N +S2+ N2 +SN+S2N +SN2+S2N2", 
              data = mydata_sp, fixed.vars = c("S2","N2", "SN", "SN2", "S2N", "S2N2"),
              intercept.fixed = FALSE,
                           bw = bw.gwr.2, 
                           diagnostic = T,
                           kernel = "gaussian", 
                           adaptive = T, 
                           longlat = F) 
print(M2_corn_mgwr_gauss)
save(M1_corn_gwr_gauss, file="M1_corn_gwr_gauss.rdata")
save(M2_corn_mgwr_gauss, file="M2_corn_mgwr_gauss.rdata")
#load("M2_corn_mgwr.rdata")
#load(M1_corn_gwr.rdata")
print(M1_corn_gwr_gauss)
print(M2_corn_mgwr_gauss)

# optimization
beta_I_G <- M2_corn_mgwr_gauss$SDF$Intercept_L
beta_S_G <- M2_corn_mgwr_gauss$SDF$S_L
beta_N_G <- M2_corn_mgwr_gauss$SDF$N_L
beta_SN_G <- M2_corn_mgwr_gauss$SDF$SN_F
beta_S2_G <- M2_corn_mgwr_gauss$SDF$S2_F
beta_N2_G <- M2_corn_mgwr_gauss$SDF$N2_F
beta_S2N_G <- M2_corn_mgwr_gauss$SDF$S2N_F
beta_SN2_G <- M2_corn_mgwr_gauss$SDF$SN2_F
beta_S2N2_G <- M2_corn_mgwr_gauss$SDF$S2N2_F

ID_range <- mydata_sp$ID
seed_range <- seq(1, 40, 1) ## SR
nitrogen_range <- seq(200, 270, 1) ## NR
corn_price <- 3.5
n_price <- 0.4
seed_price <- 1.5

evaluation_point_data2 <- expand.grid(Id =ID_range, seed = seed_range, nitrogen = nitrogen_range) %>%
  data.frame() %>%
  mutate(
    N_2 = nitrogen^2,
    S_2 = seed^2,
    S_N = seed*nitrogen,
    S2_N = seed^2 *nitrogen,
    S_N2 = seed * nitrogen^2,
    S2_N2 = seed^2 *nitrogen^2) %>%
  mutate(yield = beta_I_G + beta_S *seed + beta_N_G * nitrogen + beta_SN_G * S_N +
      beta_S2_G * S_2 + beta_SN_G * S_N +  beta_S2N_G * S2_N + beta_SN2_G * S_N2 + beta_S2N2_G * S2_N2, 
      profit = corn_price*yield - seed_price * seed - n_price *nitrogen)

# evaluation_point_data$profit %>% max()

beta_data2 <- data.frame(ID= ID_range, beta_N = runif(5293760), beta_S = runif(5293760)) %>%
  .[rep(1:nrow(.), each = 1), ] %>%
  mutate(seed = rep(evaluation_point_data2$seed, 1)) %>% 
  mutate(nitrogen = rep(evaluation_point_data2$nitrogen, 1)) %>%
  mutate(profit = rep(evaluation_point_data2$profit, 1))

tail(beta_data2)  
#
library(data.table)
ID <- beta_data2$ID
SR <- beta_data2$seed
NR <- beta_data2$nitrogen
Profit_values <- beta_data2$profit

group <- data.frame(Subject =ID,max_profit=Profit_values,seed_star =SR, n_star=NR )
group <- as.data.table(group)
group[group[, .I[max_profit ==max(max_profit)],by=Subject]$V1]

grid_profit_values2 <- group[group[, .I[max_profit ==max(max_profit)],by=Subject]$V1] %>% as.data.frame()

class(grid_profit_values)

print(grid_profit_values)

#
#combine two dataframe
new_data3 <- cbind(Subject =mydata_sp$ID,beta_I_G,beta_S_G,beta_N_G,beta_SN_G,
                   beta_S2_G,beta_N2_G,beta_S2N_G,beta_SN2_G,beta_S2N2_G, 
                   corn_price,n_price,seed_price) %>% as.data.frame()

merged_data2 <- merge.data.frame(grid_profit_values2,new_data3) %>% 
  mutate(yield_star = beta_I +
           beta_S* seed_star+ beta_N * n_star + 
           beta_SN* I(seed_star * n_star) +
           beta_S2* I(seed_star^2) + beta_N2 * I(n_star^2) +
           beta_S2N* I(seed_star^2 *n_star) + 
           beta_SN2 *I(seed_star *n_star^2) + 
           beta_S2N2 * I(seed_star^2 *n_star^2))
summary(merged_data2)


```


