---
title: "Making Tables with `gt` and `modelsummary`"
author: "AECN 396/896-002"
output:
  xaringan::moon_reader:
    # css: [default, metropolis, metropolis-fonts] 
    css: xaringan-themer.css 
    lib_dir: libs
    nature:
      ratio: 16:10
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r, child = './../setup.Rmd'}
```

```{r, include = F, cache = F}
suppressMessages(library(nycflights13)) 
suppressMessages(library(gt)) 
suppressMessages(library(tidyverse)) 
```

# Before you start

## Learning objectives

The objectives of this chapter is to learn how to use the `ggplot2` package to create figures for effective communication 

## Table of contents

1. [`ggplot2` basics](#ggplot2-basics)
2. [Different types of figures](#dif-geoms)
3. [Placing more information in one figure](#more-info)
4. [Faceted figures](#faceting)
5. [Other supplementary `geom_*()`](#other-geoms)
6. [Make your figures presentable to others](#fine-tune)
7. [Tips](#tips)
7. [Gallery of other type of figures](#gallery)
7. [Animated figures](#animated)

<br>

<span style="color:red"> Tips: </span>hitting "o" key will give you a panel view of the slides


<!-- 
#=========================================
# gt Basics
#=========================================
-->

---

class: inverse, center, middle
name: more-info

# Create awesome tables with the `gt` package 

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

---

```{r include = F}
# Define the start and end dates for the data range
start_date <- "2010-06-02"
end_date <- "2010-06-15"

# The HTML decimal references for the black
# up- and down-pointing triangles are: #9650 and #9660;
# use an in-line style to apply color
up_arrow <- "<span style=\"color:green\">&#9650;</span>"
down_arrow <- "<span style=\"color:red\">&#9660;</span>"

# Create a gt table based on a preprocessed `sp500`
sp500_data <- sp500 %>%
  dplyr::filter(date >= start_date & date <= end_date) %>%
  dplyr::select(-adj_close) 
```


```{r taste-gt, cache = F, include = F}
sp500_data %>% #BREAK
  gt() %>% #BREAK
  tab_header(
    title = "S&P 500"
  ) %>% #BREAK
  fmt_date(
    columns = vars(date),
    date_style = 7
  ) %>% #BREAK
  fmt_currency(
    columns = vars(open, high, low, close),
    currency = "USD"
  ) %>% #BREAK
  fmt_number(
    columns = vars(volume),
    scale_by = 1 / 1E9,
    pattern = "{x}B"
  ) %>% #BREAK
  text_transform(
    locations = cells_body(
      columns = "close",
      rows = close > open),
    fn = function(x) paste(x, up_arrow)
  ) %>% #BREAK
  text_transform(
    locations = cells_body(
      columns = "close",
      rows = close < open),
    fn = function(x) paste(x, down_arrow)
  ) %>% #BREAK
  cols_label(
    date = "Date", open = "Open", 
    high = "High", low = "Low", 
    close = "Close", volume = "Volume"
  )
```

`r chunk_reveal(
  chunk_name = "taste-gt", 
  title = "# Taste of the gt package",
  break_type = "user", 
  display_type = "both", 
  left_assign = FALSE,
  width = c(50, 50)
  )
`
  
---

# Data preparation

.left4[
```{r results = "hide"}
#--- Define regions ---#
Australasia <- c("AU", "NZ")
Melanesia <- c("NC", "PG", "SB", "VU")
Polynesia <- c("PF", "WS", "TO", "TV")

#--- create a dataset ---#
tab_data <- countrypops %>%
  filter(country_code_2 %in% c(
    Australasia, Melanesia, Polynesia)
  ) %>%
  filter(year %in% c(1995, 2005, 2015)) %>%
  mutate(region = case_when(
    country_code_2 %in% Australasia ~ "Australasia",
    country_code_2 %in% Melanesia ~ "Melanesia",
    country_code_2 %in% Polynesia ~ "Polynesia",
  )) %>%
  pivot_wider(
    values_from = population, 
    names_from = year,
    names_prefix = "y_"
  ) %>%
  arrange(region, desc(y_2015)) %>%
  select(-starts_with("country_code")) %>% 
  mutate(
    pop_ratio_10_15 = y_2015/y_2005,
    date = "2013-11-14"
  )  

```
]

.right6[
```{r echo = F}
tab_data 
```
]

---

# Create or Modify Parts 

.panelset[ 

.panel[.panel-name[Instruction]

.left-full[

From the most basic table you can generate by applying `gt()` function to create a table, you can add parts and modify the style using various functions:

+ `tab_header()`: add a table header
+ `tab_spanner()`: add a spanner column label
+ `tab_spanner_delim()`: create column labels and spanners via delimited names
+ `tab_row_group()`: add a row group to a `gt` table
+ `tab_stubhead()`: add label text to the stubhead
+ .orange[`tab_footnote()`]: add a table footnote
+ `tab_source_note()`: add a source note citation
+ `tab_style()`: add custom styles to one or more cells
+ `tab_options()`: modify the table output options

<br>

.red[Note]: Inside these functions, you specify the details of the **actions** and **locations** where you take actions (unless the locations of the actions are fixed like `tab_header()`).


]

  ]

<!-- panel ends here -->

.panel[.panel-name[Ex 1]

This adds the title and subtitle to the table at the top of the table (thus no need to specify the location.)

```{r eval = F}
tab_header(
  #--- actions ---#
  title = "A", 
  subtitle = "B"
)
```

.left5[

Before:

```{r echo = F}
tab_data %>% 
  select(-date, - pop_ratio_10_15, - region) %>% 
  .[1:5, ] %>% 
  gt() 
```

]

.right5[

After:

```{r echo = F}
tab_data %>% 
  select(-date, - pop_ratio_10_15, - region) %>% 
  .[1:5, ] %>% 
  gt() %>% 
  tab_header(title = "A", subtitle = "B")
```

]

  ] 

<!-- panel ends here -->

.panel[.panel-name[Ex 2]

This adds a footnote label to the right of the `y_2015` column label and put the footnote text `"This is a footnote"` at the bottom of the table.

```{r eval = F}
tab_footnote(
  #--- action ---#
  footnote = "This is a footnote", 
  #--- location ---#
  locations = cells_column_labels(columns = vars(y_2015))
  )
```

.left5[

Before:

```{r echo = F}
tab_data %>% 
  select(-date, - pop_ratio_10_15, - region) %>% 
  .[1:5, ] %>% 
  gt() 
```

]

.right5[

After:

```{r echo = F}
tab_data %>% 
  select(-date, - pop_ratio_10_15, - region) %>% 
  .[1:5, ] %>% 
  gt() %>% 
  tab_footnote(
    #--- action ---#
    footnote = "This is a footnote", 
    #--- location ---#
    locations = cells_column_labels(columns = vars(y_2015))
  )
```

]

  ] 

<!-- panel ends here -->

.panel[.panel-name[Locations]

.left-full[

In specifying the locations of the table you modify, you can use `cells_*()` functions. Some of those are (see the **Helper Functions** section of [here](https://gt.rstudio.com/reference/index.html) for the full list.):    

+ `cells_title()`: targets the table title or the table subtitle depending on the value given to the groups argument ("title" or "subtitle").
+ `cells_column_labels()`: targets the column labels.
+ `cells_row_groups()`: targets the row group labels in any available row groups using the groups argument.
+ `cells_stub()`: targets row labels in the table stub using the rows argument.
+ `cells_body()`: targets data cells in the table body using intersections of columns and rows.

Each of these functions have their own ways of specifying where exactly you want to target.

]

  ]

<!-- panel ends here -->

.panel[.panel-name[Locations examples]

<span style="color:blue"> Example 1: </span>

```{r }
locations = cells_column_labels(columns = vars(y_2015)) 
```

This will target the `y_2015` column label.

<span style="color:blue"> Example 2: </span>

```{r }
locations = cells_title(groups = "title")
```

This will target the title of the table.

<span style="color:blue"> Example 3: </span>

```{r }
locations = cells_body(
  columns = starts_with("y_"),
  rows = y_2015 > 100000  
)
```

This will target the cells of the table body (the cells that have values) that  

+ are in the columns whose labels start with `y_` (`y_1995`, `y_2005`, and `y_2015`) 
+ are in the rows where the value of the cells are greater than 100000

  ] 

<!-- panel ends here -->

.panel[.panel-name[columns and rows]

Here are the options you have for `columns = ` and `rows = ` in specifying the target locations.

.blue[`columns`]:

+ explicitly name the columns with `vars(variable name 1, variable 2, ...)`
+ use helper functions like `starts_with()`, `everything()`, etc (you saw these in the data wrangling lecture slides!)

.blue[`rows`]:

+ use logical evaluation expressions like `variable <= values` (e.g., `rows = pop_ratio_10_15 >= 1.2`)
+ use helper functions

  ] 

<!-- panel ends here -->


]

<!-- panel set ends here -->

---

```{r gt-create, cache = F, include = F}
gt_1 <- tab_data %>% #BREAK
  gt(
    rowname_col = "country_name",
    groupname_col = "region"
  ) %>% #BREAK
  tab_header(
    title = "Populations of Oceania's Countries in 1995, 2005, and 2015",
    subtitle = md("`exibble` is an R dataset")
  ) %>% #BREAK 
  tab_spanner(
    label = "2000s",
    columns = vars(y_2005, y_2015)
  ) %>% #BREAK
  tab_stubhead(
    label = "Country/Region"
  ) %>% #BREAK 
  tab_footnote(
    footnote = "This is the last year",
    locations = cells_column_labels(
      columns = vars(y_2015)
    )
  ) %>% #BREAK
  tab_source_note(
    source_note = md("From the `gt` package")
  )  
```

`r chunk_reveal(
  chunk_name = "gt-create", 
  title = "# Demonstrations 1: Create Parts",
  break_type = "user", 
  display_type = "both", 
  left_assign = TRUE,
  width = c(45, 55)
  )
`

---

# `tab_style()` and `tab_options()`

.panelset[ 

.panel[.panel-name[tab_style()]

.left-full[

You can use `tab_style()` to add custom styles to one or more cells. It consists of two parts:

+ `style` where you specify the style you would like to add  
+ `location` where you you specify where in the table you would like to apply the `style`  

<br>

.blue[Example]:

```{r eval = F}
tab_style(
  style = list(
    cell_fill(color = "lightcyan"),
    ),
  locations = cells_body(
    columns = vars(pop_ratio_10_15),
    rows = pop_ratio_10_15 >= 1.2 # this is like `dplyr::filter()`
  )
) 
```

This will work on the cells with their values greater than 1.2 (`rows = pop_ratio_10_15 >= 1.2`) in the `pop_ratio_10_15` column (`columns = vars(pop_ratio_10_15)`) to make the cell fill color `lightcyan` (`cell_fill(color = "lightcyan")`)

]

  ]

<!-- panel ends here -->

.panel[.panel-name[style()]

.left-full[

You use a series of `cell_*()` functions to direct your modification requests to specific parts of the **cells** in `style()`:

+ cell_fill()
+ cell_text()
+ cell_body()
+ cell_borders()

Inside these `cell_*()` functions, you specify what you want to do to the parts of the cells that you refer to with `cell_*()` functions. 

Some of the options are

+ `color`
+ `size`
+ `width`
+ `alpha`

To see the full list of options available, run `?cell_*()` to see their individual help pages.

]

  ] 

<!-- panel ends here -->

.panel[.panel-name[tab_options()]

.left-full[

You can control the style (theme) of a `gt` table using the `tab_options()` function. 

You can control virtually any part of a table with it (see [here](https://gt.rstudio.com/reference/tab_options.html) for the complete options you have).

Here is a few of the options:

+ `table.border.bottom.width`
+ `column_labels.vlines.width`
+ `row_group.border.left.color` 

You can easily feel how much control you have over the style of a table by just looking at these options...

.red[Note]: Unlike the other functions we have looked at, these options specify the actions (e.g., change `width`) and locations (e.g., `table.border.bottom`).

<br>

.blue[Example]:

```{r eval = F}
tab_options(
  footnotes.marks = letters, 
  heading.subtitle.font.size = px(20)
)
```

+ The first one changes the footnote marks to letters (e.g., "a", "b", ..) from numbers (e.g., 1, 2, ...) 
+ The second one changes the font size of the subtitle to 20 px. 

]

  ] 

  <!-- panel ends here -->

]

<!-- panel set ends here -->

---

```{r gt-tab-styles, cache = F, include = F}
gt_2 <- gt_1 %>% #BREAK 
  tab_style(
    style = list(
      cell_fill(color = "red")
      ),
    locations = cells_body(
      columns = vars(pop_ratio_10_15),
      rows = pop_ratio_10_15 >= 1.2 
    )
  ) %>% #BREAK
  tab_style(
    style = list(
      cell_borders(sides = "top", style =  "dotted"),
      cell_text(
        weight = "bold", 
        style = "italic", 
        color = "#E50000"
      )
    ),
    locations = cells_body(
      columns = vars(y_1995),
      rows = y_1995 < 300000 
    )
  ) %>% #BREAK
  tab_style(
    style = list(
      cell_text(
        size = px(10), 
        align = "right", 
        v_align = "top",
        decorate = "underline"
        )
      ),
    locations = cells_body(columns = vars(y_2005))
  ) 
```

`r chunk_reveal(
  chunk_name = "gt-tab-styles", 
  title = "# Demonstrations 2: tab_style()",
  break_type = "user", 
  display_type = "both", 
  left_assign = TRUE,
  width = c(45, 55)
  )
`


---

```{r tab-options, cache = F, include = F}
gt_3 <- gt_2 %>% 
  tab_options(
    table.background.color = "lightcyan", #BREAK2
    footnotes.marks = letters, #BREAK3
    data_row.padding = px(5), #BREAK4
    heading.subtitle.font.size = px(20), #BREAK5
    source_notes.border.bottom.color ="#B2B2FF", #BREAK6
    row_group.background.color = "#8B008B", #BREAK7
    table.border.bottom.width = px(5), #BREAK8
    table_body.border.top.color = "green" #BREAK9
  )

```

`r chunk_reveal(
  chunk_name = "tab-options", 
  title = "# Demonstrations 3: tab_options()",
  break_type = "non_seq", 
  display_type = "both", 
  left_assign = TRUE,
  width = c(45, 55)
  )
`

---


# Formatting 

.panelset[ 

.panel[.panel-name[Instruction]

.left-full[

Columns of data can be formatted with the `fmt_*()` functions. Here are some of those functions (see [here](https://gt.rstudio.com/reference/index.html) for the complete list). 

+ `fmt_number()`: Format numeric values
+ `fmt_percent()`: Format values as a percentage
+ `fmt_currency()`: Format values as currencies
+ `fmt_date()`: Format values as dates
+ `fmt_markdown()`: Format Markdown text
+ `fmt_missing()`: Format missing values
+ `text_transform()`: Perform targeted text transformation with a function

<br>

Using these function, you can do things like

+ change the the number of decimal points
+ add currency marks in front of numbers
+ parse markdown texts 
+ transform missing values (NA) to the text of your choice
 
]

  ]

<!-- panel ends here -->

.panel[.panel-name[]

  ]

<!-- panel ends here -->

]

<!-- panel set ends here -->


---

```{r format, cache = F, include = F}
gt_4 <- gt_3 %>% #BREAK 
  fmt_number(
    columns = vars(y_1995, y_2005),
    decimals = 1,
    use_seps = TRUE
  ) %>% #BREAK
  fmt_percent(
    columns = vars(pop_ratio_10_15),
    row = y_1995 >= 300000,
    decimals = 1,
    use_seps = TRUE
  ) %>% #BREAK
  fmt_currency(
    columns = vars(y_2005),
    currency = "dollar"
  ) %>% #BREAK
  fmt_date(
    columns = "date",
    date_style = 3
  )
```

`r chunk_reveal(
  chunk_name = "format", 
  title = "# ",
  break_type = "user", 
  display_type = "both", 
  left_assign = TRUE,
  width = c(45, 55)
  )
`

---
  
# Modify Columns 
 

 
---
  

# Save

```{r eval = F}
gtsave(gt_4, "gt4.html") 
```

```{r eval = F}
gtsave(gt_4, "gt4.tex") 
```

```{r eval = F}
gtsave(gt_4, "LectureNotes/Chapter-6-MakeTable/gt4.pdf", zoom = 0.9, vwidth = 800) 
```
