---
title: "Sensor-guided Side-dressing Decision Making"
date: "`r Sys.Date()`" 
author:
  - Taro Mieno^[University of Nebraska Lincoln, tmieno2@unl.edu] and John Doe^[Random University, johndoe@email.com]
abstract: "Abstract: this research is so awesome that you cannot reject this paper."
output:
  bookdown::word_document2:
    reference_docx: word-style.docx 
bibliography: bibliography.bib
---

```{r echo = F}
library(knitr)

# === set wd ===#
# Notes: not really necessary to set the working directory as long as this Rmd file is part of an R project and the evaluation directory was set to "Project".

opts_knit$set(root.dir = "/Users/tmieno2/Dropbox/TeachingUNL/DataScience/LectureNotes/Chapter-9-WritingArticle")

options(htmltools.dir.version = FALSE)
options(knitr.duplicate.label = "allow")

opts_chunk$set(
  fig.align = "center", 
  fig.retina = 5,
  warning = F, 
  message = F,
  cache = T,
  echo = F
  )
```

```{r eval = F}
setwd("/Users/tmieno2/Dropbox/TeachingUNL/DataScience/LectureNotes/Chapter-9-WritingArticle")
```

```{r }
# === packages ===#
library(sf)
library(RColorBrewer)
library(patchwork)
library(viridis)
library(ggthemes)
library(flextable)
library(officer)
library(modelsummary)
library(tidyverse)

#--- source functions ---#
source("./Codes/parameters.R")
```

# Keywords: 

site-specific input management, nitrogen rate, seed rate, on-farm randomized field trial, shape-constrained generalized additive modeling, economically optimal input rates

# Introduction 

bluh bluh
 
@adrian2005producers examined bluh bluh.

bluh bluh [@shrader1966estimation] 

bluh bluh [@shrader1966estimation; @adrian2005producers] 

bluh bluh [-@shrader1966estimation] 


# Materials and Methods

```{r summary_statistics}
#--- number of observations by zone ---#
num_obs <- readRDS("reg_data.rds") %>%
  .[, Zone := case_when(
    tgt_comb == "36-124" ~ "Zone 3",
    tgt_comb == "34-134" ~ "Zone 2",
  )] %>%
  filter(!is.na(Zone)) %>% 
  group_by(Zone) %>% 
  summarize(n = n())

num_obs_z2 <- filter(num_obs, Zone == "Zone 2")$n
num_obs_z3 <- filter(num_obs, Zone == "Zone 3")$n
```

The number of observations are `r num_obs_z2` and `r num_obs_z3` for Zones 2 and 3, respectively. 

Table \@ref(tab:table-1) presents summary statistics by zone.

## Experimental Design and Data Collection

bluh bluh

## Econometric and Economic Methods

Here is the statistical model to estimate:

$Y_z = f_z(S) + g_z(N) + h_z(X,Y) + \varepsilon_z$,

# Results and Discussions

Table \@ref(tab:table-1) presents the regression results.

# Conclusions

bluh bluh

# Figures 

# Tables

```{r table-1, tab.cap = "Summary Statistics by Zone"}

#--- find observations ---#
# write the function that forces the number of observations to be presented as integers
get_N <- function(x) {
  length(x) %>% as.integer()
}

#--- create a summary table by zone ---#
summary_table <- readRDS("reg_data.rds") %>%
  .[, Zone := case_when(
    tgt_comb == "36-124" ~ "Zone 3",
    tgt_comb == "34-134" ~ "Zone 2",
  )] %>%
  .[!is.na(Zone), ] %>%
  datasummary(
    Zone * (
      (`Yield (kg/ha)` = Yield) +
        (`Nitrogen Rate (kg/ha)` = aa_n) +
        (`Seed Rate (1000/ha)` = aa_s)
    ) ~
    Format(digits = 0) * get_N + Mean + SD + Min + Max,
    data = .,
    output = "flextable"
  ) %>%
  autofit() %>%
  hline(
    i = 3,
    border = fp_border(width = 1)
  )

summary_table
```


# References

<div id="refs"></div>

