---
title: "Data visualization with `ggplot2`"
author: "AECN 396/896-002"
output:
  xaringan::moon_reader:
    # css: [default, metropolis, metropolis-fonts] 
    css: xaringan-themer.css 
    lib_dir: libs
    nature:
      ratio: 16:10
      highlightStyle: monokai
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup-knitr, include = FALSE}

library(knitr)
options(htmltools.dir.version = FALSE)
options(knitr.duplicate.label = "allow")

opts_chunk$set(
  fig.align = "center", 
  fig.retina = 5,
  fig.dim = c(4, 3.5), 
  out.width = "100%", 
  warning = F, 
  message = F,
  #fig.path='Figs/',
  cache = T
  #, echo=F, warning=F, message=F
  )

# setwd("~/Dropbox/TeachingUNL/DataScience/LectureNotes/Chapter-4-DataVisualization") 

library(flipbookr)
library(flair)
```

```{r xaringanExtra, echo=FALSE, cache = F}
library(xaringanExtra)
xaringanExtra::use_xaringan_extra(c("tile_view", "animate_css", "tachyons"))
xaringanExtra::use_tile_view()
xaringanExtra::use_panelset()
```

```{r xaringan-themer, include = FALSE, warning = FALSE, cache = F}
library(xaringanthemer)
style_duo_accent(
  header_h1_font_size = "2.1rem",
  header_h2_font_size = "1.5rem",
  code_inline_background_color = "#e6e6e6", 
  text_font_size = "0.7rem",
  code_font_size = "0.6rem",
  link_color = "#ff005a"
)
```

```{r libraries, include = F, cache = F}
#--- load packages ---#
suppressMessages(library(data.table))
suppressMessages(library(exactextractr))
suppressMessages(library(prism))
suppressMessages(library(sf))
suppressMessages(library(raster))
suppressMessages(library(DT))
suppressMessages(library(tictoc))
suppressMessages(library(tmap))
suppressMessages(library(extrafont))
suppressMessages(library(patchwork))
suppressMessages(library(plotly))
suppressMessages(library(ggthemes))
suppressMessages(library(RColorBrewer))
suppressMessages(library(parallel))
suppressMessages(library(maps))
suppressMessages(library(tidyverse))
````

```{css css-setup, echo = F}
.remark-code {
  display: block;
  overflow-x: auto;
  padding: .5em;
  background: #ffe7e7;
} 

.hljs-github .hljs {
  background: #f2f2fd;
}

.remark-inline-code {
  padding-top: 0px;
  padding-bottom: 0px;
  background-color: #e6e6e6;
}

.left-full {
  width: 80%;
  height: 92%;
  float: left;
}

.left-code {
  width: 38%;
  height: 92%;
  float: left;
}

.right-plot {
  width: 60%;
  float: right;
  padding-left: 1%;
}

.left-plot5 {
  width: 48%;
  height: 92%;
  float: left;
}

.right-plot5 {
  width: 48%;
  float: right;
  padding-left: 1%;
}

ul li{
  margin: 7px;
}

ul, li{
  margin-left: 9px; 
  padding-left: 0px; 
}

ol li{
  margin: 7px;
}

ol, li{
  margin-left: 9px; 
  padding-left: 0px; 
}
```


```{r dataset, results = "hide", include = F}
county_yield <- readRDS("county_yield.rds") %>% 
  select(soy_yield, corn_yield, year, county_code, state_name, d0_5_9, d1_5_9, d2_5_9, d3_5_9, d4_5_9) %>% 
    filter(state_name %in% c("Nebraska", "Kansas", "Colorado"))
```

<!-- 
#=========================================
# ggplot2 Basics
#=========================================
-->
  
---

class: inverse, center, middle
name: inputoutput

# `ggplot2` basics

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

---

```{r taste, cache = F, include = F}
county_yield %>% #BREAK
  ggplot(data = .) + #BREAK
  aes(x = factor(year), y = corn_yield) + #BREAK
  geom_boxplot(aes(fill = state_name)) + #BREAK
  facet_grid(state_name ~ .) + #BREAK
  theme(
    legend.position = "bottom", 
    legend.title = element_text(size = 6), 
    legend.text = element_text(size = 6), 
    axis.text.x = element_text(angle = 90, size = 6),
    axis.text.y = element_text(size = 6),
    axis.title = element_text(size = 8)
  ) + #BREAK
  xlab("Year") + #BREAK
  scale_y_continuous(
    name = "Corn Yield (bu/acre)", 
    limit = c(100, 200)
  ) +  #BREAK
  scale_fill_viridis_d(
    name = "State", 
    guide = guide_legend(
      title.position = "left"
    )
  ) #BREAK
```

`r chunk_reveal(
  chunk_name = "taste", 
  title = "# Taste of ggplot2",
  break_type = "user", 
  display_type = "both", 
  left_assign = FALSE,
  width = c(50, 50)
  )
`

---

```{r facet-wrap-one, cache = F, include = F}
ggplot(data = county_yield) + 
  geom_boxplot(
    aes(x = factor(year), y = corn_yield)
  ) + #BREAK
  facet_wrap(state_name ~ .) #BREAK
```

`r chunk_reveal(
  chunk_name = "facet-wrap-one", 
  title = "# ",
  break_type = "user", 
  display_type = "both", 
  left_assign = FALSE,
  width = c(50, 50)
  )
`

---

```{r echo = F}
#--- regression ---#
reg <- lm(corn_yield ~ d3_5_9, data = county_yield)

#--- find confidence interval ---#
min_d3 <- county_yield$d3_5_9 %>% min(na.rm = TRUE) # minimum d3 observed
max_d3 <- county_yield$d3_5_9 %>% max(na.rm = TRUE) # maximum d3 observed

eval_points <- data.frame(d3_5_9 = seq(min_d3, max_d3, length = 1000)) # evaluation points

ci_bound <- predict(reg, newdata = eval_points, interval = "confidence", level = 0.9) # upper and lower bound

ci_bound_data <- cbind(eval_points, ci_bound) # combine evaluation points and ci
```

```{r multiple-geoms, cache = F, include = F}
ggplot() + 
  geom_point(
    data = county_yield, 
    aes(y = corn_yield, x = d3_5_9)
  ) + #BREAK
  geom_line(
    data = ci_bound_data, 
    aes(x = d3_5_9, y = fit)
  ) + #BREAK
  geom_ribbon(
    data = ci_bound_data,
    aes(x = d3_5_9, ymin = lwr, ymax = upr),
    fill = "red",
    alpha = 0.4
  )
```

`r chunk_reveal(
  chunk_name = "multiple-geoms", 
  title = "# ",
  break_type = "user", 
  display_type = "both", 
  left_assign = FALSE,
  width = c(50, 50)
  )
`

---

```{r include = F}
county_yield_s_b2010 <- county_yield %>% 
  filter(year >= 2005, year <= 2010) 
```

```{r axes-content, cache = F, include = F}
county_yield_s_b2010 %>% #BREAK
  ggplot(data = .) + #BREAK
  aes(x = factor(year), y = corn_yield) + #BREAK
  geom_boxplot() + #BREAK
  aes(fill = state_name) + #BREAK
  scale_x_discrete( 
    name = "Year", 
    label = gsub("20", "", as.character(2000:2018)) 
  ) + #BREAK
  scale_y_continuous(
    name = "Corn Yield (bu/acre)", 
    limit = c(100, 200), 
    breaks = seq(100, 200, by = 10) 
  ) 
```    

`r chunk_reveal(
  chunk_name = "axes-content", 
  title = "# ",
  break_type = "user", 
  display_type = "both", 
  left_assign = FALSE,
  width = c(50, 50)
  )
`

---


```{r diverging, cache = F, include = F}
county_yield %>% #BREAK
  filter(state_name == "Nebraska") %>% #BREAK 
  filter(!is.na(corn_yield)) %>% 
  group_by(county_code) %>% 
  summarize(corn_yield = mean(corn_yield)) %>% #BREAK
  mutate(
    yield_norm = (corn_yield - mean(corn_yield))/sd(corn_yield),
    below_above = ifelse(yield_norm < 0, "below", "above")
  ) %>% #BREAK
  arrange(yield_norm) %>% #BREAK
  mutate(
    county_code_f = factor(county_code, levels = .$county_code)
  ) %>% #BREAK
  ggplot(data = .) + #BREAK
  geom_bar(
    aes(fill = below_above, x = county_code_f, y = yield_norm, label = county_code_f), 
    stat = 'identity', 
    width = 0.5
  )  + #BREAK
  scale_fill_manual(
    name = "Productivity", 
    labels = c("Above Average", "Below Average"), 
    values = c("above" = "#00ba38", "below" = "#f8766d")
  ) + #BREAK
  coord_flip() + #BREAK
  labs(x = "Normalized Yield", y = "County Code") + #BREAK
  theme(
    legend.position = "bottom",
    axis.text.y = element_text(size = 6)
  )
```

`r chunk_reveal(
  chunk_name = "diverging", 
  title = "# Diverging chart (variant of a bar chart)",
  break_type = "user", 
  display_type = "both", 
  left_assign = FALSE,
  width = c(50, 50)
  )
`

---


```{r pie-cheat, cache = F, include = F}
#--- define two parameters ---#
rpie = 1 # pie radius
rlabel = 0.6 * rpie # radius of the labels

#--- create a pie chart ---#
share_data %>% #BREAk
  ggplot(data = .) + 
    geom_arc_bar(
      aes(x0 = 0, y0 = 0, r0 = 0, r = rpie, # fixed values
        start = start_angle, 
        end = end_angle, 
        fill = state_name
      )
    ) + #BREAk
    geom_text(
      aes(
        x = rlabel * sin(mid_angle), 
        y = rlabel * cos(mid_angle), 
        label = state_name
      ), 
      hjust = 0.5, 
      vjust = 0.5
    ) + #BREAk
    coord_fixed() + #BREAk
    theme_void() + #BREAk
    scale_fill_viridis_d("State")

```

`r chunk_reveal(
  chunk_name = "pie-cheat", 
  title = "# Pie chart",
  break_type = "user", 
  display_type = "both", 
  left_assign = FALSE,
  width = c(50, 50)
  )
`

---

```{r mean-se, cache = F, include = F}
data(Salaries, package="carData")

Salaries %>%
  group_by(rank) %>% #BREAK
  summarize(
    n = n(),
    mean_salary = mean(salary),
    sd = sd(salary),
    se = sd / sqrt(n),
    ci = qt(0.975, df = n - 1) * sd / sqrt(n)
  ) %>% #BREAK
  ggplot(data = .) + #BREAK
  aes(x = rank, y = mean_salary) + #BREAK
  geom_point(size = 3) + #BREAK
  geom_errorbar(
    aes(
      ymin = mean_salary - 1.96 * se, 
      ymax = mean_salary + 1.96 * se
      ), 
    width = .1
  )

```

`r chunk_reveal(
  chunk_name = "mean-se", 
  title = "# Mean and Standard Error",
  break_type = "user", 
  display_type = "both", 
  left_assign = FALSE,
  width = c(50, 50)
  )
`
