---
title: "Data visualization with `ggplot2`"
author: "AECN 396/896-002"
output:
  xaringan::moon_reader:
    # css: [default, metropolis, metropolis-fonts] 
    css: xaringan-themer.css 
    lib_dir: libs
    nature:
      ratio: 16:10
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r, child = './../setup.Rmd'}
```

```{r, include = F, cache = F}
suppressMessages(library(nycflights13)) 
```

# Before you start

## Learning objectives

The objectives of this chapter is to learn how to use the `ggplot2` package to create figures for effective communication 

## Table of contents

1. [`ggplot2` basics](#ggplot2-basics)
2. [Different types of figures](#dif-geoms)
3. [Placing more information in one figure](#more-info)
4. [Faceted figures](#faceting)
5. [Other supplementary `geom_*()`](#other-geoms)
6. [Make your figures presentable to others](#fine-tune)
7. [Tips](#tips)
7. [Gallery of other type of figures](#gallery)
7. [Animated figures](#animated)

<br>

<span style="color:red"> Tips: </span>hitting "o" key will give you a panel view of the slides


---

#  `ggplot2` package

.left-full[

Install the package if you have not.

```{r ggplot2-install, eval = FALSE}
install.packages('ggplot2')
```

When you load the `tidyverse` package, it automatically load it. 

```{r tidyverse, message = TRUE}
#--- load ggplot2 along with others in the tidyverse package ---#
library(tidyverse)

#--- or ---#
library(ggplot2) #<<
```

]

---

# The datasets we use

.panelset[ 

.panel[.panel-name[Instruction]
Go [here](https://www.dropbox.com/sh/63rlp4ydmyjm1ui/AACYSeN0f_WAgKPQKzgpGVe0a?dl=0) and download **county_yield.rds** and then read the file onto R:
  ]

.panel[.panel-name[R Code]
```{r dataset, results = "hide", cache = F}
county_yield <- readRDS("county_yield.rds") %>% 
  select(soy_yield, corn_yield, year, county_code, state_name, d0_5_9, d1_5_9, d2_5_9, d3_5_9, d4_5_9) %>% 
    filter(state_name %in% c("Nebraska", "Kansas", "Colorado"))
```
  ]

.panel[.panel-name[Output]

```{r echo = F}
county_yield
```

  ]

.panel[.panel-name[Variable Definitions]
+ `soy_yield`: soybean yield (bu/acre)
+ `corn_yield`: corn yield (bu/acre)
+ `d0_5_9`: ratio of weeks under drought severity of 0 from May to September
+ `d1_5_9`: ~ drought severity of 1 from May to September   
+ `d2_5_9`: ~ drought severity of 2 from May to September   
+ `d3_5_9`: ~ drought severity of 3 from May to September   
+ `d4_5_9`: ~ drought severity of 4 from May to September   
  ]
]


<!-- 
#=========================================
# ggplot2 Basics
#=========================================
-->
  
---

class: inverse, center, middle
name: ggplot2-basics

# `ggplot2` basics

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

---

```{r taste, cache = F, include = F}
county_yield %>% #BREAK
  ggplot(data = .) + #BREAK
  aes(x = factor(year)) + #BREAK 
  aes(y = corn_yield) + #BREAK 
  geom_boxplot() + #BREAK 
  aes(fill = state_name) + #BREAK
  facet_grid(state_name ~ .) + #BREAK
  xlab("Year") + #BREAK
  ylab("Corn Yield (bu/acre)") + #BREAK
  ylim(c(100, 200)) + #BREAK
  scale_fill_viridis_d(
    name = "State", 
    guide = guide_legend(
      title.position = "left"
    )
  ) + #BREAK
  theme_bw() + #BREAK
  theme(axis.text.x = element_text(angle = 90, size = 6)) + #BREAK
  theme(axis.text.y = element_text(size = 6)) + #BREAK
  theme(legend.position = "bottom") + #BREAK 
  theme(
    legend.title = element_text(size = 6),
    legend.text = element_text(size = 6)
  ) + #BREAK
  labs(title = "Corn Yield (bu/acre) by State") + #BREAK
  labs(caption = "Design: Taro Mieno") + #BREAK
  labs(subtitle = "Data Source: USDA-NASS")

```

`r chunk_reveal(
  chunk_name = "taste", 
  title = "# Taste of ggplot2",
  break_type = "user", 
  display_type = "both", 
  left_assign = FALSE,
  width = c(50, 50)
  )
`

---

# Basics

.panelset[ 

.panel[.panel-name[Step 1]

.left-full[

The very first job you need to do to create a figure using the `ggplot2` package is to let R know the dataset you are trying to visualize, which can be done using `ggplot()` as follows:

```{r g_fig_base}
g_fig <- ggplot(data = county_yield)
```

When you create a figure using the `ggplot2` package, `ggplot()` is always the function you call first.

]

  ]

.panel[.panel-name[g_fig]

.left-code[
Let's now see what is inside `g_fig`:

```{r g-fig, eval = F}  
g_fig
```

Well, it's blank. Obviously, `g_fig` still does not have enough information to create any kind of figures. You have not told R anything specific about how you would like to use the information in the dataset.
]

.right-plot[
```{r ref.label = "g-fig", echo = F, out.width = "90%"}  

```
]

  ]

.panel[.panel-name[Step 2]

.left-full[

So, the next thing you need to do is tell `g_fig` what type of figure you want by `geom_*()` functions. 

For example, we use `geom_point()` to create a scatter plot. 

To create a scatter plot, R needs to know which variables should be on the y-axis and x-axis. 

These information can be passed to `g_fig` by the following code:

```{r g_fig_point}
g_fig_scatter <- g_fig + geom_point(aes(x = d3_5_9, y = corn_yield))
```

Here, 

+ `geom_point()` was added to `g_fig` to declare that you want a scatter plot 
+ `aes(x = d3_5_9, y = corn_yield)` inside `geom_point()` tells R that you want to create a scatter plot where you have `d3_5_9` on the x-axis and `corn_yield` on the y-axis

]

  ]

.panel[.panel-name[Output]

.left-code[

This is what `g_fig_scatter` looks:

```{r g-fig-scatter, eval = F}  
g_fig_scatter
```
]

.right-plot[
```{r scatter-plot, ref.label = "g-fig-scatter", echo = F, out.width = "90%"}  

```
]

  ]

.panel[.panel-name[`aes()`]

.left-full[
Going back to the code,

```{r }
g_fig_scatter <- g_fig + geom_point(aes(x = d3_5_9, y = corn_yield))
```

Note that `x = d3_5_9`, `y = corn_yield` are inside `aes()`. 

<span style="color:red"> Important:</span> `aes()` is used to make the <span style='color:red'>aes</span>thetic of the figure to be a function of variables in the dataset that you told `ggplot` to use (here, `county_yield`). 

`aes(x = d3_5_9, y = corn_yield)` is telling `ggplot` to use `d3_5_9` and `corn_yield` variables in the `county_yield` dataset for the x-axis and y-axis, respectively. 

If you do not have `x = d3_5_9`, `y = corn_yield` inside `aes()`, R is going to look for `d3_5_9` and `corn_yield` themselves (but not in `county_yield`), which you have not defined.

Try

```{r eval = F}
g_fig + geom_point(x = d3_5_9, y = corn_yield)
```
]


  ]

.panel[.panel-name[summary]

.left-full[
+ `ggplot(data = dataset)` to initiate the process of creating a figure

+ add `geom_*()` to declare what kind of figure you would like to make 

+ specify what variables in the dataset to use and how they are used inside `aes()` 

+ place the `aes()` you defined above in the `geom_*()` you specified above

]

  ]

]


<!-- 
#=========================================
# Different types of figures
#=========================================
-->
  
---

class: inverse, center, middle
name: dif-geoms

# Different types of figures

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

---

# Different types of figures

.panelset[ 

.panel[.panel-name[Figure types]

<br>

`ggplot2` lets you create lots of different kinds of figures via various `geom_*()` functions.

+ `geom_histogram()`/`geom_density()`
+ `geom_line()`
+ `geom_boxplot()`
+ `geom_bar()`

How to specify aesthetics vary by `geom_*()`.

  ]

.panel[.panel-name[Histogram]

.left-code[
```{r eval = F}  
g_fig + 
  geom_histogram(
    aes(x = corn_yield)
  )
```

`geom_histogram()` only needs `x`.
]

.right-plot[
```{r hist-ex, echo = F, out.width = "90%"}  
g_fig + 
  geom_histogram(
    aes(x = corn_yield)
  )
```
]

  ]

.panel[.panel-name[Density Plot]

.left-code[
```{r density_type, eval = F}  
g_fig + 
  geom_density(
    aes(x = corn_yield)
  )
```

`geom_density()` only needs `x`.
]

.right-plot[
```{r density-ex, ref.label = "density_type", echo = F, out.width = "90%"}  

```
]

  ]

.panel[.panel-name[Line plot]

.left-code[

Create a dataset first:

```{r }  
mean_yield <- county_yield %>% 
  group_by(year) %>% 
  summarize(
    corn_yield = mean(corn_yield, na.rm = TRUE)
  ) %>% 
  filter(!is.na(year))
```

Create a line plot:

```{r line_type, eval = F}
ggplot(data = mean_yield) + 
  geom_line(aes(x = year, y = corn_yield))

```

+ `geom_line()` needs `x` and `y`.

]

.right-plot[
```{r line-ex, ref.label = "line_type", echo = F, out.width = "90%"}  
```
]

  ]

.panel[.panel-name[Boxplot]

.left-code[
```{r boxplot_type, eval = F}  
ggplot(data = county_yield) + 
  geom_boxplot(
    aes(x = factor(year), y = corn_yield)
  )
```

+ `geom_boxplot()` needs `x` and `y`

]

.right-plot[
```{r box-ex, ref.label = "boxplot_type", echo = F, out.width = "90%"}  

```
]
  ]

.panel[.panel-name[Bar plot]

.left-code[
```{r barplot_type, eval = F}  
ggplot(data = mean_yield) + 
  geom_bar(
    aes(
      x = year, 
      y = corn_yield
    ), 
    stat = "identity"
  )
```

+ `geom_bar()` needs `x` and `y`

]

.right-plot[
```{r bar-ex, ref.label = "barplot_type", echo = F, out.width = "90%"}  

```
]
  ]

]

---

# Modifying how figures look

.left-full[

All the elements in the figures we have created so far are in black and white.

You can change how figure elements look by providing options inside `geom_*()`.

Here are the list of options to control the aesthetics of figures: 

+ fill
+ color
+ size
+ shape
+ linetype

Elements of figures that you can modify differ by `geom` types

The same element name can mean different things based on `geom` types

]

---

```{r fig-scatter-f, cache = F, include = F}
g_fig + #BREAK
  geom_point( 
    aes(x = d3_5_9, y = corn_yield), 
    color = "red", #BREAK2
    size = 0.7, #BREAK3
    shape = 0 #BREAK4
  )
```

`r chunk_reveal(
  chunk_name = "fig-scatter-f", 
  title = "# Scatter Plot",
  break_type = "non_seq", 
  display_type = "both", 
  left_assign = FALSE,
  width = c(50, 50)
  )
`

+ `color = "red"`: makes all the squares red
+ `size = 0.5`: makes the size of the squares smaller 
+ `shape = 0`: change the shape of the points (find other shapes [here](http://www.sthda.com/english/wiki/ggplot2-point-shapes))

---

```{r hist-f, cache = F, include = F}  
g_fig + #BREAK
  geom_histogram(
    aes(x = corn_yield), 
    color = "blue", #BREAK2
    fill = "green", #BREAK3
    size = 2, #BREAK4 
    shape = 2 #BREAK5 
  )
```

`r chunk_reveal(
  chunk_name = "hist-f", 
  title = "# Histogram",
  break_type = "non_seq", 
  display_type = "both", 
  left_assign = FALSE,
  width = c(50, 50)
  )
`
+ `color = "blue"`: makes all the boundary of the bars blue
+ `fill = "green"`: makes the fill of the bars green
+ `size = 2`: makes the line width of the boundary of the bars thicker 
+ `shape = 2`: does nothing
 

---


```{r box-f, cache = F, include = F}
ggplot(data = county_yield) + #BREAK
  geom_boxplot(
    aes(x = factor(year), y = corn_yield),
    color = "red", #BREAK2
    fill = "orange", #BREAK3
    size = 0.2, #BREAK4
    shape = 1 #BREAK5
    )
```

`r chunk_reveal(
  chunk_name = "box-f", 
  title = "# Box Plot",
  break_type = "non_seq", 
  display_type = "both", 
  left_assign = FALSE,
  width = c(50, 50)
  )
`

+ `color = "blue"`: makes all the boundary of the boxes red
+ `fill = "orange"`: makes the fill of the boxes orange
+ `size = 0.2`: makes the line width of the boundary of the boxes thinner 
+ `shape = 1`: does nothing

---

```{r line-f, cache = F, include = F}
ggplot(data = mean_yield) + #BREAK1
  geom_line(
    aes(x = year, y = corn_yield), 
    color = "blue", #BREAK2
    size = 1.5, #BREAK3
    fill = "red", #BREAK4  
    linetype = "dotted", #BREAK5
  )
```

`r chunk_reveal(
  chunk_name = "line-f", 
  title = "# Line Plot",
  break_type = "non_seq", 
  display_type = "both", 
  left_assign = FALSE,
  width = c(50, 50)
  )
`

+ `color = "blue"`: makes the line blue
+ `size = 1.5`: makes the line thicker 
+ `fill = "red"`: does nothing
+ `linetype = "dotted"`: makes the line dotted

---

# Exercises

.panelset[ 

.panel[.panel-name[Instruction]

This exercise use the `diamonds` dataset from the `ggplot2()` package. First, load the dataset and extract observations with `Premium` cut whose color is one of `E`, `I`, and `F`:

```{r diamond, cache=TRUE}
data('diamonds')
premium <- diamonds %>%
  filter(
    cut=='Premium' & color %in% c('E','I','F')
    )

#--- take a look ---#
premium
```

  ]

.panel[.panel-name[Exercise 1]

<br>

Using `carat` and `price` variables from `premium`, generate the figure below:

```{r diamond_fig_1, echo = FALSE, cache = TRUE, out.width = "50%"}
ggplot(data=premium) +
  geom_point(aes(x = carat, y = price), color = 'red')
```

  ]

.panel[.panel-name[Exercise 2]

<br>

Using `price` variables from `premium`, generate a histogram of `price` shown below:

```{r diamond_fig_2, echo = FALSE, cache = TRUE, out.width = "50%"}
ggplot(data=premium) +
  geom_histogram(aes(x = price), fill = 'white', color = 'blue')
```

  ]

]


<!-- 
#=========================================
# Placing more information
#=========================================
-->
  
---

class: inverse, center, middle
name: more-info

# Placing more information in one figure

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

---

# Placing more information in one figure 

.panelset[ 

.panel[.panel-name[Motivation]

<br>

So far, we have learned how to create popular types of figures. 

We can make a figure much more informative by making its aesthetics data-dependent.

For example, suppose you are interested in comparing the history of irrigated corn yield by state in a line plot.

So, you want to create a line for each state and make the lines distinguishable so the readers know which line is for which state like this:   

   
```{r more-info, echo = F, out.width = "60%", fig.dim = c(6, 3)}
county_yield_mean <- county_yield %>% 
  group_by(state_name, year) %>% 
  summarize(corn_yield = mean(corn_yield, na.rm = T))

ggplot(data = county_yield_mean) +
  geom_line(
    aes(y = corn_yield, x = year, color = state_name)
  )  
```

  ]

.panel[.panel-name[How]

.left-full[

We can make the aesthetics of a figure data-dependent by specifying which variable you use for aesthetics differentiation <span style="color:red"> INSIDE </span>`aes()`.

Here is an example:

```{r how-dif, include = F}
ggplot(data = county_yield_mean) +
  geom_line(
    aes(y = corn_yield, x = year, color = state_name)
  )   
```

```{r how-dif-flair, echo = F}
decorate("how-dif", eval = F) %>% 
  flair("color = state_name") %>% 
  knit_print.with_flair()
```

In this code, `color = state_name` is inside `aes()` and it tells R to divide the data into the groups of State and draw a line by `state_name` (by state) where the lines are color-differentiated.

A legend is automatically generated.

]

  ]

.panel[.panel-name[Let's do it]

<br>

.left-code[

Create a data set of corn yield by state-year first:

```{r }  
county_yield_mean <- county_yield %>% 
  group_by(state_name, year) %>% 
  summarize(corn_yield = mean(corn_yield, na.rm = T))
```

Create a plot:

```{r corn-color-dif, eval = F}
ggplot(data = county_yield_mean) +
  geom_line(
    aes(
      y = corn_yield, 
      x = year, 
      color = state_name #<<
    )
  )
```

]

.right-plot[
```{r do-it, ref.label = "corn-color-dif", echo = F, out.width = "90%"}  

```
]

  ]

]

---


```{r density-more, cache = F, include = F}
ggplot(data = county_yield_mean) + #BREAK
  geom_density(
    aes(
      x = corn_yield, 
      fill = state_name #BREAK2 
    ),
    alpha = 0.3
  )
```

`r chunk_reveal(
  chunk_name = "density-more", 
  title = "# More examples: Density Plot",
  break_type = "non_seq", 
  display_type = "both", 
  left_assign = FALSE,
  width = c(50, 50)
  )
`

---


```{r box-more, cache = F, include = F}
county_yield %>% 
  filter(state_name %in% c("Nebraska", "Kansas")) %>% 
  ggplot(data = .) + 
  geom_boxplot(
    aes(
      x = factor(year), 
      y = corn_yield,  
      fill = state_name #BREAK2
    )
  )

```

`r chunk_reveal(
  chunk_name = "box-more", 
  title = "# More examples: Boxplot",
  break_type = "non_seq", 
  display_type = "both", 
  left_assign = FALSE,
  width = c(50, 50)
  )
`

---


```{r scatter-more, cache = F, include = F}
county_yield %>% 
  filter(state_name %in% c("Nebraska", "Kansas")) %>%
  ggplot(data = .) + #BREAK
  geom_point(
    aes(
      x = d3_5_9, 
      y = corn_yield, 
      color = state_name, #BREAK2 
      shape = state_name #BREAK3 
    ),
    size = 0.7
  )

```

`r chunk_reveal(
  chunk_name = "scatter-more", 
  title = "# More examples: Scatter Plot",
  break_type = "non_seq", 
  display_type = "both", 
  left_assign = FALSE,
  width = c(50, 50)
  )
`

---

# Exercises

.panelset[ 


.panel[.panel-name[Exercise 1]

<br>

Using `premium`, create a scatter plot of `price` (y-axis) against `depth` (x-axis) by `clarity` as shown below:

```{r ex_2_1, echo = FALSE, cache = TRUE, out.width = "60%", fig.dim = c(6, 4)}
ggplot(data = premium) +
  geom_point(aes(y = price, x = depth, color = clarity))
```

  ]

.panel[.panel-name[Exercise 2]

<br>

Using `premium`, create density plots of `carat` by `color` as shown below (set `alpha` to 0.5):

```{r ex_2_2, echo = FALSE, cache = TRUE, out.width  =  "60%", fig.dim = c(6, 4)}
ggplot(data = premium) +
  geom_density(aes(x = carat, fill = color), alpha = 0.5)
```

  ]

]

<!-- 
#=========================================
# Faceting
#=========================================
-->
  
---

class: inverse, center, middle
name: faceting

# Faceting  

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

---

# Faceting: Basics

.panelset[ 


.panel[.panel-name[Motivation]

Sometimes, you would like to visualize information across groups on separate panels. 

.left5[

Too much information in one panel?

```{r box_all, echo = F, cache=TRUE}
ggplot(data = county_yield) + 
  geom_boxplot(
    aes(
      x = factor(year), 
      y = corn_yield, 
      fill = state_name
    )
  )
```
]

.right5[

On separate panels (faceting)?

```{r box-faceted-1, echo = F, cache = TRUE}  
ggplot(data = county_yield) + 
  geom_boxplot(
    aes(
      x = factor(year), 
      y = corn_yield 
    )
  ) +
  facet_wrap(state_name ~ ., nrow = 3)
```
]

  ]

.panel[.panel-name[How]

.left-full[

We can make faceted figures by adding either `facet_wrap` or `facet_grid()` in which you specify which variable to use for faceting. 

Here is an example:

```{r how-facet, include = F}
ggplot(data = county_yield) + 
  geom_boxplot(
    aes(x = factor(year), y = corn_yield)
  ) +
  facet_wrap(state_name ~ .)
```

```{r echo = F}
decorate("how-facet", eval = F) %>% 
  flair("facet_wrap(state_name ~ .)") %>% 
  knit_print.with_flair()
```

In this code, `facet_wrap(state_name ~ .)` is added to a simple boxplot, which tells R to make a boxplot by `state_name` (state).

What does `~ .` do?

]

  ]

]

---


```{r facet-ex, cache = F, include = F}
ggplot(data = county_yield) + 
  geom_boxplot(
    aes(x = factor(year), y = corn_yield)
  ) + #BREAK
  facet_wrap(state_name ~ .) 
```

`r chunk_reveal(
  chunk_name = "facet-ex", 
  title = "# Faceting: an Example",
  break_type = "user", 
  display_type = "both", 
  left_assign = FALSE,
  width = c(50, 50)
  )
`

---

# Faceting: Two-way

.panelset[ 

.panel[.panel-name[two-way faceting]

.left-full[
Two-way faceting will 

+ divide the data into groups where each group has a unique combination of the two faceting variables  

+ create a plot for each group

```{r two-facet, eval = F}
ggplot(data = county_yield) + 
  geom_histogram(
    aes(x = corn_yield)
  ) +
  facet_wrap(state_name ~ year) #<<
```

This code will create a histogram of corn yield for each of the unique state-year combination.

]


 ]

.panel[.panel-name[Figure 2]

.left-code[

Filter `county_yield` to those in 2017 and 2018.

```{r }
county_yield_s <- county_yield %>% 
  filter(year %in% c(2017, 2018))
```

Create a faceted density plots.

```{r two-ex, eval = F}  
ggplot(data = county_yield_s) + 
  geom_histogram(
    aes(x = corn_yield)
  ) +
  facet_wrap(state_name ~ year)  
```
]

.right-plot[
```{r two-ex-fig, ref.label = "two-ex", echo = F, out.width = "90%"}  

```
]

  ]

]

---


# Faceting with `facet_grid()`

.panelset[ 

.panel[.panel-name[compare]

.left5[

`facet_wrap`

```{r wrap-ex-1, fig.dim = c(4,3)}
ggplot(data = county_yield_s) + 
  geom_histogram(aes(x = corn_yield)) +
  facet_wrap(state_name ~ year)  
```
]

.right5[

`facet_grid`

```{r frig-ex-1, fig.dim = c(4,3)}
ggplot(data = county_yield_s) + 
  geom_histogram(aes(x = corn_yield)) +
  facet_grid(state_name ~ year)  
```
]


  ]

.panel[.panel-name[facet_grid()]

.left-code[

Unlike `facet_wrap()`, which side you put faceting variables matters a lot. 

+ left hand side: rows 
+ right hand side: columns 

```{r grid-ex-2, eval = F}  
ggplot(data = county_yield_s) + 
  geom_histogram(aes(x = corn_yield)) +
  facet_grid(state_name ~ year)
```

In the code above, `state_name` values become the rows, and `year` values become columns.
]

.right-plot[
```{r grid-matter, ref.label = "grid-ex-2", echo = F, out.width = "90%", out.width = "90%"}  

```
]

  ]

.panel[.panel-name[order]

.left5[
```{r wrap-left, ref.label = "grid-ex-2"}
  
```
]

.right5[
```{r grid-right}
ggplot(data = county_yield_s) + 
  geom_histogram(aes(x = corn_yield)) +
  facet_grid(year ~ state_name) 
```
]


  ] <!-- panel ends here -->

.panel[.panel-name[scale]

.left-full[

`facet_grid()` allows 

+ the figures in different columns to have different scales for the x-axis (figures in the same column have the same scale for the x-axis)

+ the figures in different rows to have different scales for the y-axis (figures in the same rows have the same scale for the x-axis)

]

  ] <!-- panel ends here -->

.panel[.panel-name[free x]

.left-code[
```{r free-x, eval = F}  
ggplot(data = county_yield_s) + 
  geom_histogram(aes(x = corn_yield)) +
  facet_grid(
    state_name ~ year, 
    scales = "free_x" #<<
  )
```

]

.right-plot[
```{r free-x-fig, ref.label = "free-x", echo = F, out.width = "90%"}  

```
]

  ] <!-- panel ends here -->

.panel[.panel-name[free y]

.left-code[
```{r free-y, eval = F}  
ggplot(data = county_yield_s) + 
  geom_histogram(aes(x = corn_yield)) +
  facet_grid(
    state_name ~ year, 
    scales = "free_y" #<<
  )  
```

]

.right-plot[
```{r free-y-fig, ref.label = "free-y", echo = F, out.width = "90%"}  

```
]


  ] <!-- panel ends here -->

.panel[.panel-name[both free]

.left-code[
```{r free, eval = F}  
ggplot(data = county_yield_s) + 
  geom_histogram(aes(x = corn_yield)) +
  facet_grid(
    state_name ~ year, 
    scales = "free" #<<
  )  
```

]

.right-plot[
```{r both-free-fig, ref.label = "free", echo = F, out.width = "90%"}  

```
]

  ] <!-- panel ends here -->

]

---

# Exercises

.panelset[ 

.panel[.panel-name[Exercise 1]

<br>

Using `premium`, create scatter plots of `price` (y-axis) against `carat` (x-axis) by `color` on separate panels as shown below:

```{r ex_3_1, echo = FALSE, cache = TRUE, out.width = "50%"}
ggplot(data=premium) +
  geom_point(aes(x=carat,y=price)) +
  facet_grid(color~.)
```

  ]

.panel[.panel-name[Exercise 2]

<br>

Using premium, create histogram of `carat` by `color` and `clarity` on separate panels as shown below:

```{r ex_3_2, echo = FALSE, cache = TRUE, out.width = "50%"}
ggplot(data=premium) +
  geom_histogram(aes(x=carat)) +
  facet_grid(color~clarity)
```

  ]

]

<!-- 
#=========================================
# Other useful geom_* 
#=========================================
-->
  
---

class: inverse, center, middle
name: other-geoms

# Other supplementary `geom_*()`  

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

---

# Other supplementary `geom_*()`  

.panelset[ 

.panel[.panel-name[geom_*]

.left-full[

Here are the list of useful **geom_**.

+ `geom_vline()`: draw a vertical line
+ `geom_hline()`: draw a horizontal line
+ `geom_abline()`: draw a line with the specified intercept and slope
+ `geom_smooth()`: draw an OLS-estimated regression line (other regression methods available)
+ `geom_ribbon()`: create a shaded area
+ `geom_text()` and `annotate()`: add texts in the figure

We will use `g_fig_scatter` to illustrate how these functions work.

]

  ]

.panel[.panel-name[vline and hline]

.left-code[
```{r hv-line, eval = F}  
g_fig_scatter +
  geom_vline(
    xintercept = 10, 
    color = "blue"
  ) +
  geom_hline(
    yintercept = 100, 
    color = "red"
  ) 
```
]

.right-plot[
```{r hv-line-fig, ref.label = "hv-line", echo = F, out.width = "90%"}  
```
]

  ]

.panel[.panel-name[abline]

.left-code[


```{r ab-line, eval = F}  
g_fig_scatter +
  geom_abline(
    #--- a ---#
    intercept = 50, 
    #--- b ---#
    slope = 4,
    color = "blue"
  )  
```

$$y = a + b\times x$$

+ `intercept`: $a$
+ `slope`: $b$

]

.right-plot[
```{r ab-line-fig, ref.label = "ab-line", echo = F, out.width = "90%"}  
```
]

  ] <!-- panel ends here -->

.panel[.panel-name[smooth]

.left-code[
```{r smooth, eval = F}  
g_fig_scatter +
  geom_smooth(
    aes(
      y = corn_yield, 
      x = d3_5_9
    )
  )
```

Also try

```{r , eval = F}  
g_fig_scatter +
  geom_smooth(
    aes(
      y = corn_yield, 
      x = d3_5_9
    ),
    method = "lm"
  )
```

]

.right-plot[
```{r smooth-fig, ref.label = "smooth", echo = F, out.width = "90%"}  

```
]


  ] <!-- panel ends here -->

.panel[.panel-name[ribbon]

.left-code[
```{r ribbon, eval = F}  
g_fig_scatter +
  geom_ribbon(
    aes(
      x = d3_5_9, 
      ymin = 100,
      ymax = 200
    ),
    fill = "green",
    alpha = 0.3
  )

```

+ `ymin`: lower bound of the ribbon
+ `ymax`: upper bound of the ribbon

Useful when drawing confidence intervals.

]

.right-plot[
```{r ribbon-fig, ref.label = "ribbon", echo = F, out.width = "90%"}  

```
]

  ] <!-- panel ends here -->

.panel[.panel-name[text]

.left-code[
```{r text-ex, eval = F}  
g_fig_scatter +
  geom_text(
    aes(
      x = d3_5_9,
      y = corn_yield,
      label = state_name,
    )
  )
```
+ `x`, `y`: position of where texts are placed
+ `label`: variable to print

]

.right-plot[
```{r text-ex-fig, ref.label = "text-ex", echo = F, out.width = "90%"}  

```
]

  ] <!-- panel ends here -->


.panel[.panel-name[annotate]

.left-code[
```{r annotate, eval = F}  
g_fig_scatter +
  annotate(
    'text',
    x = 10,
    y = 50,
    label = 'Drought hurts \n a lot!!',
    size = 3,
    color = "red"
  )
```

+ `x`: where on x-axis
+ `y`: where on y-axis
+ `label`: text to print (\n break the line)
+ size: font size

]

.right-plot[
```{r annotate-fig, ref.label = "annotate", echo = F, out.width = "90%"}  

```
]

  ] <!-- panel ends here -->

]

---



# Multiple datasets in one figure

.panelset[ 

.panel[.panel-name[multiple datasets]

.left-full[

<span style="color:red">Important: </span>`data = county_yield` declared inside `ggplot()` applies to ALL the subsequent `geom_*()`s unless overwritten locally inside individual `geom_*()`s.

Try this:

```{r eval = F}
ggplot() + 
  geom_point(data = county_yield, aes(y = corn_yield, x = d3_5_9)) +
  geom_smooth(aes(y = corn_yield, x = d3_5_9))
```

It is easy to use multiple datasets inside a single `ggplot` object (or a figure). 

You just need to specify what dataset to use locally inside individual `geom_*()`s.

Let's see how this works using an example of drawing the  confidence intervals around the regression lie of the following regression:

<br>

$$corn\_yield = \beta_0 + \beta_1 d3\_5\_9 + v$$

]

  ]

.panel[.panel-name[Preparation]

.left-full[

```{r }
#--- regression ---#
reg <- lm(corn_yield ~ d3_5_9, data = county_yield)

#--- find confidence interval ---#
min_d3 <- county_yield$d3_5_9 %>% min(na.rm = TRUE) # minimum d3 observed
max_d3 <- county_yield$d3_5_9 %>% max(na.rm = TRUE) # maximum d3 observed

eval_points <- data.frame(d3_5_9 = seq(min_d3, max_d3, length = 1000)) # evaluation points

ci_bound <- predict(reg, newdata = eval_points, interval = "confidence", level = 0.9) # upper and lower bound

ci_bound_data <- cbind(eval_points, ci_bound) # combine evaluation points and ci
```

```{r }
head(ci_bound_data) 
```
]

  ] <!-- panel ends here -->

]

---


```{r mult-geom, cache = F, include = F}
ggplot() + 
  #--- scatter plot ---#
  geom_point(
    data = county_yield, 
    aes(y = corn_yield, x = d3_5_9)
  ) + #BREAK
  #--- regression line ---#
  geom_line(
    data = ci_bound_data, 
    aes(x = d3_5_9, y = fit),
    color = "blue",
    size = 1.2
  ) + #BREAK
  #--- confidence interval ---#
  geom_ribbon(
    data = ci_bound_data,
    aes(x = d3_5_9, ymin = lwr, ymax = upr),
    fill = "red",
    alpha = 0.4
  )
```

`r chunk_reveal(
  chunk_name = "mult-geom", 
  title = "# Multiple datasets in one figure",
  break_type = "user", 
  display_type = "both", 
  left_assign = FALSE,
  width = c(50, 50)
  )
`

<!-- 
#=========================================
# Make your figures presentable to others    
#=========================================
-->
  
---

class: inverse, center, middle
name: fine-tune

# Make your figures presentable to others

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

---

# Make your figures presentable to others 


.panelset[ 

.panel[.panel-name[Motivation]

.left-full[

+ Figures we have created so far cannot be used for formal presentations or publications. They are simply too crude.

+ We need fine-tune raw figures before they are publishable. 

+ You can control virtually every element of a figure under the `ggplot2` framework.

+ Take a look at [here](https://ggplot2.tidyverse.org/reference/theme.html) for the complete list of options you can use to modify the theme of figures

<span style="color:red"> Key:</span> The most important thing is actually to know which part of a figure a theme option refers to (e.g., `axis.text`)

]

  ]

.panel[.panel-name[two types]

.left-full[

## Two types of operations

Operations to make your figures presentable can be categorized into two types:

+ Content-altering
+ Theme-altering
 
They are two separate things.

## Examples:

For the y-axis title,

+ The axis title text itself (say "Corn Yield (bu/acre)") falls under the **content** category.   

+ The position of or the font size of the axis-title fall under the **theme** category

The content itself does not change when theme is altered.

]
 
  ] <!-- panel ends here -->

.panel[.panel-name[content-altering]

```{r echo = F}
county_yield_s_b2010 <- county_yield %>% 
  filter(year >= 2005, year <= 2010)

g_box <- ggplot(data = county_yield_s_b2010) + 
  geom_boxplot(
    aes(
      x = factor(year), 
      y = corn_yield, 
      fill = state_name
    )
  )
```

.left5[

Original

```{r original-f, echo = F}
g_box 
```
]

.right5[

Altered

```{r altered-f, echo = F}
g_box +
  xlab("Year") +
  ylab("Corn Yield (bu/acre)") +
  scale_fill_viridis_d()
```
]


  ] <!-- panel ends here -->

.panel[.panel-name[theme-altering]

.left5[

Original

```{r t-original-f, echo = F}
g_box 
```
]

.right5[

Altered

```{r t-altered-f, echo = F}
g_box +
  theme(
    axis.title = element_text(size = 20)
  )
```
]

  ] <!-- panel ends here -->

.panel[.panel-name[Note]

.left-full[

<br>

+ Distinctions between the two types of actions are not always clear

+ But, typically, you use 
  * `scale_*()` function series to alter contents
  * `theme()` function to alter the theme

+ Note that there are shorthand convenience functions to alter figure contents for commonly altered parts of figures
]

  ] <!-- panel ends here -->

]

---

# Axes content  

.panelset[ 

.panel[.panel-name[Preparation]

We are going to build on this figure in this section:

.left-code[

```{r g-box, results = "hide"}  
county_yield_s_b2010 <- county_yield %>% 
  filter(year >= 2005, year <= 2010)

g_box <- ggplot(data = county_yield_s_b2010) + 
  geom_boxplot(
    aes(
      x = factor(year), 
      y = corn_yield, 
      fill = state_name
    )
  )
```
]

.right-plot[
```{r g-box-f-a, echo = F, out.width = "90%"}  
g_box
```
]

  ] <!-- panel ends here -->

.panel[.panel-name[How]

.left-full[
We can use 

+ `scale_x_discrete()`/`scale_x_continuous()` for x-axis
+ `scale_y_discrete()`/`scale_y_continuous()` for y-axis

to control the following elements of axes:

+ `name`: an axis title 
+ `limit`: the range of an axis
+ `breaks`: axis ticks positions
+ `label`: axis texts at ticks

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

We use `scale_x_discrete()` if `x` is a discrete variable (not numeric) and `scale_x_continuous()` if `x` is a continuous variable (numeric).

The same applies for `y`.

]


  ] <!-- panel ends here -->

.panel[.panel-name[axis title]

.left5[


```{r axis-title, eval = F}  
g_box + 
  scale_x_discrete(name = "Year") + #<<
  scale_y_continuous(name = "Corn Yield (bu/acre)") #<< 
```

Or just this,

```{r eval = F}
g_box + 
  xlabel("Year") + #<<
  ylabel("Corn Yield (bu/acre)") #<< 
```

]

.right5[
```{r ref.label = "axis-title", echo = F, out.width = "90%"}  

```
]

  ]

.panel[.panel-name[limit]

.left5[
```{r axis-limit, eval = F}  
g_box + 
  scale_x_discrete(
    name = "Year" 
  ) +
  scale_y_continuous(
    name = "Corn Yield (bu/acre)",
    #--- first min, second max ---#
    limit = c(100, 200) #<<
  ) 
```

Or just,

```{r eval = F}
g_box + 
  xlabel("Year") +
  ylabel("Corn Yield (bu/acre)") +
  ylim(100, 200) #<<
     
```

Or,

You can filter the data first and then use the filtered data.

]

.right5[
```{r ref.label = "axis-limit", echo = F, out.width = "90%"}  

```
]


  ]

.panel[.panel-name[breaks and labels]

.left5[
```{r axis-breaks, eval = F}  
g_box + 
  scale_x_discrete(
    name = "Year",
    label = gsub("20", "", as.character(2000:2018)) #<<
  ) +
  scale_y_continuous(
    name = "Corn Yield (bu/acre)",
    limit = c(100, 200),
    breaks = seq(100, 200, by = 10) #<<
  ) 
```

<br>

+ `breaks`: determines where the ticks are located
+ `labels`: defines the texts at the ticks

]

.right5[
```{r ref.label = "axis-breaks", echo = F, out.width = "90%"}  

```
]


  ] <!-- panel ends here -->

]

---

# Legends content

.panelset[ 

.panel[.panel-name[Preparation]

We are going to build on this figure in this section:

.left-code[
```{r start-leg-content, results = "hide"}  
g_axis <- g_box + 
  scale_x_discrete(
    name = "Year",
    label = gsub("20", "", as.character(2000:2018))
  ) +
  scale_y_continuous(
    name = "Corn Yield (bu/acre)",
    limit = c(100, 200),
    breaks = seq(100, 200, by = 10)
  )
```
]

.right-plot[
```{r start-leg-content-f, echo = F, out.width = "90%"}  
g_axis
```
]

 

  ] <!-- panel ends here -->

.panel[.panel-name[title]


.left-code[

```{r l-title, eval = F}  
g_axis + 
  scale_fill_brewer(name = "State") #<<
```

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=350px></html>

Or,

```{r , eval = F}  
g_axis + 
  labs(fill = "State") #<<
```

]

.right-plot[
```{r ref.label = "l-title", echo = F, dependson = "starting-point", out.width = "90%"}  

```
]
  ] <!-- panel ends here -->

.panel[.panel-name[fill color]

.left-code[

```{r l-fill, eval = F}  
g_axis + 
  scale_fill_brewer(
    name = "state", 
    palette = "Set1" #<<
    )
```

<span style="color:red"> We are going to spend lots of time on color scheme later. </span>

]

.right-plot[
```{r ref.label = "l-fill", echo = F, dependson = "starting-point", out.width = "90%"}  

```
]

  ] <!-- panel ends here -->

.panel[.panel-name[title position]

.left-code[

```{r l-title-pos, eval = F}  
g_axis + 
  scale_fill_brewer(
    name = "state", 
    palette = "Set1",
    guide = guide_legend( #<<
      title.position = "left" #<<
    ) #<<
  )
```

]

.right-plot[
```{r ref.label = "l-title-pos", echo = F, dependson = "starting-point", out.width = "90%"}  

```
]

  ] <!-- panel ends here -->

.panel[.panel-name[wrapping]

.left-code[

```{r l-wrapping, eval = F}  
g_axis + 
  scale_fill_brewer(
    name = "state", 
    palette = "Set1",
    guide = guide_legend(
      title.position = "left",
      nrow = 2 #<<
    )
  )
```

]

.right-plot[
```{r ref.label = "l-wrapping", echo = F, dependson = "starting-point", out.width = "90%"}  

```
]

  ] <!-- panel ends here --> 
]

---


# Theme

.panelset[ 

.panel[.panel-name[Naming rules]

.left-full[

When specifying the theme of figure elements, it is good to know the naming convention of figure elements:

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

For example:

+ `axis.title` 

This refers to the title of both x- and y-axis. Any aesthetic theme you apply to this element will be reflected on the title of both x- and y-axis.  

+ `axis.title.x` 

This refers to the title of only x-axis. Any aesthetic theme you apply to this element will be reflected on the title of only x-axis.

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

So, basically appending `.name` narrows down the scope of the figure elements the element name refers to.

]

  ] <!-- panel ends here -->

.panel[.panel-name[Common functions]

.left-full[

There are common functions we use to specify the aesthetic nature of figure elements based on the type of the elements:

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

<br>

+ `element_text()`: for text elements like `axis.text`, `axis.title`, `legend.text`

Inside the function, you specify things like font size, font family, angle, etc.

+ `element_rect()`: for box-like elements like `legend.background`, `plot.background`, `strip.background`

Inside the function, you specify things like font background color, border line color, etc.

+ `element_line()`: for line elements like `panel.grid.major`, `axis.line.x`

Inside the function, you specify things like line thickness, line color, etc.

+ `element_blank()`: any components

It makes the specified component disappear.

+ `unit()`: for attributes of figure elements like `legend.key.width`, `legend.box.spacing` 
]

  ]

]


---

# Axis theme 

.panelset[ 

.panel[.panel-name[Preparation]

We are going to build on this figure in this section:

.left-code[
```{r start-axis-theme, eval = F}  
g_axis
```
]

.right-plot[
```{r start-axis-theme-f, echo = F, out.width = "90%"}  
g_axis
```
]

  ]

.panel[.panel-name[title and text]

.left5[
```{r at-font, eval = F}
g_axis +
  theme(
    axis.title.x = element_text(size = 8, color = "red"), #<<
    axis.text = element_text(size = 14, family = "Times") #<<
  )
```
]

.right5[
```{r at-font-f, ref.label = "at-font", echo = F, out.width = "90%"}

```
]

  ]

.panel[.panel-name[line]

.left5[
```{r at-line, eval = F}
g_axis +
  theme(
    axis.title.x = element_text(size = 8, color = "red"),
    axis.text = element_text(size = 14, family = "Times"),
    axis.line.y = element_line(size = 2, color = "blue") #<<
  )
```

]

.right5[
```{r at-line-f, ref.label = "at-line", echo = F, out.width = "90%"}

```
]

  ]

.panel[.panel-name[ticks]

.left5[
```{r at-ticks, eval = F}
g_axis +
  theme(
    axis.title.x = element_text(size = 8, color = "red"),
    axis.text = element_text(size = 14, family = "Times"),
    axis.line.y = element_line(size = 2, color = "blue"),
    axis.ticks.length.x = unit(2, "cm") #<<
  )
```
]

.right5[
```{r at-ticks-f, ref.label = "at-ticks", echo = F, out.width = "90%"}

```
]


  ]

  <!-- panel ends here -->

]


---


# Legends theme

.panelset[ 


.panel[.panel-name[How]

<br>

We can use `them()` to change the aesthetics of legends. Some of the elements include

+ title
+ position
+ key 
+ text
+ direction
+ background

See [here](https://ggplot2.tidyverse.org/reference/theme.html) for the full list of options related to legends.

We will discuss how to change the color scheme of legends later in much detail.

  ]

.panel[.panel-name[Preparation]

This is what we will build on: 

.left-code[
```{r starting-point}  
g_l <- g_axis +
  scale_fill_brewer(
    palette = "Paired",
    guide = guide_legend(
      title.position = "left",
      nrow = 2
    )
  )
```
]

.right-plot[
```{r starting-point-f, echo = F, dependson = "starting-point", out.width = "90%"}  
g_l
```
]

  ] 

.panel[.panel-name[position]

.left-code[
```{r l-position, eval = F}  
g_l +
  labs(fill = "State") +
  theme( #<<
    legend.position = "bottom" #<<
  ) #<<
```

]

.right-plot[
```{r l-position-f, ref.label = "l-position", echo = F, dependson = "starting-point", out.width = "90%"}  

```
]

  ]

.panel[.panel-name[key]

.left-code[
```{r key, eval = F}  
g_l +
  labs(fill = "State") +
  theme(
    legend.position = "bottom",
    legend.key.height = unit(0.5, "cm"), #<<
    legend.key.width = unit(2, "cm") #<<
  ) 
```

]

.right-plot[
```{r l-key-f, ref.label = "key", echo = F, out.width = "90%"}  

```
]

  ]

.panel[.panel-name[font]

.left-code[
```{r font, eval = F}  
g_l +
  labs(fill = "State") +
  theme(
    legend.position = "bottom",
    legend.key.height = unit(0.5, "cm"),
    legend.key.width = unit(2, "cm"),
    legend.text = element_text( #<<
      size = 16, #<<
      family = "Times" #<<
    ), 
    legend.title = element_text( #<<
      size = 6, #<<
      family = "Courier", #<<
      color = "red" #<<
    ), 
  ) 
```
]

.right-plot[
```{r l-font-f, ref.label = "font", echo = F, out.width = "90%"}  

```
]

  ]

.panel[.panel-name[background]

.left-code[
```{r background, eval = F}  
g_l +
  labs(fill = "State") +
  theme(
    legend.position = "bottom",
    legend.key.height = unit(0.5, "cm"),
    legend.key.width = unit(2, "cm"),
    legend.text = element_text(
      size = 16, 
      family = "Times"
    ),
    legend.title = element_text(
      size = 6, 
      family = "Courier",
      color = "red"
    ),
    legend.background = element_rect( #<<
      fill = "lightblue", #<<
      linetype = "solid" #<<
    ) #<<
  ) 
```
]

.right-plot[
```{r l-bcg-f, ref.label = "background", echo = F, out.width = "90%"}  

```
]
  ] <!-- panel ends here -->

]

---


# Pre-made themes

.panelset[ 

.panel[.panel-name[Instruction]

.left-full[
There are a bunch of pre-made themes from the `ggplot2` and `ggthemes` packages that can quickly change how figures look.

Install and library  `ggthemes` package first:

```{r eval = F}
#--- install ---#
install.packages("ggthemes") 

#--- library ---#
library("ggthemes") 
```

See the full list of pre-made themes [here](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/).

]

  ]

.panel[.panel-name[bw]

.left-code[
```{r bw, eval = F}  
g_axis +
  theme_bw()
```
]

.right-plot[
```{r bw-f, ref.label = "bw", echo = F, out.width = "90%"}  

```
]

  ]

.panel[.panel-name[void]

.left-code[
```{r void, eval = F}  
g_axis +
  theme_void()
```
]

.right-plot[
```{r void-f, ref.label = "void", echo = F, out.width = "90%"}  

```
]

  ]

.panel[.panel-name[stata]

.left-code[
```{r stata, eval = F}  
g_axis +
  theme_stata()
```
]

.right-plot[
```{r stata-f, ref.label = "stata", echo = F, out.width = "90%"}  

```
]

  ]

.panel[.panel-name[gdocs]

.left-code[
```{r gdocs, eval = F}  
g_axis +
  theme_gdocs()
```
]

.right-plot[
```{r gdocs-f, ref.label = "gdocs", echo = F, out.width = "90%"}  

```
]

  ]

.panel[.panel-name[economist]

.left-code[
```{r economist, eval = F}  
g_axis +
  theme_economist()
```
]

.right-plot[
```{r economist-f, ref.label = "economist", echo = F, out.width = "90%"}  

```
]

  ]

.panel[.panel-name[excel]

.left-code[
```{r excel, eval = F}  
g_axis +
  theme_excel()
```
]

.right-plot[
```{r excel-f, ref.label = "excel", echo = F, out.width = "90%"}  

```
]

  ]

]

---

# Panel (build on a pre-made theme)

.panelset[ 

.panel[.panel-name[How]

.left-full[

## Build on pre-made theme
You can simply override parts of the pre-made theme by adding theme options like this (see more on this [here](#custom-theme)):

```{r eval = F}
g_axis +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank()
  )
```

So, you can pick the pre-made theme that looks the closest to what you would like, and then add on theme elements to the part you do not like.

]

  ]

.panel[.panel-name[Preparation]

.left-code[
This is what we will be building on:

```{r start-panel, eval = F}  
g_axis +
  theme_bw()  
```
]

.right-plot[
```{r start-panel-f, ref.label = "start-panel", echo = F, out.width = "90%"}  

```
]


  ] <!-- panel ends here -->

.panel[.panel-name[minor grid]


.left-code[
```{r panel-grid-minor, eval = F}
g_axis +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank() #<<
  )
```

]

.right-plot[
```{r panel-grid-minor-f, ref.label = "panel-grid-minor", echo = F, out.width = "90%"}

```
]

  ]

.panel[.panel-name[major grid]

.left-code[
```{r panel-grid-major, eval = F}
g_axis +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(), #<<
    panel.grid.major.y = element_line( #<<
      size = 1,  #<<
      color = "blue", #<< 
      linetype = "dotted" #<<
    ) #<<
  )
```

See [here](http://sape.inf.usi.ch/quick-reference/ggplot2/linetype) for the line types available.

]

.right-plot[
```{r panel-grid-major-f, ref.label = "panel-grid-major", echo = F, out.width = "90%"}

```
]

  ]

]

---


# Faceted figure theme 

.panelset[ 

.panel[.panel-name[Instruction]

<br>

Faceted figures have `strip` elements that do no exist for non-faceted figures like

+ `strip.background`
+ `strip.placement`
+ `strip.text`
+ `panel.spacing`

We learn how to modify these elements. 

  ]

.panel[.panel-name[Preparation]

.left-code[

Create a dataset for this section:

```{r }
county_yield_f <- county_yield %>% 
  filter(state_name %in% c("Nebraska", "Colorado", "Kansas")) %>% 
  filter(year %in% c(2005, 2006))  
```

Create a faceted figure we will build on:

```{r f-prep, echo = F}
g_f <- ggplot(data = county_yield_f) + 
  geom_histogram(aes(x = corn_yield)) +
  facet_grid(
    state_name ~ year, 
    scales = "free_x"
  )
```

]

.right-plot[
```{r f-prep-f, echo = F, out.width = "90%"}  
g_f
```
]

  ] 

.panel[.panel-name[text]

.left-code[
```{r st-text, eval = F}
g_f +
  theme(
    strip.text.x = element_text( #<<
      size = 12, #<<
      family = "Times", #<< 
      color = "red" #<<
    ), #<<
    strip.text.y = element_text( #<<
      angle = 0, #<< 
      size = 6, #<<
      color = "blue" #<<
    ) #<<
  )
```

]

.right-plot[
```{r st-text-f, ref.label = "st-text", echo = F, out.width = "90%"}

```
]

  ]

.panel[.panel-name[background]

.left-code[
```{r st-background, eval = F}
g_f +
  theme(
    strip.text.x = element_text(
      size = 12, 
      family = "Times", 
      color = "red"
    ),
    strip.text.y = element_text(
      angle = 0, 
      size = 6, 
      color = "blue"
    ),
    strip.background.x = element_rect( #<<
      color = "blue" #<<
    ), #<<
    strip.background.y = element_blank() #<<
  ) 
```

]

.right-plot[
```{r st-background-f, ref.label = "st-background", echo = F, out.width = "90%"}

```
]

  ] 

.panel[.panel-name[panel spacing]


.left-code[
```{r p-spacing, eval = F}
g_f +
  theme(
    strip.text.x = element_text(
      size = 12, 
      family = "Times", 
      color = "red"
    ),
    strip.text.y = element_text(
      angle = 0, 
      size = 6, 
      color = "blue"
    ),
    strip.background.x = element_rect(
      color = "blue"
    ),
    strip.background.y = element_blank(),
    panel.spacing.x = unit(2, "cm"), #<<
    panel.spacing.y = unit(0.01, "cm") #<<
  ) 
```

]

.right-plot[
```{r p-spacing-f, ref.label = "p-spacing", echo = F, out.width = "90%"}

```
]

  ] <!-- panel ends here -->

]

---


# More font families 

.panelset[ 

.panel[.panel-name[Instruction]

.left-full[

You can use more font families by taking advantage of the `extrafont` package. 

Install and library the package, and then run `font_import()` to import font families (you need to do this only once). 

```{r eval = F}
install.packages(extrafont)

library(extrafont)

font_import()
```

Next time around, just run `library(extrafont)`.

]

  ]

.panel[.panel-name[font families]

```{r echo = F}
fonttable() %>% 
  select(FontName, Bold,  Italic) %>% 
  datatable()
```

  ]

.panel[.panel-name[try a family]


.left-code[
```{r f-fam, eval = F}
g_f +
  theme(
    strip.text.y = element_text(
      family = "Georgia", #<<
      color = "red"
    )
  )
```
]

.right-plot[
```{r f-fam-f, ref.label = "f-fam", echo = F, out.width = "90%"}

```
]

  ] <!-- panel ends here -->

]

---

# More flexible color options with HEX 

.panelset[ 

.panel[.panel-name[Instruction]

.left-full[
Instead of naming the color you want to use, you can use **HEX color codes** instead.

<span style="color:red"> Direction: </span>

+ Visit [here](https://www.color-hex.com/)
+ Click on any color you like
+ Then you will see two sets of color gradients (thicker and lighter from the color you picked)
+ Pick the color you like from the color bar and copy the HEX color code beneath the color you picked

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

You could alternatively use the RGB codes, but I do not see any reasons to do so because the use of HEX codes is sufficient. 

]

  ]

.panel[.panel-name[Example]


.left-code[
```{r color-hex, eval = F}
ggplot(data = county_yield) +
  geom_point(
    aes(y = corn_yield, x = d3_5_9), 
    color = "#824283" #<<
  ) 
```

You can use HEX color codes for any color-related elements in a figure.

]

.right-plot[
```{r color-hex-f, ref.label = "color-hex", echo = F, out.width = "90%"}

```
]

  ]

]


---

# Color scale

.panelset[ 
 
.panel[.panel-name[Intro 1]

.left-full[

The choice of color schemes for your figures are very important (not so much for academic journals ...)

We use `scale_A_B()` functions to for color specification:

+ **A** is the name of aesthetic (`color` or `fill`)
+ **B** is the type of color specification method

]

   ]
 
.panel[.panel-name[Intro 2]

.left-full[

For example, consider the following code:

```{r color-aes, eval = F, echo = F}
ggplot(data = county_yield) +
  geom_point(aes(y = corn_yield, x = d3_5_9, color = corn_yield))
```

```{r echo = F}
decorate("color-aes", eval = F) %>% 
  flair("color = corn_yield") %>%
  knit_print.with_flair()
```

Since it is the `color` aesthetic that we want to work on, **A** = `color`.

There are many options for **B**. Indeed, there are so many that, it gets confusing!

+ `scale_color_brewer()` (discrete)
+ `scale_color_distiller()` (continuous)
+ `scale_color_viridis_d()` (discrete)
+ `scale_color_viridis_c()` (continuous)
+ `scale_color_continuous()` (continuous)
+ `scale_color_discrete()` (discrete)
+ `scale_color_hue()` (discrete)

One thing to remember is that you need to be aware of whether the aesthetic variable (here, `corn_yield`) is numeric or not as that determines acceptable type of **B**.

]


   ]
 
 ]
  
---

# Viridis 

.panelset[ 

.panel[.panel-name[Instruction]

<br>

We have four `scale` functions for Viridis color map:

+ `scale_color_viridis_c()`: for `color` aesthetic with a continuous variable
+ `scale_color_viridis_d()`: for `color` aesthetic with a discrete variable
+ `scale_fill_viridis_c()`: for `fill` aesthetic with a continuous variable
+ `scale_fill_viridis_d()`: for `fill` aesthetic with a discrete variable

There are five color scheme types under the Viridis color map:

+ magma
+ inferno
+ plasma
+ viridis
+ civiris

You can use `option` to specify which one of them you want to use inside the `scale` functions.

These color schemes are color-blind sage.

  ] <!-- panel ends here -->

.panel[.panel-name[types]

```{r viridis-ex, echo = F, fig.dim = c(6, 3.5), out.width = "80%"}
data("geyser", package = "MASS")

ggplot(geyser, aes(x = duration, y = waiting)) +
  xlim(0.5, 6) + ylim(40, 110) +
  stat_density2d(aes(fill = ..level..), geom="polygon") +
  theme_bw() +
  theme(panel.grid=element_blank()) -> gg

((gg + scale_fill_viridis_c(option="A") + labs(x="magma (option A)", y=NULL)) |
(gg + scale_fill_viridis_c(option="B") + labs(x="inferno (option B)", y=NULL)))/ 
((gg + scale_fill_viridis_c(option="C") + labs(x="plasma (option C)", y=NULL)) |
(gg + scale_fill_viridis_c(option="D") + labs(x="viridis (default)", y=NULL)))

```

  ]

.panel[.panel-name[Example 1]


.left-code[
```{r viridis-ex-1, eval = F}
ggplot(data = filter(county_yield, corn_yield > 50)) +
  geom_point(
    aes(
      y = corn_yield, 
      x = d3_5_9, 
      color = corn_yield
    )
  ) +
  scale_color_viridis_c() #<<
```

]

.right-plot[
```{r viridis-ex-1-f, ref.label = "viridis-ex-1", echo = F, out.width = "90%"}

```
]

  ]

.panel[.panel-name[Example 2]

.left-code[
```{r viridis-ex-2, eval = F}
ggplot(data = filter(county_yield, corn_yield > 50)) +
  geom_point(
    aes(
      y = corn_yield, 
      x = d3_5_9, 
      color = corn_yield
    )
  ) +
  scale_color_viridis_c(option = 2) #<<
```

]

.right-plot[
```{r viridis-ex-2-f, ref.label = "viridis-ex-2", echo = F, out.width = "90%"}

```
]

  ] <!-- panel ends here -->

.panel[.panel-name[reverse]

.left-code[
```{r viridis-reverse, eval = F}
ggplot(data = filter(county_yield, corn_yield > 50)) +
  geom_point(
    aes(
      y = corn_yield, 
      x = d3_5_9, 
      color = corn_yield
    )
  ) +
  scale_color_viridis_c(
    option = 2, 
    direction = - 1 #<<
  )
```

]

.right-plot[
```{r viridis-reverse-f, ref.label = "viridis-reverse", echo = F, out.width = "90%"}

```
]

  ] <!-- panel ends here -->

]

---

# RColorBrewer 

.panelset[ 

.panel[.panel-name[Instruction]

.left-full[
`RColorBrewer` package provides a number of color palettes of three types:

+ sequential: suitable for a variable that has ordinal meaning (e.g., temperature, precipitation)
+ diverging: suitable for variables that take both negative and positive values (e.g., changes in groundwater level)
+ qualitative: suitable for qualitative or categorical variable

These palettes are particularly suitable for maps.

<span style="color:red"> Direction: </span>visit [here](https://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3). 

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

We use two types of scale functions for the palettes:

+ `scale_A_brewer()`: for discrete aesthetic variable
+ `scale_A_distiller()`: for continuous aesthetic variable 
 
]

  ]

.panel[.panel-name[sequential]

```{r seq-s, out.width = "60%", fig.dim = c(6, 5)}
display.brewer.all(type = "seq")  
```

  ]

.panel[.panel-name[diverging]

```{r div-s, out.width = "60%", fig.dim = c(6, 4)}
display.brewer.all(type = "div")  
```

  ]

.panel[.panel-name[qualitative]

```{r qua-s, out.width = "60%", fig.dim = c(6, 4)}
display.brewer.all(type = "qual")  
```

  ]

.panel[.panel-name[Example 1]


.left-code[

Generate a dataset for visualization:

```{r , results = "hide"}  
county_yield_s_b2010 <- county_yield %>% 
  filter(year >= 2005, year <= 2010)
```

Create a figure:

```{r rb-ex-1, eval = F}
ggplot(data = county_yield_s_b2010) + 
  geom_boxplot(
    aes(
      x = factor(year), 
      y = corn_yield, 
      fill = state_name
    )
  ) +
  scale_fill_brewer(palette = "Set2") #<<
```

]

.right-plot[
```{r rb-ex-1-f, ref.label = "rb-ex-1", echo = F, out.width = "90%"}

```
]

  ] <!-- panel ends here -->

.panel[.panel-name[Example 2]

.left-code[

```{r rb-ex-2, eval = F}
ggplot(data = filter(county_yield, corn_yield > 50)) +
  geom_point(
    aes(
      y = corn_yield, 
      x = d3_5_9, 
      color = corn_yield
    )
  ) +
  scale_color_distiller(palette = "RdYlGn") #<<
```

]

.right-plot[
```{r rb-ex-2-f, ref.label = "rb-ex-2", echo = F, out.width = "90%"}

```
]

  ] <!-- panel ends here -->

]

---


# Set discrete color scale manually 

.panelset[ 

.panel[.panel-name[Instruction]

.left-full[

Sometimes, you just want to pick colors yourself. In that case, you can use 

+ `scale_color_manual()`
+ `scale_fill_manual()`

You specify a sequence of colors and supply it to the `scale` function via the `values` option.

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

It is recommended that you use a named vector like.

For example, consider the box plot of corn yield for four states: Colorado, Kansas, Nebraska, and South Dakota. Then, a sample named vector looks like this:

```{r eval = F}
cols <- c("Colorado" = "red", "Nebraska" = "blue", "Kansas" = "orange", "South Dakota" = "#ff0080")
```

Now that a named vector is created, you can do the following to impose the color scheme you just defined.

```{r eval = F}
scale_fill_manual(values = cols)
```

]
 

  ]

.panel[.panel-name[Example]

.left-code[

Define a named color vector:

```{r }
cols <- c("Colorado" = "red", "Nebraska" = "blue", "Kansas" = "orange", "South Dakota" = "#ff0080")
```

Create a figure:

```{r manu-ex, eval = F}
ggplot(data = county_yield_s_b2010) + 
  geom_boxplot(
    aes(
      x = factor(year), 
      y = corn_yield, 
      fill = state_name
    )
  ) +
  scale_fill_manual(values = cols) #<<
```

]

.right-plot[
```{r manu-ex-f, ref.label = "manu-ex", echo = F, out.width = "90%"}

```
]


  ]

]

 

---


# Patchwork (grid of figures)

.panelset[ 

.panel[.panel-name[Instruction]

.left-code[

The `patchwork` package allows you to combine and arrange multiple figures (and even tables and texts) like the figure to the right:

```{r eval = F}
install.packages("patchwork")  

library("patchwork")
```

See the [`patchwork` package website](https://patchwork.data-imaginist.com/index.html) for a fuller treatment of this package.

]

.right-plot[
```{r patch-1, echo = F, out.width = "90%"}  
g_fig_scatter | g_box
```
]

  ]

.panel[.panel-name[Preparation]

.left-full[

We are going to use following figures:

```{r gen-ggs}
g_1 <- ggplot(data = county_yield) + 
  geom_histogram(aes(x = corn_yield)) +
  ggtitle("g_1") 

g_2 <- ggplot(data = county_yield) + 
  geom_boxplot(aes(x = factor(year), y = corn_yield)) +
  ggtitle("g_2")

g_3 <- ggplot(data = county_yield) + 
  geom_density(aes(x = corn_yield)) +
  ggtitle("g_3")

g_4 <- ggplot(data = mean_yield) + 
  geom_line(aes(x = year, y = corn_yield)) +
  ggtitle("g_4")
```

Note: `mean_yield` for `g_4` is created in the **Line Plot** tab in slide 9.

]

  ] 

.panel[.panel-name[+]

```{r plus-1, out.width = "75%", fig.dim = c(8, 5), dependson = "gen-ggs"}
g_1 + g_2 
```

  ] <!-- panel ends here -->

.panel[.panel-name[++]

```{r pp-1, out.width = "75%", fig.dim = c(8, 5), dependson = "gen-ggs"}
g_1 + g_2 + g_3 + g_4 
```

  ] <!-- panel ends here -->

.panel[.panel-name[/]

```{r v-1, out.width = "75%", fig.dim = c(8, 5), dependson = "gen-ggs"}
g_1 / g_2
```

  ] <!-- panel ends here -->

.panel[.panel-name[|]

```{r h-1, out.width = "75%", fig.dim = c(8, 5), dependson = "gen-ggs"}
g_1 | g_2
```

  ] <!-- panel ends here -->

.panel[.panel-name[||]

```{r hh-1, out.width = "75%", fig.dim = c(8, 5), dependson = "gen-ggs"}
g_1 | g_2 | g_3 | g_4
```

  ] <!-- panel ends here -->

.panel[.panel-name[()]

```{r group-p-1, out.width = "75%", fig.dim = c(8, 5), dependson = "gen-ggs"}
g_1 | (g_3 / g_4)
```

  ] <!-- panel ends here -->

.panel[.panel-name[faceting?]

.left-full[

The difference between faceted figures and panel of independent figures

+ `facet_*()`: faceted figures share the same legend   
+ `patchwork`: panel of independent figures can clearly have independent legends    

Faceting is not suitable for presenting multiple distinct variables as they need to share the same legend.

Imagine plotting temperature (in Celsius) and precipitation (in mm) as faceted figures.

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

<span style="color:red"> Note: </span>

+ Before you use the `patchwork` packages to do figure arrangement, think a bit if you really need to do it. Can't you just arrange figures in WORD or latex? 
+ It is very useful if the destination of the panels of figures is html (e.g., shiny, flex dashboard) because it is not trivial to arrange figures the way the `patchwork` allows you to (you certainly need to know how **css** works).

]

  ] 

]

<!-- 
#=========================================
# Random tips    
#=========================================
-->
  
---

class: inverse, center, middle
name: tips

# Tips 

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

---

# Change the order 

.panelset[ 

.panel[.panel-name[Preparation]

.left-full[

Create a dataset for this section:

```{r }
county_yield_f <- county_yield %>% 
  filter(state_name %in% c("Nebraska", "Colorado", "Kansas")) %>% 
  filter(year %in% c(2005, 2006))  
```

Create a faceted figure we will work on:

```{r eval = F}
g_f <- ggplot(data = county_yield_f) + 
  geom_histogram(aes(x = corn_yield)) +
  facet_grid(
    state_name ~ year, 
    scales = "free_x"
  )
```

]

  ]

.panel[.panel-name[Problem]

<br>

+ You want the panels of figures to appear in the order of Nebraska, Kansas, and Colorado.

+ But, by default, `ggplot2` orders panels in alphabetical order

```{r problem-f, echo = F, fig.dim = c(8, 4), out.width = "70%"}
g_f 
```

  ]

.panel[.panel-name[Solution]

.left-full[

You can turn `state_name` into a factor with the preferred order of state names.

```{r }
county_yield_f <- county_yield_f %>% 
  mutate(
    state_name_f = factor(state_name, level = c("Nebraska", "Kansas", "Colorado")
  )
) 
```  

```{r }
county_yield_f$state_name_f %>% head()
```

]

  ] <!-- panel ends here -->

.panel[.panel-name[Problem solved]

.left-code[
```{r factor, echo = F, eval = F}
ggplot(data = county_yield_f) + 
  geom_histogram(aes(x = corn_yield)) +
  facet_grid(
    state_name_f ~ year, 
    scales = "free_x"
  )
```

```{r echo = F}
decorate("factor", eval = F) %>% 
  flair("state_name_f") %>%
  knit_print.with_flair()
```
]

.right-plot[
```{r problem-f-s, ref.label = "factor", echo = F, out.width = "90%"}

```
]

  ] <!-- panel ends here -->


]

---

# Box plot 

.panelset[ 

.panel[.panel-name[this works]

.left-code[

```{r}
county_yield_s_b2010 <- county_yield %>% 
  filter(year >= 2005, year <= 2010)
```

```{r box-disc, eval = F, echo = F}
ggplot(data = county_yield_s_b2010) + 
  geom_boxplot(
    aes(
      x = factor(year), 
      y = corn_yield, 
      fill = state_name
    )
  ) 
```

```{r echo = F}
decorate("box-disc", eval = F) %>% 
  flair('factor(year)') %>%
  knit_print.with_flair()
```

+ The `x` variable has to be discrete (character or factor). 
+ `factor(year)` converts `year` into a `factor`

]

.right-plot[
```{r box-disc-f, ref.label = "box-disc", echo = F, out.width = "90%"}
```
]

  ] <!-- panel ends here -->

.panel[.panel-name[not this]

.left-code[

```{r}
county_yield_s_b2010 <- county_yield %>% 
  filter(year >= 2005, year <= 2010)
```

```{r box-num, eval = F, echo = F}
ggplot(data = county_yield_s_b2010) + 
  geom_boxplot(
    aes(
      x = year, 
      y = corn_yield, 
      fill = state_name
    )
  ) 
```

```{r echo = F}
decorate("box-num", eval = F) %>% 
  flair('x = year') %>%
  knit_print.with_flair()
```

+ The `x` variable has to be discrete (character or factor). 
+ `year` is numeric

]

.right-plot[
```{r box-num-f, ref.label = "box-num", echo = F, out.width = "90%"}
```
]

  ] <!-- panel ends here -->

.panel[.panel-name[this works]

.left-code[

```{r}
county_yield_s_b2010 <- county_yield %>% 
  filter(year >= 2005, year <= 2010)
```

```{r box-char, eval = F, echo = F}
ggplot(data = county_yield_s_b2010) + 
  geom_boxplot(
    aes(
      x = as.character(year), 
      y = corn_yield, 
      fill = state_name
    )
  ) 
```

```{r echo = F}
decorate("box-char", eval = F) %>% 
  flair('x = as.character(year)') %>%
  knit_print.with_flair()
```

+ The `x` variable has to be discrete (character or factor). 
+ `as.character(year)` converts `year` into a `character` variable

]

.right-plot[
```{r box-char-f, ref.label = "box-char", echo = F, out.width = "90%"}
```
]


  ] <!-- panel ends here -->


.panel[.panel-name[this?]



.left-code[

Load the `nycflights13` package to get access to the `weather` dataset.

```{r }
library(nycflights13)  
```

Create a box plot of temperature by month:
```{r weather, eval = F, echo = F}
ggplot(data = weather) +
  geom_boxplot(
    aes(
      y = temp, 
      x = as.character(month)
    )
  )
```

```{r echo = F}
decorate("weather", eval = F) %>% 
  flair('x = as.character(month)') %>%
  knit_print.with_flair()
```

+ Remember using `month` (which is a numeric variable) would not have worked
+ Notice that 10 and 11 come after 1 

]

.right-plot[
```{r weather-f, ref.label = "weather", echo = F, out.width = "90%"}  

```
]


  ] <!-- panel ends here -->

.panel[.panel-name[this works]

.left-code[

Load the `nycflights13` package to get access to the `weather` dataset.

```{r }
library(nycflights13)  
```

Create a box plot of temperature by month:
```{r weather-2, eval = F, echo = F}
ggplot(data = weather) +
  geom_boxplot(
    aes(
      y = temp, 
      x = factor(month)
    )
  )
```

```{r echo = F}
decorate("weather-2", eval = F) %>% 
  flair('x = factor(month)') %>%
  knit_print.with_flair()
```

]

.right-plot[
```{r weather-2-f, ref.label = "weather-2", echo = F, out.width = "90%"}  

```
]

  ] <!-- panel ends here -->
]

---
name: custom-theme

# Custom theme 

.panelset[ 

.panel[.panel-name[Custom theme]

.left-full[

You can create your own theme, save it, and then use it later. 

Here, I am creating my own theme off of `theme_economist()`, where axis titles and major panel grids are absent.

```{r }
my_theme <- theme_economist() +
  theme(
    axis.title = element_blank(),
    panel.grid.major = element_blank()  
  )
```

You can add `my_theme` like below just like a regular pre-made theme:

```{r eval = F}
ggplot(data = weather) +
  geom_boxplot(
    aes(y = temp, x = factor(month))
  ) +
  my_theme
```

]

  ]

.panel[.panel-name[Compare]

.left5[

```{r t-econ-f, out.width = "80%"}
ggplot(data = weather) +
  geom_boxplot(
    aes(y = temp, x = factor(month))
  ) +
  theme_economist() #<<
```
]

.right5[

```{r t-econ-custom-f, out.width = "80%"}
ggplot(data = weather) +
  geom_boxplot(
    aes(y = temp, x = factor(month))
  ) +
  my_theme #<< 
```
]

  ]

.panel[.panel-name[theme_set()]

.left-full[

If you would like to apply your theme to all the figures you generate, then use `theme_set()` like below:

```{r eval = F}
theme_set(my_theme)
```

After this, all of your figures will follow  `my_theme`.

]

  ] <!-- panel ends here -->

]


<!-- 
#=========================================
# Extensions    
#=========================================
-->
  
---

class: inverse, center, middle
name: gallery

# Gallery of other type of figures

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

---


# Pie charts

.panelset[ 

.panel[.panel-name[Instruction]

.left-full[

+ You can use `geom_arc_bar()` from the `ggforce` package to create a pie chart

```{r }
library(ggforce)
```

+ First create a dataset where the starting and ending angles based on the share for each group (see next slide)
+ Then use `geom_arc_bar()` to create a pie chart

]

  ]

.panel[.panel-name[Prepare a dataset]

.left-code[

```{r prep-data, results = "hide"}  
share_data <- county_yield %>% 
  select(state_name, corn_yield) %>% 
  group_by(state_name) %>% 
  summarize(yield = mean(corn_yield, na.rm = T)) %>% 
  mutate(yield_ratio = yield/sum(yield)) %>% 
  mutate(
    end_angle = 2*pi*cumsum(yield_ratio),
    start_angle = lag(end_angle, default = 0),
    mid_angle = (end_angle + start_angle)/2
  )
```

]

.right-plot[
```{r echo = F}  
share_data
```
]

  ]

.panel[.panel-name[pie chart]

.left-code[

+ You can use the following code as a template to create a pie chart 
+ Replace `state_name` in `fill = state_name` with the variable representing the groups in your share data

```{r pie-chart, eval = F}  
#--- define two parameters ---#
rpie = 1 # pie radius
rlabel = 0.6 * rpie # radius of the labels

#--- create a pie chart ---#
share_data %>% 
  ggplot(data = .) + 
    geom_arc_bar(
      aes(x0 = 0, y0 = 0, r0 = 0, r = rpie, # fixed values
        start = start_angle, 
        end = end_angle, 
        fill = state_name
      )
    ) +
    geom_text(
      aes(
        x = rlabel * sin(mid_angle), 
        y = rlabel * cos(mid_angle), 
        label = state_name
      ), 
      hjust = 0.5, 
      vjust = 0.5
    ) +
    coord_fixed() +
    theme_void() +
    scale_fill_viridis_d("State")

```
]

.right-plot[
```{r pie-chart-f, ref.label = "pie-chart", echo = F}  

```
]

  ] 

]


---

# Correlation plot 

.panelset[ 

.panel[.panel-name[Instruction]

.left-full[

+ Correlation plots visualize the degree of correlation between variables from a correlation matrix 

+ Here, we use the `ggcorrplot` package, which is one of the packages that lets you make a correlation plot (others include the `corrplot`, `GGally` packages).

```{r cache = F}
library(ggcorrplot)
```

]

  ]

.panel[.panel-name[How]

.left-full[

+ Create a correlation matrix using `cor()` on a dataset 
+ Apply `ggcorrplot()` to the correlation matrix

]

  ] <!-- panel ends here -->

.panel[.panel-name[Example]

.left4[
```{r cor-plot, eval = F}  
weather %>% 
  na.omit() %>% 
  select(where(is.numeric)) %>% 
  select(- year) %>% 
  cor() %>% # create a cor matrix
  #--- create a cor plot ---#
  ggcorrplot(., 
    type = "lower", 
    hc.order = TRUE
  ) +
  theme(
    legend.position = "bottom"
  ) 

```

+ You can add `lab = TRUE` in `ggcorrplot()` to display the correlation coefficient numbers in the squares as well

]

.right6[
```{r cor-plot-f, ref.label = "cor-plot", echo = F, cache = F, fig.dim = c(7, 5)}  

```
]

  ]

]

---

# Diverging chart (variant of a bar chart)

.left-code[
```{r diverging2, eval = F}  
county_yield %>% 
  filter(state_name == "Nebraska") %>% 
  filter(!is.na(corn_yield)) %>% 
  group_by(county_code) %>% 
  summarize(corn_yield = mean(corn_yield)) %>% 
  mutate(
    yield_norm = (corn_yield - mean(corn_yield))/sd(corn_yield),
    below_above = ifelse(yield_norm < 0, "below", "above")
  ) %>% 
  arrange(yield_norm) %>% 
  mutate(
    county_code_f = factor(county_code, levels = .$county_code)
  ) %>% 
  ggplot(data = .) +
  geom_bar(
    aes(fill = below_above, x = county_code_f, y = yield_norm, label = county_code_f), 
    stat = 'identity', 
    width = 0.5
  )  +
  scale_fill_manual(
    name = "Productivity", 
    labels = c("Above Average", "Below Average"), 
    values = c("above" = "#00ba38", "below" = "#f8766d")
  ) + 
  coord_flip() +
  labs(x = "Normalized Yield", y = "County Code") +
  theme(
    legend.position = "bottom",
    axis.text.y = element_text(size = 6)
  )

```
]

.right-plot[
```{r diverging2-f, ref.label = "diverging2", echo = F, fig.dim = c(4, 4.3), out.width = "90%", dependson = "diverging"}  

```
]

---

# Area chart 

.left-code[

Area charts are useful when comparing a variable across multiple groups over time.

```{r area-chart, eval = F}  
county_yield %>% 
  filter(!is.na(corn_yield)) %>% 
  group_by(state_name, year) %>% 
  summarize(corn_yield = mean(corn_yield)) %>% 
  group_by(year) %>% 
  mutate(corn_yield_p = corn_yield / sum(corn_yield)) %>% 
  ggplot(data = .) + 
  geom_area(
    aes(
      y = corn_yield_p, 
      x = year, 
      fill = state_name
    ), 
    stat = "identity"
  ) + 
  labs(x = "Year", y = "Corn Yield Ratio") +
  scale_fill_viridis_d(name = "State") +  
  theme_bw() +
  theme(legend.position = "bottom")

```

+ Do not forget `stat = "identity"`

]

.right-plot[
```{r area-chart-f, ref.label = "area-chart", echo = F, out.width = "90%"}  
```
]


---

# Mean and SE chart

.left5[
```{r mean-se, eval = F}
data(Salaries, package="carData")

Salaries %>%
  group_by(rank) %>% #BREAK
  summarize(
    n = n(),
    mean_salary = mean(salary),
    sd = sd(salary),
    se = sd / sqrt(n),
    ci = qt(0.975, df = n - 1) * sd / sqrt(n)
  ) %>% #BREAK
  ggplot(data = .) + #BREAK
  aes(x = rank, y = mean_salary) + #BREAK
  geom_point(size = 3) + #BREAK
  geom_errorbar( #>>
    aes( #>>
      ymin = mean_salary - 1.96 * se, #>>
      ymax = mean_salary + 1.96 * se #>>
      ), #>>
    width = .1 #>>
  ) #>>

```
]

.right5[
```{r mean-se-f, ref.label = "mean-se", echo = F}  
```
]


---

# More supplementary `geom_*()` 

.panelset[ 

.panel[.panel-name[encircle]

.left5[

Create a dataset first:

```{r cache = F}
library(ggalt)

yield_group <- county_yield %>% 
  filter(state_name %in% c("Colorado", "Nebraska")) %>% 
  filter(!(state_name == "Colorado" & corn_yield > 130)) %>% 
  filter(!(state_name == "Nebraska" & corn_yield > 150))  
```  

Create a figure:

```{r encircle, eval = F}  
ggplot(data = yield_group) +
  geom_point(
    aes(y = corn_yield, x = d3_5_9, color = state_name)
  ) +
  geom_encircle( #<<
    aes(y = corn_yield, x = d3_5_9, color = state_name) #<<
  ) #<<

```
+ Can be useful for illustrating clusters

]

.right5[
```{r encircle-f, ref.label = "encircle", echo = F, out.width = "90%"}  

```
]


  ]

.panel[.panel-name[segment]

.left5[
```{r segment, eval = F}  
ggplot(data = yield_group) +
  geom_point(
    aes(y = corn_yield, x = d3_5_9, color = state_name)
  ) +
  geom_segment( #<<
    x = 10, #<<
    y = 50, #<<
    xend = 18.8, #<< 
    yend = 88, #<<
    arrow = arrow(length = unit(0.5, "cm")) #<<
  ) #<<
```
]

.right5[
```{r segment-f, ref.label = "segment", echo = F, out.width = "90%"}  

```
]

  ]

.panel[.panel-name[curve]

.left5[
```{r curve, eval = F}  
ggplot(data = yield_group) +
  geom_point(
    aes(y = corn_yield, x = d3_5_9, color = state_name)
  ) +
  geom_curve( #<<
    x = 10, #<<
    y = 50, #<<
    xend = 18.8, #<< 
    yend = 88, #<<
    curvature = 0.2, #<<
    arrow = arrow(length = unit(0.5, "cm")) #<<
  ) #<<
```
]

.right5[
```{r curve-f, ref.label = "curve", echo = F, out.width = "90%"}  

```
]

  ]

]

<!-- 
#=========================================
# Animation    
#=========================================
-->
  
---

class: inverse, center, middle
name: animated

# Animated figures 

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

---


# Animated figures using the `gganimate` package 

.panelset[ 

.panel[.panel-name[Instruction]

.left-full[

+ Install the `gganimate` package:

```{r eval = F}
install.packages(gganimate)   
``` 

+ Install the `png` and `gifski` packages as well:

```{r eval = F}
install.packages(png)   
install.packages(gifski)   
``` 

+ You need to library only the `gganimate` package

```{r }
library(gganimate)
```

]
  ]

.panel[.panel-name[How]

.left-full[

**One state a a time**:

+ Create a regular ggplot object **without** the dimension you intend to animate over 
+ add `transition_states(transition variable)` to make the `ggplot` object animated

**Reveal a state at a time**:

+ Create a regular ggplot object **with** the dimension you intend to animate over 
+ add `transition_reveal(transition variable)` to make the `ggplot` object animated

]

  ]

.panel[.panel-name[state]

.left-code[

```{r animate, eval = F}  
weather %>% 
  ggplot(data = .) +
  geom_boxplot(
    aes(y = temp, x = origin, fill = origin)
  ) + 
  transition_states(month) #<<
```
+ `weather` is from the the `nycflights13` package has daily weather information at three airports in NY.
+ Each frame has a boxplot of temperature for the three airports in NY in its corresponding month.

]

.right-plot[
```{r animate-f-1, ref.label = "animate", echo = F, out.width = "90%"}  
```
]

  ]

.panel[.panel-name[reveal]

.left-code[
```{r animate-2, eval = F}  
weather %>% 
  filter(month == 8) %>% 
  filter(day <= 10) %>% 
  ggplot(data = .) +
  geom_boxplot(
    aes(y = temp, x = factor(day), fill = origin)
  ) + 
  transition_reveal(day) #<<
```
]

.right-plot[
```{r animate-2-f, ref.label = "animate-2", echo = F, out.width = "90%"}  
```
]
  ]

]


---

# Animated figures  

.panelset[ 

.panel[.panel-name[Instruction]

.left-full[

You can also use the `plotly` package to create an animated figure. 

```{r eval = F}
install.packages(plotly)  

library(plotly)
```

+ It is very easy to create an animated figure using `plotly`
+ But, it is a bit of a hassle to integrate animated figures generated by `plotly`

]

  ] 

.panel[.panel-name[How 1]

.left-full[

+ Create a regular `ggplot` object (figure) where `frame = transition variable` is added in `aes()` along with other necessary arguments
+ Apply `ggploty()` to the `ggplot` object 
+ Save the results as an html file 
+ Import the html file and put it in an **iframe**

]

  ] 

.panel[.panel-name[How 2]

.left-full[

```{r eval = F}
#--- create a ggplot object ---#
county_yield %>% 
  filter(state_name %in% c("Nebraska", "Colorado", "Kansas")) %>% 
  ggplot(data = .) + 
  geom_boxplot(
    aes(
      y = corn_yield, 
      x = state_name,
      #--- add frame ---#
      frame = year #<<
    )
  ) +
  labs(x = "State", y = "Corn Yield (bu/acre)")

#--- apply ggplotly() to the ggplot object ---#
ggplotly(g_box, width = 800, height = 400) %>% #<< 
  #--- save as html file ---#
  htmltools::save_html(file = "g_box.html") #<<

```

Then, add this in your Rmd file:

```
<iframe src="g_box.html" width="1000" height="550" scrolling="yes" seamless="seamless" frameBorder="0"> </iframe>
```

]

  ] <!-- panel ends here -->

.panel[.panel-name[time-series box plot]

```{r echo = F, eval = F}
county_yield_3 <- county_yield %>% 
  filter(state_name %in% c("Nebraska", "Colorado", "Kansas"))

g_box <- ggplot(data = county_yield_3) + 
  geom_boxplot(
    aes(
      y = corn_yield, 
      x = state_name,
      frame = year
    )
  ) +
  labs(x = "State", y = "Corn Yield (bu/acre)")

htmltools::save_html(ggplotly(g_box, width = 800, height = 400), file = "g_box.html")

```

<iframe src="g_box.html" width="1000" height="550" scrolling="yes" seamless="seamless" frameBorder="0"> </iframe>
  ]

.panel[.panel-name[time-series scatter plot]

```{r echo = F, eval = F}
library(plotly)
library(gapminder)

test <- ggplot(gapminder, aes(gdpPercap, lifeExp, color = continent)) +
  geom_point(aes(size = pop, frame = year, ids = country)) +
  scale_x_log10()  

htmltools::save_html(ggplotly(test, width = 800, height = 400), file = "fig.html")

```

<iframe src="fig.html" width="900" height="500" scrolling="yes" seamless="seamless" frameBorder="0"> </iframe>  

  ] <!-- panel ends here -->

]


<!-- 
#=========================================
# Exporting a figure as an image    
#=========================================
-->
  
---

class: inverse, center, middle
name: inputoutput

# Exporting a figure as an image 

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

---

# Exporting a figure as an image 

.panelset[ 

.panel[.panel-name[Instruction]

.left-full[

You can use the `ggsave()` function with the following syntax:

```{r eval = F}
#--- Syntax (NOT RUN) ---#
ggsave(filename = file name, plot = ggplot object)  

#--- or just this ---#
ggsave(file name, ggplot object)  
```

## Example

```{r }
ggsave("ex_boxplot.pdf", g_box)  
```

This will save `g_box` as **ex_boxplot.pdf** in the working directory.

]

  ]

.panel[.panel-name[output file format]

.left-full[

+ Many different file formats are supported including pdf, svg, eps, png, jpg, tif, etc. One thing you want to keep in mind is the type of graphics:
  * vector graphics (pdf, svg, eps)
  * raster graphics (jpg, png, tif)

+ While vector graphics are scalable, raster graphics are not. 

+ If you enlarge raster graphics, the cells making up the figure become visible, making the figure unappealing. 

+ Unless it is required to save figures as raster graphics, it is encouraged to save figures as vector graphics.

+ **pdf** is almost always a good choice

]

  ]

.panel[.panel-name[Options]

.left-full[

## Image width and height

+ You can control the width and height of the output image using the `width` and `height` options (the default unit is inch.):

```{r }
ggsave("ex_boxplot.pdf", g_box, height = 5, width = 7)   
```

## Image resolution

+ You can control the resolution of the output image by specifying DPI (dots per inch) using the dpi option. 

+ The default DPI value is 300, but you can specify any value suitable for the output image, including retina (320) or screen (72). 

+ 600 or higher is recommended when a high resolution output is required.

```{r eval = F}
#--- dpi = 320 ---#
ggsave("nc_dpi_320.png", g_nc, height = 5, width = 7, dpi = "retina")

ggsave("nc_dpi_600.png", g_nc, height = 5, width = 7, dpi = 600)
```

]

  ] 

]



<!-- 
#=========================================
# Resources    
#=========================================
-->
  
---

class: inverse, center, middle
name: inputoutput

# Resources 

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

---

# Resources

## Books and tutorials

+ [ggplot2: Elegant Graphics for Data Analysis](https://ggplot2-book.org/)
+ [Data Visualization with R](https://rkabacoff.github.io/datavis/)
+ [ggplot2 tutorial by Cedric Scherer](https://cedricscherer.netlify.app/2019/08/05/a-ggplot2-tutorial-for-beautiful-plotting-in-r/)
+ [Interactive html document](https://plotly-r.com/index.html)
+ [R Graphics Cookbook, 2nd edition](https://r-graphics.org/)

## Packages
+ [ggplot2 extensions](https://exts.ggplot2.tidyverse.org/)
+ [gganimat](https://exts.ggplot2.tidyverse.org/)



