
---


```{r fig-scatter-f, cache = F, include = F}
g_fig + #BREAK
  geom_point( 
    aes(x = d3_5_9, y = corn_yield), 
    color = "red", #BREAK2
    size = 0.7, #BREAK3
    shape = 0 #BREAK4
  )
```

`r chunk_reveal(
  chunk_name = "fig-scatter-f", 
  title = "Scatter Plott",
  break_type = "non_seq", 
  display_type = "both", 
  left_assign = FALSE,
  width = c(50, 50)
  )
`


```{r taste, cache = F, include = F}
tibble(mtcars) %>% #BREAK
  ggplot(data = .) + #BREAK2 
  aes(x = hp, y = mpg) + #BREAK3
  geom_point(aes(color = factor(cyl)), size = 3) + #BREAK4
  geom_smooth(method = "lm", color = "black") + #BREAK5
  facet_wrap(. ~ cyl) + #BREAK6
  scale_color_viridis_d(name = "Number of Cylinders") + #BREAK7
  labs(x = "Horse Power", y = "Mile per Gallon") + #BREAK8
  theme(legend.position = "bottom") #BREAK9
```

`r chunk_reveal(
  chunk_name = "taste", 
  title = "# Taste of ggplot2",
  break_type = "non_seq", 
  display_type = "both", 
  left_assign = FALSE,
  width = c(50, 50)
  )
`

---


```{r taste, cache = F, include = F}
county_yield %>%#BREAK
  ggplot(data = .) + #BREAK2
  aes(x = factor(year), y = corn_yield) + #BREAK3
  geom_boxplot(aes(fill = state_name)) + #BREAK4
  facet_grid(state_name ~ .) + #BREAK5
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90)) + #BREAK6
  xlab("Year") + #BREAK7
  scale_y_continuous(name = "Corn Yield (bu/acre)", limit = c(100, 200)) +  #BREAK8
  scale_fill_viridis_d(name = "State", guide = guide_legend(title.position = "top")) + #BREAK9
  ggtitle("Corn yield (bu/acre) over the \n past 20 years by State") #BREAK10

```

`r chunk_reveal(
  chunk_name = "taste", 
  title = "# Taste of ggplot2",
  break_type = "non_seq", 
  display_type = "both", 
  left_assign = FALSE,
  width = c(50, 50)
  )
`

---

```{r fig-scatter-f, cache = F, include = F}
g_fig + #BREAK
  geom_point( 
    aes(x = d3_5_9, y = corn_yield), 
    color = "red", #BREAK2
    size = 0.7, #BREAK3
    shape = 0 #BREAK4
  )
```

`r chunk_reveal(
  chunk_name = "fig-scatter-f", 
  title = "# Scatter Plot",
  break_type = "non_seq", 
  display_type = "both", 
  left_assign = FALSE,
  width = c(50, 50)
  )
`

+ `color = "red"`: makes all the squares red
+ `size = 0.5`: makes the size of the squares smaller 
+ `shape = 0`: change the shape of the points (find other shapes [here](http://www.sthda.com/english/wiki/ggplot2-point-shapes))

---

```{r hist-f, cache = F, include = F}  
g_fig + #BREAK
  geom_histogram(
    aes(x = corn_yield), 
    color = "blue", #BREAK2
    fill = "green", #BREAK3
    size = 2, #BREAK4 
    shape = 2 #BREAK5 
  )
```

`r chunk_reveal(
  chunk_name = "hist-f", 
  title = "# Histogram",
  break_type = "non_seq", 
  display_type = "both", 
  left_assign = FALSE,
  width = c(50, 50)
  )
`
+ `color = "blue"`: makes all the boundary of the bars blue
+ `fill = "green"`: makes the fill of the bars green
+ `size = 2`: makes the line width of the boundary of the bars thicker 
+ `shape = 2`: does nothing
 

---


```{r box-f, cache = F, include = F}
ggplot(data = county_yield) + #BREAK
  geom_boxplot(
    aes(x = factor(year), y = corn_yield),
    color = "red", #BREAK2
    fill = "orange", #BREAK3
    size = 0.2, #BREAK4
    shape = 1 #BREAK5
    )
```

`r chunk_reveal(
  chunk_name = "box-f", 
  title = "# Box Plot",
  break_type = "non_seq", 
  display_type = "both", 
  left_assign = FALSE,
  width = c(50, 50)
  )
`

+ `color = "blue"`: makes all the boundary of the boxes red
+ `fill = "orange"`: makes the fill of the boxes orange
+ `size = 0.2`: makes the line width of the boundary of the boxes thinner 
+ `shape = 1`: does nothing

---

```{r line-f, cache = F, include = F}
ggplot(data = mean_yield) + #BREAK1
  geom_line(
    aes(x = year, y = corn_yield), 
    color = "blue", #BREAK2
    size = 1.5, #BREAK3
    fill = "red", #BREAK4  
    linetype = "dotted", #BREAK5
  )
```

`r chunk_reveal(
  chunk_name = "line-f", 
  title = "# Line Plot",
  break_type = "non_seq", 
  display_type = "both", 
  left_assign = FALSE,
  width = c(50, 50)
  )
`

+ `color = "blue"`: makes the line blue
+ `size = 1.5`: makes the line thicker 
+ `fill = "red"`: does nothing
+ `linetype = "dotted"`: makes the line dotted

---

