---
title: "Making Regression and Summary Tables with `modelsummary`"
author: "AECN 396/896-002"
output:
  xaringan::moon_reader:
    # css: [default, metropolis, metropolis-fonts] 
    css: xaringan-themer.css 
    lib_dir: libs
    nature:
      ratio: 16:10
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r, child = './../setup.Rmd'}
```

```{r, include = F, cache = F}
suppressMessages(library(nycflights13)) 
suppressMessages(library(modelsummary)) 
suppressMessages(library(flextable)) 
suppressMessages(library(sandwich)) 
suppressMessages(library(gt)) 
suppressMessages(library(tidyverse)) 
```


# Before you start

## Learning objectives

The objectives of this lecture is to learn how to use the `modelsummary` package to create regression and summary tables that are of publication quality. 

Here is the [modelsummary package website](https://vincentarelbundock.github.io/modelsummary/).

## Table of contents

1. [Create regression tables](#reg-table)
2. [Create summary tables](#summary-table)

<br>

<span style="color:red"> Tips: </span>hitting "o" key will give you a panel view of the slides

---

# The datasets we use

```{r eval = F, echo = F}
setwd("~/Dropbox/TeachingUNL/DataScience/LectureNotes/Chapter-6-MakeTable") 
```

.panelset[ 

.panel[.panel-name[Instruction]
Go [here](https://www.dropbox.com/sh/63rlp4ydmyjm1ui/AACYSeN0f_WAgKPQKzgpGVe0a?dl=0) and download **county_yield.rds** and then read the file onto R:

```{r dataset, results = "hide", cache = F}
county_yield <- readRDS("county_yield.rds") %>% 
  tibble() %>% 
  select(soy_yield, corn_yield, year, county_code, state_name, d0_5_9, d1_5_9, d2_5_9, d3_5_9, d4_5_9) %>% 
  filter(state_name %in% c("Nebraska", "Kansas", "Colorado"))
```

]

.panel[.panel-name[Variable Definitions]
+ `soy_yield`: soybean yield (bu/acre)
+ `corn_yield`: corn yield (bu/acre)
+ `d0_5_9`: ratio of weeks under drought severity of 0 from May to September
+ `d1_5_9`: ~ drought severity of 1 from May to September   
+ `d2_5_9`: ~ drought severity of 2 from May to September   
+ `d3_5_9`: ~ drought severity of 3 from May to September   
+ `d4_5_9`: ~ drought severity of 4 from May to September   

  ]

]


<!-- 
#=========================================
# Create regression tables  
#=========================================
-->

---

class: inverse, center, middle
name: reg-table

# Create regression tables with the `modelsummary` package 

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

---

```{r include = F}
model_1_corn <- lm(corn_yield ~ d1_5_9 + d2_5_9, data = county_yield)
vcov_1_corn <- vcovHC(model_1_corn)
model_2_corn <- lm(corn_yield ~ d1_5_9 + d2_5_9 + d3_5_9 + d4_5_9, data = county_yield)
vcov_2_corn <- vcovHC(model_2_corn)
model_1_soy <- lm(soy_yield ~ d1_5_9 + d2_5_9, data = county_yield)
vcov_1_soy <- vcovHC(model_1_soy)
model_2_soy <- lm(soy_yield ~ d1_5_9 + d2_5_9 + d3_5_9 + d4_5_9, data = county_yield)
vcov_2_soy <- vcovHC(model_2_soy)
```


```{r basics-ms, cache = F, include = F}
list(
  "Corn 1" = model_1_corn, 
  "Corn 2" =  model_2_corn, 
  "Soy 1" = model_1_soy, 
  "Soy 2" = model_2_soy
) %>% #BREAK
  msummary(
    title = "Awesome Regression Results", #BREAK2
    coef_map = c("d1_5_9" = "DI: category 1", "d2_5_9" = "DI: category 2", "d3_5_9" = "DI: category 3", "d4_5_9" = "DI: category 4", "(Intercept)" = "Constant"), #BREAK3
    notes = list("A note"), #BREAK4
    fmt = "%.2f", #BREAK5
    statistic_override = list(vcov_1_corn, vcov_2_corn, vcov_1_soy, vcov_2_soy), #BREAK6
    stars = TRUE, #BREAK7
    gof_omit = "IC|Log|Adj", #BREAK8
  )

```

`r chunk_reveal(
  chunk_name = "basics-ms", 
  title = "# Taste of modelsummary",
  break_type = "non_seq", 
  display_type = "both", 
  left_assign = FALSE,
  width = c(50, 50)
  )
`

---

```{r basics-ms-gt, cache = F, include = F}
list(
  "Corn 1" = model_1_corn, 
  "Corn 2" =  model_2_corn, 
  "Soy 1" = model_1_soy, 
  "Soy 2" = model_2_soy
) %>% 
  msummary(
    output = "gt",
    title = "Awesome Regression Results", 
    coef_map = c("d1_5_9" = "DI: category 1", "d2_5_9" = "DI: category 2", "d3_5_9" = "DI: category 3", "d4_5_9" = "DI: category 4", "(Intercept)" = "Constant"), 
    notes = list("A note"), 
    fmt = "%.2f", 
    statistic_override = list(vcov_1_corn, vcov_2_corn, vcov_1_soy, vcov_2_soy), 
    stars = TRUE, 
    gof_omit = "IC|Log|Adj", 
  ) %>%  #BREAK
  tab_spanner(
    label = "Corn",
    columns = vars("Corn 1", "Corn 2")
  ) %>% #BREAK
  tab_spanner(
    label = "Soy",
    columns = vars("Soy 1", "Soy 2")
  ) %>% #BREAK
  tab_style(
    style = cell_text(color = 'red'),
    locations = cells_body(rows = 7)
  )

```

`r chunk_reveal(
  chunk_name = "basics-ms-gt", 
  title = "# Taste of modelsummary and gt",
  break_type = "user", 
  display_type = "both", 
  left_assign = FALSE,
  width = c(50, 50)
  )
`

---

# Create a regression table 

.panelset[ 

.panel[.panel-name[prepare]

Let's first run regressions which we are going to report in tables.

```{r eval = F}
model_1_corn <- lm(corn_yield ~ d1_5_9 + d2_5_9, data = county_yield)
model_2_corn <- lm(corn_yield ~ d1_5_9 + d2_5_9 + d3_5_9 + d4_5_9, data = county_yield)
model_1_soy <- lm(soy_yield ~ d1_5_9 + d2_5_9, data = county_yield)
model_2_soy <- lm(soy_yield ~ d1_5_9 + d2_5_9 + d3_5_9 + d4_5_9, data = county_yield)
```

Get White-Huber robust variance-covariance matrix for the regressions:

```{r }
library(sandwich)

vcov_1_corn <- vcovHC(model_1_corn)
vcov_2_corn <- vcovHC(model_2_corn)
vcov_1_soy <- vcovHC(model_1_soy)
vcov_2_soy <- vcovHC(model_2_soy)
```

  ]

<!-- panel ends here -->

.panel[.panel-name[default table]

.left-code[
You can supply a list of regression results to msummary() to create a default regression table.

```{r model-sum-default, eval = F}  
msummary(
  list(
    model_1_corn, 
    model_2_corn, 
    model_1_soy, 
    model_2_soy
  )
)
```

]

.right-plot[
```{r model-sum-default-f, ref.label = "model-sum-default", echo = F}  

```
]


  ]

<!-- panel ends here -->

.panel[.panel-name[modify]

`msummary()` offers multiple options to modify the default regression table to your liking:

+ `title`: put a title to the table
+ `stars`: place significance symbols (and modify the symbol placement rules)
+ `coef_map`: change the order and label of variable names  
+ `notes`: add footnotes
+ `fmt`: change the format of numbers 
+ `statistic`: type of statistics you display along with coefficient estimates
+ `gof_map`: define which model statistics to display 
+ `gof_omit`: define which model statistics to omit from the default selection of model statistics
+ `add_rows`: add rows of arbitrary contents to the table

  ] 

<!-- panel ends here -->

.panel[.panel-name[stars]

.left5[

Add `stars = TRUE` in `msummary()` to add significance markers.

You can modify significance levels and markers by supplying a named vector with its elements being the significance levels and their corresponding names being the significance markers.

**Example**:


```{r stars, eval = F}
#--- create a named vector ---#
stars_label <- c("+" = 0.1, "&+" = 0.05, "+*+" = 0.01)

#--- create a table ---#
msummary(model_1_corn, stars = stars_label)
```

]


.right5[

```{r ref.label = "stars", echo = F}

```

]

  ] 

  <!-- panel ends here -->

.panel[.panel-name[coef_map]

.left-code[
`coef_map` allows you to reorder coefficient rows and change their labels.  

Similarly with the `stars` option, you supply a named vector where its names are the existing labels and their corresponding elements are the new labels. 

In the table, the coefficient rows are placed in the order they are ordered in the named vector.     

```{r coef, eval = F}
#--- define a coef_map vector ---#
coef_map_vec <- c(
  "d1_5_9" = "DI: category 1", 
  "d2_5_9" = "DI: category 2", 
  "d3_5_9" = "DI: category 3", 
  "d4_5_9" = "DI: category 4", 
  "(Intercept)" = "Constant"
) 

#--- create a table ---#
msummary(
  list(model_2_corn, model_2_soy), 
  coef_map = coef_map_vec
)

```
]

.right-plot[
```{r coef-f, ref.label = "coef", echo = F}  

```
]



  ] 

<!-- panel ends here -->

.panel[.panel-name[coef_omit]

.left-code[

`coef_omit()` lets you omit coefficient rows from the default selections.

You supply a vector of strings (and/or regular expressions), and coefficient rows that match the string pattern will be omitted.

**Example**
```{r coef-omit, eval = F}  
msummary(
  list(model_2_corn, model_2_soy), 
  coef_omit ="d2"
)
```

`d2` matches with `d2_5_9`, and rows associated with the coefficients on `d2_5_9` are removed.

]

.right-plot[
```{r coef-omit-f, ref.label = "coef-omit", echo = F}  


```
]

  ] 

  <!-- panel ends here -->

.panel[.panel-name[gof_omit]

.left-code[

`gof_omit()` lets you omit model statistics like $R^2$ from the default selections.


You supply a vector of strings (and/or regular expressions), and statistics that match the string pattern will be omitted.

**Example**

```{r omit-stat, eval = F}  
msummary(
  list(model_2_corn, model_2_soy), 
  gof_omit ="IC|Adj"
)
```

`IC` matches with `AIC` and `BIC`, and `Adj` matches with `R2 Adj`

]

.right-plot[
```{r omit-stat-f, ref.label = "omit-stat", echo = F}  

```
]

  ] 

.panel[.panel-name[add_rows]

.left-code[

`add_rows()` can be used to insert arbitrary rows into a table. 

Adding rows using `add_rows()` involves:

+ Creating a `data.frame` (or `tibble`) to insert

+ Tell which rows you will inset the `data.frame` by `attr(data.frame, "position") <- row number`. 

```{r add-rows, eval = F}  
#--- create a table (data.frame) to insert ---#
(
rows <- data.frame(
  c1 = c("FE: County", "FE: Year"),
  c2 = c("Yes", "Yes"),
  c3 = c("No", "Now")
  )
)

#--- tell where to insert ---#
attr(rows, "position") <- c(3, 4)

#--- create a table with rows inserted ---#
msummary(
  list(Moddel1 = model_2_corn, Model2 = model_2_soy), 
  gof_omit ="IC|Adj",
  coef_omit = "d",
  add_row = rows #<<
)

```
]

.right-plot[
```{r add-rows-f, ref.label = "add-rows", echo = F}  

```
]

  ] 

  <!-- panel ends here -->

.panel[.panel-name[fmt]


**Example**:

.left-code[

`fmt()` option lets you format numbers in the table. It follows the `sprintf` C-library standard. 

By far the most common use cases is to change the number of digits after the decimal point.

For this purpose, we do not really have to understand how the `sprintf` standard works.

You can simply replace `x` in `%.xf` with the number of digits.

```{r format, eval = F}  
msummary(
  list(Moddel1 = model_2_corn, Model2 = model_2_soy), 
  gof_omit ="IC|Adj",
  coef_omit = "d",
  add_row = rows,
  #--- 2 digits after the decimal point ---#
  fmt = "%.2f" #<<
)
```

]

.right-plot[
```{r format-f, ref.label = "format", echo = F}  

```
]


  ] 

  <!-- panel ends here -->

]
<!-- panel set ends here -->

---

# Swapping the default VCOV with another VCOV

.panelset[ 

.panel[.panel-name[Instruction]

.left-full[

It is often the case that we replace the default variance-covariance matrix with a robust one for valid statistical testing.

You can achieve this using the `statistic_override` option. You will give a list of variance-covariance matrices in the order their corresponding regression results appear on the table.

**Syntax:**
```{r eval = F}
statistic_override = list(vcov_1, vcov_2, ...)
```

]

  ]

<!-- panel ends here -->

.panel[.panel-name[Compare]

.left5[

Default: 

```{r }
msummary(
  list(Moddel1 = model_2_corn, Model2 = model_2_soy), 
  gof_omit = "IC|R",
  coef_omit = "d3|d4",
  #--- no statistical override ---#
)  
```
]

.right5[

VCOV swapped:

```{r }
msummary(
  list(Moddel1 = model_2_corn, Model2 = model_2_soy), 
  gof_omit = "IC|R",
  coef_omit = "d3|d4",
  statistic_override = list(vcov_2_corn, vcov_2_soy) #<<
)  
```
]

  ]

<!-- panel ends here -->

]

<!-- panel set ends here -->


---
name: save-table

# Saving the table 

.left-full[

You can save the table to a file by providing a file name to the `output` option.

The supported file types are:

+ .html 
+ .tex 
+ .md 
+ .txt 
+ .docx, pptx
+ .png 
+ .jpg

<br>

**Example**:

The `docx` option may be particularly useful for those who want to put finishing touches on the table manually on WORD:

```{r eval = F}
msummary(
  list(Moddel1 = model_2_corn, Model2 = model_2_soy),
  output = "reg_results_table.docx" #<<
)
```

]

---
name: save-R

# Further modify regression tables with other packages

.panelset[ 

.panel[.panel-name[output type]

.left-full[

Using the `output` option in `msummary()`, you can turn the regression table into R objects that are readily modifiable by the `gt`, `kableExtra`, and `flextable` packages.


.content-box-green[**Example: gt**]

```{r }
#--- create a regression table and turn it into a gt_tbl ---#
reg_table_gt <- list(model_1_corn, model_1_soy)%>% 
  msummary(output = "gt")

#--- check the class ---#
class(reg_table_gt)
```

.content-box-green[**Example: flextable**]

```{r }
#--- create a regression table and turn it into a gt_tbl ---#
reg_table_ft <- list(model_1_corn, model_1_soy)%>% 
  msummary(output = "flextable")

#--- check the class ---#
class(reg_table_ft)
```

]

  ]
<!-- panel ends here -->

.panel[.panel-name[edit with flextable]

.left-code[

Now that the regression table created using `msummary()` with `output = "flextable"` is a `flextable` object. 

So, we can use our knowledge of the `flextable` package to further modify the regression table if you would like. 

For the details of how to use the `flextable` package visit the `flextable` lecture notes. 

Here I will just given you an example of the use of `flextable` operations.

**Example**

```{r edit-ft, eval = F}  
list(
  "Corn 1" = model_1_corn, 
  "Corn 2" =  model_2_corn, 
  "Soy 1" = model_1_soy, 
  "Soy 2" = model_2_soy
) %>% 
msummary(
  output = "flextable",
  gof_omit ="IC|Adj",
) %>%  
bold(i = 9, j = c(3, 5), bold = TRUE) %>% 
color(i = 3, j = 2, color = "red")

```
]

.right-plot[
```{r edit-ft-f, ref.label = "edit-ft", echo = F}  

```
]

  ]
<!-- panel ends here -->

.panel[.panel-name[edit with gt]

.left-code[

Now that the regression table is a `gt_tbl` object, we can use our knowledge of the `gt` package to modify the regression table. 

For the details of how to use the `gt` package go [here](). Here I will just given you an example of the use of `gt` operations.

**Example**
```{r edit-gt, eval = F}  
list(
  "Corn 1" = model_1_corn, 
  "Corn 2" =  model_2_corn, 
  "Soy 1" = model_1_soy, 
  "Soy 2" = model_2_soy
) %>% 
  msummary(
    output = "gt",
    gof_omit ="IC|Adj",
  ) %>%  
  tab_spanner( #<<
    label = "Corn", #<<
    columns = vars("Corn 1", "Corn 2") #<<
  ) %>% #<<
  tab_style( #<<
    style = cell_text(color = 'red'), #<<
    locations = cells_body(rows = 7:8) #<<
  ) #<<
```
]

.right-plot[
```{r edit-gt-f, ref.label = "edit-gt", echo = F}  

```
]

  ]


<!-- panel ends here -->

]

<!-- panel set ends here -->


<!-- 
#=========================================
# Create regression tables  
#=========================================
-->

---

class: inverse, center, middle
name: summary-table

# Create summary tables 

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

---


# Example table with `tablesummary()` 

.left-code[
```{r exmaple-table, eval = F}  
county_yield %>% 
  filter(year %in% 2010:2012) %>% 
  datasummary(
    (Year = factor(year)) * (
      (`Corn Yield (bu/acre)` = corn_yield) + 
      (`Soy Yield (bu/acre)` = soy_yield) + 
      (`DI: category 4` = d4_5_9)
    ) ~ 
    state_name * (Mean + SD) ,
    data = .
  )  
```
]

.right-plot[
```{r exmaple-table-f, ref.label = "exmaple-table", echo = F}  
```
]


---


# `datasummary()` 

.panelset[ 

.panel[.panel-name[basics]

.left-code[

**Syntax**:

```{r eval = F}
datasummary(formula, data = dataset)
```

`formula` has two sides separated by `~` just like formula for regression. 

Variables/statistics on the left-hand side (right-hand side) comprise rows (columns).

]

.right-plot[

**Example**:

```{r }
datasummary(
  corn_yield ~ Mean, #<<
  data = county_yield
)
```

<br>

Switching the order changes the structure of the resulting table:

```{r }
datasummary(
  Mean ~ corn_yield, #<<
  data = county_yield
)
```
]

  ]

<!-- panel ends here -->

.panel[.panel-name[summary functions]


.left-code[

The `modelsummary` package offers multiple summary functions of its own:

+ `Mean`
+ `SD`
+ `Min`
+ `Max`
+ `P0`, `P25`, `P50`, `P75`, `P100`
+ `Histogram`

These functions have `na.rm = NA` hidden inside it, so they avoid having NA when simply applying their counterparts from the `base` package.

]

.right-plot[

For example, compare these two:

```{r }
datasummary(
  corn_yield ~ Mean, #<<
  data = county_yield
)
```

```{r }
datasummary(
  #--- mean from the base package ---#
  corn_yield ~ mean, #<<
  data = county_yield
)
```

]

  ]

<!-- panel ends here -->

.panel[.panel-name[user-defined function]

.left-full[

You can use a user-defined function that takes a vector of values and return a single value. 

**Example**:

```{r user-defined}  
#--- define a function ---#
MinMax <- function(x){
  paste0('[', min(x, na.rm = TRUE), ', ', max(x, na.rm = TRUE), ']')
} 

#--- use it ---#
datasummary(corn_yield ~ MinMax, data = county_yield) 
```

]

  ] 

  <!-- panel ends here -->

.panel[.panel-name[add stats/variables]

.left-full[

You can add more variables and statistics using `+`.

**Example**:

```{r plus}  
datasummary(
  corn_yield + soy_yield + d0_5_9 + d1_5_9
  ~ Mean + SD+ MinMax + Histogram, 
  data = county_yield
)
```

For each of the variables on the left-hand side, each of the statistics on the right-hand side is calculated and displayed.

]

  ] 

  <!-- panel ends here -->

.panel[.panel-name[All]

You can use `All()` to create a summary table for all the numeric variables in the dataset.

At the moment, `All()` does not work on `tibble`. So, if your dataset is `tibble`, convert it to a `data.frame` on the fly in the code like below:

**Example**:

```{r all, cache = F}  
datasummary(
  All(data.frame(county_yield)) 
  ~ Mean + SD, 
  data = county_yield
)
```

  ] 

  <!-- panel ends here -->



  <!-- panel ends here -->
]

<!-- panel set ends here -->


---


# `datasummary()` (cont.) 

.panelset[ 

.panel[.panel-name[nesting]

.left-code[

You can nest categorical variables with `*`, meaning you can get summary statistics for each value of the categorical variable (like `group_by() %>% summarize()`).

**Syntax**

```{r eval = F}
#--- single stat ---#
variable ~ category_variable * stat  

#--- multiple stats ---#
variable ~ category_variable * (stat 1 + stat 2 + ...)  
```

]

.right-plot[

**Examples**:

```{r nesting-f}  
datasummary(
  corn_yield + soy_yield + d0_5_9 + d1_5_9
  ~ state_name * (Mean + SD) + MinMax, #<< 
  data = county_yield
)
```

For each value of `state_name` (`Nebraska`, `Colorado`, `Kansas`), `Mean` and `SD` are shown for each of the variables on the left-hand side. But, `MinMax` is for the entire sample.

]

  ] 

.panel[.panel-name[deeper]

.left-full[

You can nest with multiple categorical variables by multiplying stats with multiple categorical variables. 

**Example**:

```{r deeper}  
county_yield %>% 
  filter(year %in% 2011:2012) %>% 
  filter(state_name %in% c("Kansas", "Nebraska")) %>% 
  datasummary(
    corn_yield + soy_yield + d0_5_9 + d1_5_9
    ~ factor(year) * state_name * (Mean + SD) + MinMax, #<< 
    data = .
  )
```

For each of the unique combinations of `state_name` (`Nebraska`, `Kansas`) and `year` (2011, 2012), `Mean` and `SD` are shown for each of the variables on the left-hand side. But, `MinMax` is for the entire sample.

]

  ] 

  <!-- panel ends here -->

.panel[.panel-name[renaming]

.left-full[

By default variable and statistics names are used as the labels in the table.

You can provide labels by the following syntax: `(label = variable/stat)`

**Example**:

```{r renaming}  
county_yield %>% 
  filter(year %in% 2011:2012) %>% 
  filter(state_name %in% c("Kansas", "Nebraska")) %>% 
  datasummary(
    (`Corn Yield (bu/acre)` = corn_yield) #<<
    ~ state_name * (Mean + (Std.Dev. = SD)), #<< 
    data = .
  )
```

+ `corn_yield` is labeled as `Corn Yield (bu/acre)`
+ `SD` is labeled as `Std.Dev.`

.content-box-red[Note: when you have spaces in the label, surround the label with back quotes.]

If you do not like this way of changing labels, you can always use `gt` package.

]

  ] 

  <!-- panel ends here -->

.panel[.panel-name[function arguments]

.left-full[

You can pass option arguments to the stats function by: `stat * Argument(options)`

<br>

**Example**:

```{r arguments}  
county_yield %>% 
  filter(year %in% 2011:2012) %>% 
  filter(state_name %in% c("Kansas", "Nebraska")) %>% 
  datasummary(
    corn_yield 
    ~ state_name * (mean + sd) * Arguments(na.rm = TRUE) + #<<
      quantile * Arguments(prob = 0.1, na.rm = TRUE), #<< 
    data = .
  )
```

<br>

+ `(mean + sd) * Arguments(na.rm = TRUE)` adds `na.rm = TRUE` option to `mean()` and `sd()`
+ `quantile * Arguments(prob = 0.1, na.rm = TRUE)` adds `prob = 0.1` and `na.rm = TRUE` to `quantileo()`

]

  ] 

  <!-- panel ends here -->

.panel[.panel-name[title and notes]

See the example below:

.left-full[
**Example**:

```{r title}  
county_yield %>% 
  filter(year %in% 2011:2012) %>% 
  filter(state_name %in% c("Kansas", "Nebraska")) %>% 
  datasummary(
    corn_yield 
    ~ state_name * (mean + sd) * Arguments(na.rm = TRUE) + 
      quantile * Arguments(prob = 0.1, na.rm = TRUE),  
    data = .,
    title = "A title", #<<
    notes = c("first note", "second note") #<<
  )
```
]
  ] 

  <!-- panel ends here -->

.panel[.panel-name[align columns]

You can use `align` to align columns. Available alignment are:

+ `l`: left
+ `r`: right
+ `c`: center

Inside `align()`, you provide a sequence of the option letters (e.g., `"lrcle"`)

The `n`th letter corresponds to `n`th column.

.left-full[
**Example**:

```{r align}  
county_yield %>% 
  filter(year %in% 2011:2012) %>% 
  filter(state_name %in% c("Kansas", "Nebraska")) %>% 
  datasummary(
    corn_yield 
    ~ state_name * (`This is M E A N` = mean) * Arguments(na.rm = TRUE) + 
      (`This is Q U A N T I L E` = quantile) * Arguments(prob = 0.1, na.rm = TRUE),  
    data = .,
    align = "lrlc" #<<
  )
```
]

  ] 

  <!-- panel ends here -->

.panel[.panel-name[Output]

You can use the `output` option to either export the table as a file or save it as R objects which you can further modify. 

This works exactly the same way as the `msummary()` function. 

+ Go [here](#save-table) for exporting the table as a file in various format 
+ Go [here](#save-R) for saving the table as an R object for further modifications

  ] 

  <!-- panel ends here -->
]

<!-- panel set ends here -->

---

# Convenience functions

.panelset[ 

.panel[.panel-name[categorical variables]

.left-full[

You can use `datasummary_skim()` with 'categorical' option to quickly generate a summary table for categorical variables:

**Syntax**:
```{r eval = F}
datasummary_skim(dataset, 'categorical')
```

**Example**:
```{r }
county_yield %>% 
  select(- county_code) %>% 
  data.frame() %>% 
  datasummary_skim(., 'categorical') 
```

Note: `datasummary_skim()` does work well on a `tibble` object.

]

  ] 

  <!-- panel ends here -->

.panel[.panel-name[balance table]

If your data was generated through randomized experiments (or you are using natural experiments), then `datasummary_balance()` can be very useful as it can generate a variable balance table.

**Syntax**:
```{r eval = F}
datasummary_balance(variables to summarize ~ treatment dummy)
```

+ `variables to summarize`: list of variables to summarize
+ `treatment dummy`: a dummy variable that indicates whether in the treated or control group

.left-code[

**Example**:

```{r balance-tab, eval = F}  
county_yield %>% 
  filter(state_name %in% c("Nebraska", "Kansas")) %>% 
  dplyr::select(c(state_name, where(is.numeric))) %>% 
  dplyr::select(- year) %>% 
  datasummary_balance(
    All (data.frame(.))~ state_name, #<<
    data = .
  )
```
]

.right-plot[
```{r balance-tab-f, ref.label = "balance-tab", echo = F}  

```
]

  ]

<!-- panel ends here -->

.panel[.panel-name[correlation table]

.left-full[

You can create a correlation table with `datasummary_correlation()`.

```{r }
county_yield %>% 
  filter(state_name %in% c("Nebraska", "Kansas")) %>% 
  dplyr::select(c(state_name, where(is.numeric))) %>% 
  dplyr::select(- year) %>% 
  datasummary_correlation()
```

]

  ]

<!-- panel ends here -->

]

<!-- panel set ends here -->


