---
title: "Miscellaneous Useful Functions in R"
author: "AECN 396/896-002"
output:
  xaringan::moon_reader:
    # css: [default, metropolis, metropolis-fonts] 
    css: xaringan-themer.css 
    lib_dir: libs
    nature:
      ratio: 16:10
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r, child = './../setup.Rmd'}
```

```{r, include = F, cache = F}
suppressMessages(library(nycflights13)) 
```

# Before you start

## Learning objectives

The objectives of this chapter is to learn how to use the `ggplot2` package to create figures for effective communication 

## Table of contents

1. [`ggplot2` basics](#date)
2. [Different types of figures](#dif-geoms)
3. [Placing more information in one figure](#more-info)
4. [Faceted figures](#faceting)
5. [Other supplementary `geom_*()`](#other-geoms)
6. [Make your figures presentable to others](#fine-tune)
7. [Tips](#tips)
7. [Gallery of other type of figures](#gallery)
7. [Animated figures](#animated)

<br>

<span style="color:red"> Tips: </span>hitting "o" key will give you a panel view of the slides

---

class: inverse, center, middle
name: df_tbl

# `Date` object

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---

class: inverse, center, middle
name: df_tbl

# Strings

```{r }
paste()
```

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---
