---
title: "Miscellaneous Useful Functions and Tips"
author: "AECN 396/896-002"
output:
  xaringan::moon_reader:
    # css: [default, metropolis, metropolis-fonts] 
    css: xaringan-themer.css 
    lib_dir: libs
    nature:
      ratio: 9:12
      highlightStyle: github
      highlightLines: true
      highlightSpans: true
      countIncrementalSlides: false
---

```{r, child = './../setup.Rmd'}
```

```{r, include = F, cache = F}
suppressMessages(library(nycflights13)) 
suppressMessages(library(extrafont)) 
suppressMessages(library(gt)) 
data(pizzaplace)
suppressMessages(library(lubridate)) 
suppressMessages(library(patchwork))
```

# Before you start

## Learning objectives

The objectives of this chapter is to learn various functions that are useful to make your programming efficient.   

## Table of contents

1. [Dates](#date)
2. [Strings](#string)

<br>

<span style="color:red"> Tips: </span>hitting "o" key will give you a panel view of the slides


---

# The datasets we use

Install the `gt` package if you have not:

```{r eval = F}
install.packages("gt")  
```

Load the package: 

```{r eval = F}
library(gt)
```

The `pizzaplace` dataset is available immediately after you loaded the package.

```{r dataset}
pizzaplace
```

<!-- 
#=========================================
# Dates
#=========================================
-->
  
---

class: inverse, center, middle
name: date

# Date and time

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---

# `Date` using the `lubridate` package


R has an object class called 

```{r eval = F}
install.packages(lubridate)

library(lubridate) 
```

Recording dates as an `Date` object instead of a character object has several benefits:

+ calendar math is possible with `Date` objects
+ you can `filter()` based on the chronological order of dates
+ converting date into an another format is easy

---

# `Date` operations

.panelset[ 

<!-- panel ends here -->

.panel[.panel-name[as.Date()]

Dates comes in various formats. Several of them are:

+ `2010-12-15`
+ `12/15/2010`
+ `Dec 15 10`
+ `15 December 2010`

They all represent the same date. 

We can use `as.Date()` to transform dates stored as characters into `Date`s. 

```{r eval = F}
#--- NOT RUN ---#  
as.Date(date in character, format)
```

In `format` you specify how day, month, and year are represented in the date characters you intend to convert using special symbols including:

+ **%d**: day as a number (0-31)
+ **%m**: month (00, 01, 02, $\dots$, 12)
+ **%b**: abbreviated month (Jan, $\dots$, Dec)
+ **%B**: unabbreviated month (January, $\dots$, December)
+ **%y**: 2-digit year (96 for 1996, 02 for 2002)
+ **%Y**: 4-digit year (1996, 2012)

<br>

.content-box-green[**Example**]

```{r }
#--- convert from a character to a Date ---#
(
a_date <- as.Date("Dec 15 10", format = "%b %d %y")
)

#--- check the class ---#
class(a_date)
```
 
  ]

<!-- panel ends here -->

.panel[.panel-name[lubridate]

Alternatively, you can use the `lubridate` package to easily convert dates recorded in characters into `Date`s.

Using `lubridate`, you do not need to provide the format information unlike `as.Date()`

Instead, you simply use `y` (year), `m` (month), `d` (day) in the order they appear in the dates in character.

<br>

.content-box-green[**Example**]

```{r }
ymd("2010-12-15")
mdy("12/15/2010")
mdy("Dec 15 10")
dmy("15 December 2010") 
```

  ] 

  <!-- panel ends here -->

.panel[.panel-name[format]
  
It is often the case that date values are not formatted in the way you want (e.g., when you are creating figures). 

While you can use string manipulation functions to reformat dates (which we learn next in this lecture), it is easier to just use the `format()` function.


```{r eval = F}
#--- NOT RUN ---#  
format(Date, format)
```

You can use the same rule for the `format` argument as the one we saw earlier when using `as.Date()`.

<br>

.content-box-green[**Example**]

```{r }
pizzaplace %>% 
  #--- convert to Date ---#
  mutate(date = ymd(date)) %>% 
  #--- format ---#
  mutate(date_txt_f1 = format(date, "%m/%d/%y")) %>% 
  mutate(date_txt_f2 = format(date, "%m %B %Y")) %>% 
  select(date_txt_f1, date_txt_f2)
```

  ]  
  
<!-- panel ends here -->  

.panel[.panel-name[components]

You can extract components (year, month, day) from a `Date` object using various helper functions offered by `lubridate`. 

+ `year()`: get year
+ `month()`: get month
+ `mday()`: get day of month 
+ `yday()`: get day of year
+ `wday()`: get day of week

<br>

.content-box-green[**Examples**]

```{r eval = F}
pizzaplace %>% 
  mutate(date = ymd(date)) %>%
  distinct(date) %>% 
  filter(month(date) >= 8) %>% 
  mutate(
    year = year(date),
    month = month(date),
    mday = mday(date),
    yday = yday(date),
    wday = wday(date)
  ) %>% 
  select(year, month, mday, yday, wday) %>% 
  head(10)
```

```{r echo = F}
pizzaplace %>% 
  mutate(date = ymd(date)) %>%
  distinct(date) %>% 
  filter(month(date) >= 8) %>% 
  mutate(
    year = year(date),
    month = month(date),
    mday = mday(date),
    yday = yday(date),
    wday = wday(date)
  ) %>% 
  select(date, year, month, mday, yday, wday) %>% 
  head(10) 
```

]

.panel[.panel-name[math 1]

Unlike dates in character, you can do some math on `Date` objects.

.content-box-green[**add years, months, and days to a `Date` object**]

You can use `years()`, `months()`, `days()` from the `lubridate` package to add specified years, months, and days, respectively.

```{r }
ymd("2015-05-01") + years(3)
ymd("2015-05-01") - months(3)
ymd("2015-05-01") + days(1)
```

<br>

.content-box-green[**sequence of dates**]

You can use `seq()` to create a sequence of dates, where the incremental step is defined by the `by` option.

```{r }
seq(ymd("2015-05-01"), ymd("2020-05-01"), by = "years")   

seq(ymd("2015-05-01"), ymd("2015-06-01"), by = "weeks")   

seq(ymd("2015-05-01"), ymd("2015-05-09"), by = "days")   
```  

  ] 

  <!-- panel ends here -->

.panel[.panel-name[math 2]

.content-box-green[**filter (logical evaluation)**]

```{r }
pizzaplace %>% 
  #--- convert to Date ---#
  mutate(date = ymd(date)) %>%  
  filter(date >= as.Date("2015-05-01"))
```

  ] 

  <!-- panel ends here -->
]
  
---

class: inverse, center, middle
name: string

# Strings manipulation

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---

# String manipulation 

.content-box-green[**Package**]

For string (character) manipulation, we use the `stringr` package, which is part of the `tidyverse` package. So, you have installed it already.

`stringr` is loaded automatically when you load `tidyverse`. So, just load `tidyverse`.

```{r eval = F}
library(tidyverse)  
```

<br>

.content-box-green[**Functions**]

Here are the select functions we learn in this lecture:

+ join and split
  * `str_c()`
  * `str_split()` (`dplyr::separate()`)
+ mutate strings
  * `str_replace()`
+ detect matches
  * `str_detect()`
+ manage lengths
  * `str_trim()`
  * `str_pad()`

<br>

.content-box-green[**Resources**]

+ [stringr cheatsheet](https://github.com/rstudio/cheatsheets/blob/master/strings.pdf)
+ [stringr packagedown website](https://stringr.tidyverse.org/index.html)

<br>

---

# Join (concatenate) with `str_c()`

`str_c()` lets you concatenate a vector of strings. It is basically the same as `paste()`.

.panelset[ 

.panel[.panel-name[join 1]


.content-box-green[**concatenate**]


```{r }
str_c("R", "rocks")
```

<br>

.content-box-green[**order matters**]


```{r }
str_c("rocks", "R")
```

<br>

.content-box-green[**separator**]


```{r }
str_c("R", "rocks", sep = "+")
```

<br>

.content-box-green[**more than two strings**]


```{r }
str_c("R ", "rocks", ", right", "?", sep = "")
```

  ]

<!-- panel ends here -->

.panel[.panel-name[join 2]

.content-box-green[**a string and a vector of strings**]

```{r }
verbs <- c("sucks", "rocks", "is just okay")

str_c("R", verbs, sep = "+")
```

+ Each of the vector elements (`verbs`) are concatenated with a string (`"R"`)
+ The separator (`"+"`) applied to all the vector elements

<br>

.content-box-green[**collapsing a vector of strings to a single string**]

```{r }
str_c("R", verbs, sep = "+", collapse = "%")
```

+ The `collapse` option collapse all the vector elements into a single string with the collapse separator (here, `%`) placed between the individual vector elements
+ `sep = "+"` is applied when concatenating a vector of strings and a string, and `collapse = "%"` is applied when concatenating the resulting vector of strings. 

  ]

<!-- panel ends here -->

.panel[.panel-name[join 3]


.content-box-green[**two vectors of equal length**]


```{r }
software_types <- c("R", "Python", "SAS")

str_c(software_types, verbs, sep = " ")
```

+ `n`th element of a vector (`software_types`) is met with `n` the element of the other vector (`verbs`).

  <!-- panel ends here -->


.content-box-green[**two vectors of different lengths**]


```{r }
software_types <- c("R", "Python", "SAS", "STATA", "Matlab")

str_c(software_types, verbs, sep = " ")
```

+ `n`th element of a vector (`software_types`) is met with `n` the element of the other vector (`verbs`) with `verbs` recycled for the elements in `software_type` that are missing positional matches.


.content-box-green[**all combinations**]


```{r eval = F}
software_types <- c("R", "Python", "SAS", "STATA", "Matlab")

str_c(software_types, `rep(verbs, 5)`, sep = " ")
```

```{r echo = F}
software_types <- c("R", "Python", "SAS", "STATA", "Matlab")

str_c(software_types, rep(verbs, 5), sep = " ")
```

+ take advantage of the recycling feature to create all possible combinations of values    

  ] 

  <!-- panel ends here -->

.panel[.panel-name[use case]


.content-box-green[**Concatenating string variables in a dataset**]


Sometimes, you want to concatenate two (or more) string variables into one variable. 

For example, suppose you would like to combine pizza `size` and `type` into a single variable to make it easier to create faceted figures by `size`-`type`.

```{r }
pizzaplace <- mutate(pizzaplace, type_size = str_c(type, size, sep = "-"))
```

```{r echo = F}
pizzaplace   
```  


.content-box-green[**Reading files**]


You can use `str_c()` to create a vector of file names that have a common pattern. 

For example suppose you have files that are named following this convention: "corn_yield_X.csv", where X represents year. 

You have such csv files starting from 2000 to 2020. Then,

```{r }
file_names <- str_c("corn_yield_", 2000:2020, ".csv")  
```

```{r }
head(file_names) 
```

Now, you can easily read each of them iteratively using a loop.

  ] 

  <!-- panel ends here -->

.panel[.panel-name[split]


`str_split()` splits a string based on a pattern you provide:

```{r }
str_split(pizzaplace$type_size, "-") %>% head()
```

But, if you are splitting a variable into two variables, `dplyr::separate()` is a better option.

```{r }
pizzaplace %>% separate(type_size, c("type_2", "size_2"), "-") %>% 
  select(id, type, type_2, size, size_2)
```

  ] 

  <!-- panel ends here -->
]

<!-- panel set ends here -->

---

# Mutate 

.panelset[ 

.panel[.panel-name[replace]

You can use `str_replace()` to replace parts of the texts matched with the user-specified texts.

```{r eval = F}
#--- Syntax ---#
str_replace(string, pattern, replacement)
```

<br>

.content-box-green[**Example**]:

```{r }
strings_vec <- c("I think R rocks, but some think it does not rock", "Python also rocks, but Julia does not rock")  

str_replace(strings_vec, "rock", "rock big time")
```

Note that the only the first occurrence of "rock" in each of the string vector element was replaced with "rock big time." 

You need to use `str_replace_all()` to replace all the occurrences. 


```{r }
str_replace_all(strings_vec, "rock", "rock big time") 
```

  ]

<!-- panel ends here -->
.panel[.panel-name[use case]


.content-box-green[**Formating dates**]

Suppose you would like to have a particular format of date in a figure you are trying to create using `pizzaplace`: e.g., 07/08/20 (month, day, year without the first 2 digits).

Pretend that `date_text` is the variable that indicates date and it looks like this:

```{r echo = F}
pizzaplace_for_plot <- mutate(pizzaplace, date_text = ymd(date) %>%  format("%m/%d/%Y")) 
```   

```{r }
pizzaplace_for_plot$date_text %>% head()
```

So, you would like to replace "20" with "" (nothing).

```{r }
pizzaplace_for_plot %>% 
  mutate(date_text = str_replace(date_text, "20", "")) %>% 
  select(id, date_text, size, type)
```

Now you can create a figure with the dates in the desired format. 

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

From `pizzaplace`, you could have just done this:

```{r }
pizzaplace %>% 
  mutate(
    date = ymd(date),
    date = format(date, "%m/%d/%y")
  ) %>% 
  .$date %>% head()
``` 

  ]

<!-- panel ends here -->

]

<!-- panel set ends here -->


---

# Detect 

.panelset[ 

.panel[.panel-name[detect]

You can use `str_detect()` to check if a user-specified texts are part of strings.

It takes a vector of strings and a text pattern, and then return a vector of TRUE/FALSE.

<br>

.content-box-green[**Example**]

```{r }
fruit <- c("apple", "banana", "pear", "pineapple")

str_detect(fruit, "apple")   
``` 

  ]

<!-- panel ends here -->

.panel[.panel-name[use case 1]

.content-box-green[**get the list of file names** ]

Suppose you have two sets of files in a single folder: `corn_experiment_x.rds` and `soy_experiment_y.rds`, where both `x` and `y` range from 1 to 50.

You want to read only the soy files. 

First, let's get the name of the whole list of files in the working directory:

```{r eval = F}
all_files <- list.files("../../Datasets/Chapter_5/experiment_data")
```

```{r echo = F}
all_files <- list.files("../../Datasets/Chapter_5/experiment_data")
```

```{r }
head(all_files)
tail(all_files) 
```

Now use `str_detect()` to find which elements of `all_files` include "soy."

```{r }
is_soy <- str_detect(all_files, "soy")  
```

Okay so, here is the list of all the "soy" files:

.scroll-box-12[
```{r }
all_files[is_soy]  
```
]


Now, you can loop to read all the files. 

  ]

.panel[.panel-name[use case 2]

.content-box-green[**Define a group from a varible** ]

Consider the following dataset of plant genes.

```{r }
gene_data <- expand.grid(
  id = c("Zm_1", "Zm_2"), 
  gene = c("20_WW_BL_TP1", "20_WW_BL_TP", "20_WW_ML_TP1", "20_WW_ML_TP", "20_WW_TL_TP1", "20_WW_TL_TP3")
) 
```

```{r echo = F}
gene_data 
```

There are three different types of genes: those that have `_BL_`,`_ML_`, and `_TL_`. The objective here is to make a variable that indicates gene group from the `gene` variable.

```{r }
gene_data %>% 
  mutate(gene_group = case_when(
    str_detect(gene, "_BL_") ~ "BL",
    str_detect(gene, "_ML_") ~ "ML",
    str_detect(gene, "_TL_") ~ "TL"
  ))
```

  ] 

  <!-- panel ends here -->
]

<!-- panel set ends here -->

---

# Other convenient functions 

.panelset[ 

.panel[.panel-name[letter case]

Here are the collection of functions that let you change the letter case of strings.

.content-box-green[**To upper case**]

```{r eval = F}
pizzaplace %>% 
  mutate(name = str_to_upper(name))
```

.scroll-box-10[
```{r echo = F}
pizzaplace %>% 
  mutate(name = str_to_upper(name))
```
]

.content-box-green[**To lower case**]

```{r eval = F}
pizzaplace %>% 
  mutate(size = str_to_lower(size))
```

.scroll-box-10[
```{r echo = F}
pizzaplace %>% 
  mutate(size = str_to_lower(size))
```
]

.content-box-green[**Only the first letter is capitalized**]

```{r eval = F}
pizzaplace %>%  
  mutate(type = str_to_title(type)) 
```

.scroll-box-10[
```{r echo = F}
pizzaplace %>%  
  mutate(type = str_to_title(type)) 
```
]

  ]

<!-- panel ends here -->

.panel[.panel-name[padding]

You can pad strings with symbols of your choice so the resulting string are of the length you specify.   

```{r eval = F}
#--- NOT RUN ---#
str_pad(strings, string length, side, padding symbol)
```

<br>

.content-box-green[**Examples**]

```{r }
str_pad("I am sick of R", 20, side = "left", pad = "?") 
```

```{r }
str_pad("I am sick of R", 30, side = "both", pad = "?") 
```

```{r }
str_pad("I am sick of R", 20, side = "right", pad = "-") 
```


  ]

<!-- panel ends here -->

]

<!-- panel set ends here -->


---

# Exercises 


.content-box-green[**Exercise 1**]

Create the following dataset:

```{r }
date_data <- tibble(
  date = seq(ymd("2012-04-01"), ymd("2012-09-30"), by = "days")
) %>%    
  mutate(
    year = year(date),
    month = month(date),
    day = mday(date)
  ) %>% 
  select(- date)
```

Use `str_c()` to combine, `year`, `month`, and `day` using "-" as the separator and convert the combined text to Date using `lubridate`.

<br>

.content-box-green[**Exercise 2**]

Create the following dataset:

```{r }
date_data <- tibble(
  date = seq(ymd("2012-04-01"), ymd("2012-09-30"), by = "days")
) %>%    
  mutate(
    year = year(date),
    day_of_year = yday(date)
  ) %>% 
  select(- date)
```

Using `Date` math to recover the dates from `year` and `day_of_year`. 


