---
title: "Making Tables with `flextable`"
author: "AECN 396/896-002"
output:
  xaringan::moon_reader:
    # css: [default, metropolis, metropolis-fonts] 
    css: xaringan-themer.css 
    lib_dir: libs
    nature:
      ratio: 16:10
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r, child = './../setup.Rmd'}
```

```{r, include = F, cache = F}
suppressMessages(library(nycflights13)) 
suppressMessages(library(gt)) 
suppressMessages(library(tidyverse)) 
suppressMessages(library(flextable)) 
suppressMessages(library(printr))
suppressMessages(library(officer)) 
```

# Before you start

## Learning objectives

The objectives of this chapter is to learn how to use the `flextable` package to create publication-quality tables.

## Table of contents

1. [Create and modify tables: the basics](#create-modify)
2. [Formatting values in tables](#formating)
3. [Modify by column](#columns)
4. [Save tables as a file](#save)

<br>

<span style="color:red"> Tips: </span>hitting "o" key will give you a panel view of the slides


<!-- 
#=========================================
# gt Basics
#=========================================
-->

---

class: inverse, center, middle

# Create tables with the `flextable` package 

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

---

```{r taste-flextable, cache = F, include = F}
head(mtcars, 20) %>% #BREAK
  select(am, carb, gear, mpg, drat) %>% #BREAK
  flextable() %>% #BREAK
  theme_vanilla() %>% #BREAK 
  set_header_labels(mpg = "miles per gallon") %>% #BREAK 
  autofit() %>% #BREAK 
  merge_v(j = c("am", "carb")) %>% #BREAK 
  italic(j = 1) %>% #BREAK 
  bg(bg = "#C90000", part = "header") %>% #BREAK 
  color(color = "blue", j = 5) %>% #BREAK 
  color(color = "red", i = 5:10, j = 2) %>% #BREAK 
  color(color = "white", part = "header") %>% #BREAK 
  bold(~ drat > 3.2, ~ gear, bold = TRUE) %>% #BREAK 
  align(j = 1) %>% #BREAK
  fontsize(i = 12:18, size = 8) %>% #BREAK
  add_footer_row(values = "blah blah", colwidths = 5) %>% #BREAK
  border_outer(fp_border(color="red", width = 2)) %>% #BREAK 
  line_spacing(space = 1.5) %>% #BREAK
  theme_tron()  
  
```

`r chunk_reveal(
  chunk_name = "taste-flextable", 
  title = "# Taste of the gt package",
  break_type = "user", 
  display_type = "both", 
  left_assign = FALSE,
  width = c(40, 60)
  )
`
  
<!-- 
#=========================================
# Create or Modify Parts  
#=========================================
-->
  
---

class: inverse, center, middle
name: create-modify

# Create table and modify Parts  

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1000px></html>

---


# Initiate a table 

.panelset[ 

.panel[.panel-name[Instruction]

.content-box-green[**Before we start**]

Install the `flextable` package and library it.

```{r eval = F}
install.pacakges(flextable) 
```

```{r eval = F}
library(flextable)  
```

.content-box-red[**How**]

+ We can apply `flextable()` to a `data.frame` to initiate a table. 

+ We can then modify its part using many functions offered by the `flextable` package

  ] 

  <!-- panel ends here -->

.panel[.panel-name[Data Preparation]

.left4[
```{r results = "hide"}
#--- Define regions ---#
Australasia <- c("AU", "NZ")
Melanesia <- c("NC", "PG", "SB", "VU")
Polynesia <- c("PF", "WS", "TO", "TV")

#--- create a dataset ---#
tab_data <- countrypops %>%
  filter(country_code_2 %in% c(
    Australasia, Melanesia, Polynesia)
  ) %>%
  filter(year %in% c(1995, 2005, 2015)) %>%
  mutate(region = case_when(
    country_code_2 %in% Australasia ~ "Australasia",
    country_code_2 %in% Melanesia ~ "Melanesia",
    country_code_2 %in% Polynesia ~ "Polynesia",
  )) %>%
  pivot_wider(
    values_from = population, 
    names_from = year,
    names_prefix = "y_"
  ) %>%
  arrange(region, desc(y_2015)) %>%
  select(-starts_with("country_code")) %>% 
  mutate(
    pop_ratio_10_15 = y_2015/y_2005,
    date = "2013-11-14"
  )  

```
]

.right6[
```{r echo = F}
tab_data 
```
]

  ] 

  <!-- panel ends here -->
.panel[.panel-name[Initiate]

.left4[
```{r eval = F}
flextable(tab_data, 
  col_keys = c(
    "country_name", 
    "lufhierh", 
    "region", 
    "y_1995", 
    "y_2005")
)
```

+ `col_keys`: the name of the variables from the `data.frame` (providing variable names that do no exist in the dataset creates blank columns) 

]

.right6[
```{r echo = F}
flextable(tab_data, col_keys = c("country_name", "lufhierh", "region", "y_1995", "y_2005")) 
```
]

]

]

<!-- panel set ends here -->

---

# Selector

.panelset[ 
 
 .panel[.panel-name[Instruction]

Many functions let you choose specifically where you apply changes. Those functions have 

+ `i` for rows selection 
  * formula 
  * integer vector 
  * logical vector
+ `j` for cols selection
  * formula
  * character vector
  * integer vector
 
You can use any combinations of the reference methods for `i` and `j`.

   ]
 
 <!-- panel ends here -->
 
 .panel[.panel-name[i, j: formula]


.left5[
<br>
.content-box-green[**Syntax**]

```{r eval = F}
function(
  i = ~ condition based on variables,
  j = ~ vairable 1 + variable 2 + ...,
  .
)  
```

.content-box-red[**Example**]

```{r i_formula, eval = F}
tab_data %>% 
  flextable(
    col_keys = c("country_name", "region", "y_1995", "y_2005")
  ) %>% 
  color(
    i = ~ y_2005 > 2e5, #<<
    j = ~ country_name + region, #<<
    color = "red"
  )
```
]

.right5[

<br>

```{r i_formula-t, ref.label = "i_formula", echo = F}
```
]

 
   ]
 
 <!-- panel ends here -->

.panel[.panel-name[i, j: integer, character]

.left-code[
<br>
```{r i-c, eval = F}  
tab_data %>% 
  flextable(
    col_keys = c("country_name", "region", "y_1995", "y_2005")
  ) %>% 
  color(
    i = c(2, 4:9), #<<
    j = c("country_name", "y_1995"), #<<
    color = "red"
  )
```

Using a character vector for `j` is not recommended because using a formula involves less typing.

]
 
.right-plot[
<br>
```{r i-c-f, ref.label = "i-c", echo = F}  

```
]
  
   ] 
 
   <!-- panel ends here --> 

.panel[.panel-name[i, j: logical, integer]

.left-code[
<br>
```{r l-i, eval = F}  
tab_data %>% 
  flextable(
    col_keys = c("country_name", "region", "y_1995", "y_2005")
  ) %>% 
  color(
    i = rep(c(TRUE, FALSE), each = 5), #<<
    j = c(1, 3), #<<
    color = "red"
  )
```
]
 
.right-plot[
<br>
```{r l-i-f, ref.label = "l-i", echo = F}  

```
]
  
   ] 
 
   <!-- panel ends here --> 

.panel[.panel-name[by parts]

.left-code[
<br>
You can refer to parts of the table using `part =` option. The available options are

+ `header`: the header part of the table
+ `footer`: the footer part of the table
+ `body`: the body part of the table
+ `all`: the body and the header parts of the table

<br>
```{r parts, eval = F}  
tab_data %>% 
  flextable(
    col_keys = c("country_name", "region", "y_1995", "y_2005")
  ) %>% 
  color(
    part = "header", #<<
    color = "red"
  )
```

Different functions have different default values for `part`.

]
 
.right-plot[
<br>
```{r parts-f, ref.label = "parts", echo = F}  

```
]


  ] 

  <!-- panel ends here -->
]
 
 <!-- panel set ends here -->

---

# Style 

.panelset[ 

.panel[.panel-name[Introduction]

.left5[

.content-box-green[**Explanation**]

You can use the `style()` function to change the style and format of a table.  

.content-box-green[**Syntax**]

```{r eval = F}
style(
  x,
  i = NULL,
  j = NULL,
  pr_t = NULL,
  pr_p = NULL,
  pr_c = NULL,
  part = "body"
)
```

We use `fp_*()` functions from the `officer` package to specify the style of .blue[texts], .blue[paragraphs], and .blue[cells].

+ `pr_t = fp_text()`: format texts 
+ `pr_p = fp_par()`: format paragraphs
+ `pr_c = fp_celll()`: format cells

(`pr` in `pr_* =`  stands for property.)

]

.right5[

.content-box-green[**paragraph vs cell**]

+ paragraphs: rectangular boxes around the texts inside the cells 
+ cells: rectangular boxes that contain paragraphs and texts inside 

```{r }
head(tab_data, 5) %>% #BREAK 
  flextable(
    col_keys = c("country_name", "region", "y_1995", "y_2005")
  ) %>% 
  style(
    pr_p = fp_par(
      border.bottom = fp_border(width = 2, color = "blue")
    ),
    pr_c = fp_cell(
      border.bottom = fp_border(width = 2, color = "red")
    ) 
  )
```

It seems that cell formatting is sufficient and paragraph formatting is not necessary...

]


  ]

<!-- panel ends here -->

.panel[.panel-name[fp_text]

.left5[

.content-box-green[**Explanation**]

`fp_text()` lets you update the appearance of texts, including color, font size, bold or not, etc (see the help page below for the complete list of options).

.content-box-green[**Syntax**]

```{r eval = F}
?fp_text  
```

.scroll-box-24[
```{r echo = F}
?fp_text  
```
]

]

.right5[

.content-box-green[**Example**]

```{r }
tab_data %>% 
  flextable(
    col_keys = c("country_name", "region", "y_1995", "y_2005")
  ) %>% 
  style(
    i = 4:6,
    j = ~ country_name,
    pr_t = fp_text( #<<
      color = "red", #<<
      bold = TRUE, #<<
      underlined = TRUE #<<
    ) #<<
  )
```

]

  ]

<!-- panel ends here -->

.panel[.panel-name[fp_par]

.left5[

.content-box-green[**Explanation**]

`fp_paragraph()` lets you update the appearance of paragraphs (see the help page below for the complete list of options).

.content-box-green[**Syntax**]

.scroll-box-24[
```{r }
?fp_par  
```
]

  ]


.right5[

.content-box-green[**Example**]

```{r }
tab_data %>% 
  flextable(
    col_keys = c("country_name", "region", "y_1995", "y_2005")
  ) %>% 
  style(
    i = 4:6,
    j = ~ country_name,
    pr_p = fp_par( #<<
      text.align = "left", #<<
      shading.color = "grey" #<<
    ) #<<
  )
```

]

  ]

<!-- panel ends here -->

.panel[.panel-name[fp_cell]

.left5[

.content-box-green[**Explanation**]

`fp_cell()` lets you update the appearance of cells (see the help page below for the complete list of options).

.content-box-green[**Syntax**]

.scroll-box-24[
```{r }
?fp_cell  
```
]

  ]


.right5[

.content-box-green[**Example**]

```{r }
tab_data %>% 
  flextable(
    col_keys = c("country_name", "region", "y_1995", "y_2005")
  ) %>% 
  style(
    i = 4:6,
    j = ~ country_name,
    pr_c = fp_cell( #<<
      margin = 2, #<<
      vertical.align = "top", #<<
      background.color = "red" #<<
    ) #<<
  )
```

]

  ]

<!-- panel ends here -->


]

<!-- panel set ends here -->

---

# Formatting: convenience functions 

.left5[

.content-box-green[**Instruction**]

+ Instead of using the `style()` function in combination with `fp_*()` functions, you can easily work on a specific aesthetic feature using convenience functions like below:
  * `align()`, `align_text_col()`, `align_nottext_col()`: Set text alignment
  * `bg()`: Set background color
  * `font()`: Set font
  * `fontsize()`: Set font size
  * `italic()`: Set italic font
  * `bold()`: Set bold font
  * `color()`: Set font color
  * `padding()`: Set paragraph paddings
  * `valign()`: Set vertical alignment
  * `rotate()`: rotate cell text
  * `empty_blanks()`: make blank columns as transparent

+ You can use the selector syntax to specify where the specified effects take place

]

.right5[

.content-box-green[**Example**]

```{r }
tab_data %>%  
  flextable(
    col_keys = c("country_name", "region", "y_1995", "y_2005")
  ) %>%  
  color(i = 1, j = 2, "#fcba03") %>%  
  bg(i = ~ y_2005 < 2e5, j = c("region"), bg = "grey") %>%  
  fontsize(i = 7, size = 16)   
```
]


---

```{r format-flip, cache = F, include = F}
tab_data %>% #BREAK 
  flextable(
    col_keys = c("country_name", "region", "y_1995", "y_2005")
  ) %>% #BREAK 
  #--- text color ---#
  color(i = 1, j = 2, "#fcba03") %>% #BREAK 
  #--- background ---#
  bg(i = ~ y_2005 < 2e5, j = c("region"), bg = "grey") %>% #BREAK 
  #--- font type ---#
  font(i = 5, j = ~ country_name + y_2005, fontname = "Times") %>% #BREAK 
  #--- font size ---#
  fontsize(i = 7, size = 16) %>% #BREAK 
  #--- italicize ---#
  italic(j = 2) %>% #BREAK 
  #--- bold ---#
  bold(j = 4) %>% #BREAK 
  #--- vertical text alignment ---#
  valign(i = ~ region == "Australasia", j = 4, valign = "top") %>% #BREAK 
  #--- text direction ---#
  rotate(i = 1, j = 2, rotation = "tbrl") 

```

`r chunk_reveal(
  chunk_name = "format-flip", 
  title = "# Demonstration: Formatting",
  break_type = "user", 
  display_type = "both", 
  left_assign = FALSE,
  width = c(50, 50)
  )
`

---

# Borders: convenience functions  

.panelset[ 

.panel[.panel-name[Instruction]

.left5[

.content-box-green[**List of functions**]

Here is a list of convenience functions that you can use to draw border lines on a table:
+ `hline()`: set horizontal borders
+ `hline_bottom()`: set bottom horizontal border
+ `hline_top()`: set top horizontal border
+ `vline()`: set vertical borders
+ `vline_left()`: set flextable left vertical borders
+ `vline_right()`: set flextable right vertical borders
+ `border()`: Set cell borders
+ `border_inner()`: set vertical & horizontal inner borders
+ `border_inner_h()`: set inner borders
+ `border_inner_v()`: set vertical inner borders
+ `border_outer()`: set outer borders
+ `border_remove()`: remove borders
+ `fix_border_issues()`: fix border issues when cell are merged

]

.right5[

.content-box-green[**Instruction**]

+ Use the selector syntax to specify where
+ Use the `border = ` option along with `fp_border()` from the `officer` package to specify what kind of borders you would like to draw

.content-box-green[**Example**]

```{r eval = F}
hline(
    i = 3, 
    j = 1:3, 
    border = fp_border( 
      color = "red", 
      style = "dotted", 
      width = 4 
    ) 
  ) 
```

]


] 

<!-- panel ends here -->

.panel[.panel-name[fp_border()]

.left5[

.content-box-green[**Explanation**]

`fp_border()` lets you specify the aesthetics of the borders you are drawing.

.content-box-green[**Syntax**]

```{r eval = F}
?fp_border  
```

.scroll-box-24[
```{r echo = F}
?fp_border  
```
]

]

.right5[

.content-box-green[**Example**]

```{r }
tab_data %>%  
  flextable(
    col_keys = c("country_name", "region", "y_1995", "y_2005")
  ) %>%  
  hline(
    i = 3, 
    j = 1:3, 
    border = fp_border( #<<
      color = "red", #<<
      style = "dotted", #<<
      width = 4 #<<
    ) #<<
  ) 
```

]

  ] 

  <!-- panel ends here -->

]

<!-- panel set ends here -->


---

```{r borders-f, cache = F, include = F}
tab_data %>% #BREAK 
  flextable(
    col_keys = c("country_name", "region", "y_1995", "y_2005")
  ) %>% #BREAK 
  #--- remove all borders ---#
  border_remove() %>% #BREAK
  #--- horizontal lines ---#
  hline(i = 3, j = 1:3, border = fp_border(color = "red")) %>% #BREAK 
  #--- horizontal line at the bottom ---#
  hline_bottom(j = 3:4, border = fp_border(color = "green")) %>% #BREAK 
  #--- horizontal line at the top ---#
  hline_top(j = 1:3, border = fp_border(color = "orange")) %>% #BREAK 
  #--- vertical lines ---#
  vline(border = fp_border(color = "orange")) %>% #BREAK 
  #--- vertical on the left edge  ---#
  vline_left(border = fp_border(color = "grey", width = 2)) %>% #BREAK 
  #--- vertical on the right edge  ---#
  vline_right(border = fp_border(color = "red", width = 2)) %>% #BREAK 
  #--- borders of individual cells ---#
  border(i = 4, j = 2, 
    border.top = fp_border(color = "red", width = 3),
    border.left = fp_border(color = "green", width = 3),
    border.right = fp_border(color = "black", width = 3),
    border.bottom = fp_border(color = "pink", width = 3)
  ) %>% #BREAK 
  #--- horizontal lines (inner) ---#
  border_inner_h(border = fp_border(color = "black")) %>% #BREAK 
  #--- vertical lines (inner) ---#
  border_inner_v(border = fp_border(color = "black")) %>% #BREAK 
  border_remove() %>% #BREAK 
  #--- the outer lines ---#
  border_outer(border = fp_border(color = "red", width = 4)) 

```

`r chunk_reveal(
  chunk_name = "borders-f", 
  title = "# Demonstration: Borders",
  break_type = "user", 
  display_type = "both", 
  left_assign = FALSE,
  width = c(50, 50)
  )
`

---

# Layout

.content-box-green[**List of functions**]

Here is a list of functions you can use to change the layout of a table:

+ `merge_at()`: Merge flextable cells into a single one
+ `merge_h()`: Merge flextable cells horizontally
+ `merge_h_range()`: rowwise merge of a range of columns
+ `merge_v()`: Merge flextable cells vertically
+ `height()`, `height_all()`: Set flextable rows height
+ `width()`: Set flextable columns width
+ `hrule()`: Set flextable rule for rows heights
+ `autofit()`: Adjusts cell widths and heights
+ `fit_to_width()`: fit a flextable to a maximum width
+ `as_grouped_data()`: grouped data transformation

]

---

```{r layout, cache = F, include = F}
tab_data %>% #BREAK 
  flextable(
    col_keys = c("country_name", "region", "y_1995", "y_2005")
  ) %>% #BREAK 
  merge_v(j = ~ region) %>% #BREAK 
  height(i = 5, height = 4) %>% #BREAK 
  width(j = 2, width = 2) %>% #BREAK 
  autofit()
```

`r chunk_reveal(
  chunk_name = "layout", 
  title = "# Layout",
  break_type = "user", 
  display_type = "both", 
  left_assign = FALSE,
  width = c(50, 50)
  )
`

---

# Content transformation and formatting

.panelset[ 

<!-- panel ends here -->

.panel[.panel-name[Instruction]

.left5[
<br>

Here is a list of functions that let you alter the content of a table:

+ `colformat_char()`: format character cells
+ `colformat_int()`: format integer cells
+ `colformat_num()`: format numeric cells
+ `compose()`, `mk_par()`: Define flextable displayed values
+ `as_paragraph()`: concatenate chunks in a flextable
+ `as_chunk()`: chunk of text wrapper
+ `as_image()`: image chunk wrapper
+ `minibar()`: mini barplots chunk wrapper
+ `linerange()`: mini linerange chunk wrapper
+ `lollipop()`: mini lollipop chart chunk wrapper
+ `as_b()`: bold chunk
+ `as_bracket()`: chunk with values in brackets
+ `as_i()`: italic chunk
+ `as_sub()`: subscript chunk
+ `as_sup()`: superscript chunk
+ `hyperlink_text()`: chunk of text with hyperlink wrapper
+ `set_caption()`: set caption
+ `footnote()`: add footnotes to flextable
+ `void()`: Delete flextable content
]

.right5[

<br>

.content-box-green[**Note**]

Most of the functions are replaceable with either other `flextable` functions or R operations before applying `flextable()` to initiate a table from the `data.frame`.

.content-box-green[**Example**]

`colformat_char()` lets you add prefix and suffix to character columns. But, you can easily achieve this using `stringr::str_c` to a `data.frame` prior to apply `flextable()` to the `data.frame`.

.content-box-green[**Here**]

We will take a look at those that are likely to be useful and cannot be easily achieved pre-flextable.

+ `as_image()`
+ `hyperlink_text()`
+ `set_caption()`
+ `footnote()`

]

]
<!-- panel ends here -->

.panel[.panel-name[How]

.left5[

<br>

.content-box-green[**Syntax**]

```{r eval = F}
compose(i, j, value = as_paragraph(a list of elements))
```

`as_paragraph()` concatenates list of the elements specified inside it. 

.content-box-green[**Example**]

The following code first concatenate the values of y_2015, the text "uses ", and "image.png" row-wise and then place it to the [row = 1, col = 2] element of the table, overriding the existing content in the element.

```{r eval = F}
compose(i = 1, j = 2, 
  value = as_paragraph(
    #--- y_2015 ---#
    as_chunk(y_2015), #<<
    #--- text ---#
    "uses ", #<<
    #--- image ---#
    as_image("image.png") #<<
  )
)
```

]

.right5[

<br>

.content-box-green[**Example with outcome**]

```{r }
img.file <- file.path( R.home("doc"), "html", "logo.jpg" )

tab_data %>% #BREAK 
  flextable(
    col_keys = c("country_name", "region", "y_1995", "y_2005")
  ) %>% #BREAK 
  compose(i = c(1, 3), j = 1,
    value = as_paragraph(
      as_chunk(country_name), #<<
      "uses ", #<<
      as_image(src = img.file) #<<
    ) 
  ) %>% 
  autofit()  


```

]


]

<!-- panel ends here -->

.panel[.panel-name[footnote()]

  ] 

  <!-- panel ends here -->

.panel[.panel-name[set_caption()]

  ] 

  <!-- panel ends here -->

]

<!-- panel set ends here -->


---

```{r format-content-flip, cache = F, include = F}
tab_data %>% #BREAK 
  flextable(
    col_keys = c("country_name", "region", "y_1995", "y_2005")
  ) %>% #BREAK 
  compose(j = 1,
    value = as_paragraph(
      as_chunk(region),
      "-",
      as_chunk(country_name)
    )
  )
  

```

`r chunk_reveal(
  chunk_name = "format-content-flip", 
  title = "# Format",
  break_type = "user", 
  display_type = "both", 
  left_assign = FALSE,
  width = c(50, 50)
  )
`

---

# Footnote 

---

# Output 




