---
title: "R as GIS: Vector Data"
author: "AECN 396/896-002"
output:
  xaringan::moon_reader:
    # css: [default, metropolis, metropolis-fonts] 
    css: xaringan-themer.css 
    lib_dir: libs
    nature:
      ratio: 16:10
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r, child = './../setup.Rmd'}
```

```{r, include = F, cache = F}
suppressMessages(library(sf)) 
suppressMessages(library(ggplot2)) 
suppressMessages(library(patchwork))
```

# Before you start

## Learning objectives

The objectives of this chapter is to learn how to use R as GIS 

## Table of contents

1. [Understanding the data structure of `sf`](#data-structure)
2. [Create sf objects yourself](#create-sf)
3. [Importing and exporting spatial datasets](#input-output)
4. [Miiscellaneous operations](#misc)
5. [Non-interactive transformation](#non-int-trans)

<br>

<span style="color:red"> Tips: </span>hitting letter "o" key will give you a panel view of the slides


---

# 

