---
title: "Introduction to R"
author: "AECN 396/896-002"
output:
  xaringan::moon_reader:
    # css: [default, metropolis, metropolis-fonts] 
    css: xaringan-themer.css 
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

# Table of contents

1. [Introduction to R and RStudio](#intro)
2. [Object types](#objects)
3. [Functions and packages](#functions)
4. [Some fundamentals on vector, matrix, list, and data.frame](#objects_basics)

---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_duo_accent(
  header_h1_font_size = "2.1rem",
  header_h2_font_size = "1.5rem",
  text_font_size = "0.9rem",
  code_font_size = "0.8rem"
)

```

```{r , include=FALSE}
library(knitr)
opts_chunk$set(
  fig.align = "center", #fig.width=6, fig.height=4.5, 
  # out.width="748px", #out.length="520.75px",
  # dpi = 100, #fig.path='Figs/',
  cache = T#, echo=F, warning=F, message=F
  )

opts_knit$set(root.dir = "~/Box/Teaching/UNL/DataScience")
# setwd("~/Box/Teaching/UNL/DataScience") 
```

```{r, include = F}
#--- load packages ---#
suppressMessages(library(data.table))
suppressMessages(library(exactextractr))
suppressMessages(library(prism))
suppressMessages(library(sf))
suppressMessages(library(raster))
suppressMessages(library(tidyverse))
suppressMessages(library(DT))
suppressMessages(library(tictoc))
suppressMessages(library(tmap))
suppressMessages(library(parallel))
suppressMessages(library(maps))
````


class: inverse, center, middle
name: intro

# Introduction to R and RStudio

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---

# Learning Objectives

+ become familiar with programming 
+ become capable of using R software to conduct research independently
  + manipulate data
  + visualize data 
  + report results 
  + spatial data management

---

# R and RStudio

## R 
+ a very popular statistical programming language used in academia and industry  
+ started out as software to do statistics, designed by a number of statisticians
+ is open-source and free
+ has been and is evolving rapidly by the contributions of its users 
 + state-of-the-art statistical methods (e.g., machine learning algorithms) written by the developers of the methods
 + geographic information system (GIS)
 + big data handling and analysis

## RStudio:

+ R has a terrible graphic user interface 
+ by far the most popular graphic user interface of R

---


# Install R and RStudio 

+ Install [R](https://www.r-project.org/)
+ Install [RStudio](https://www.rstudio.com/)

---

# Introduction to RStudio

## Four panes

+ R script (upper left)
+ Console (lower left)
+ Environment (upper right)
+ Files, plots, packages, and help (lower right)

## Small tips

+ Appearance
+ Pane Layout 

---

# Getting started with R and RStudio 

+ do basic mathematical operations 
+ define objects in R 
+ learn different object types
+ use RStudio at the same time 

---

<!-- 
#/*=================================================*/
#' # new slide stars here 
#/*=================================================*/
-->

#  Basic element types (atomic mode)

+ integer: e.g., 1, 3,  
+ numeric (double): e.g., 1, 1.3
+ complex: 
+ logical (boolean): true or false 
+ character: combination of letters (numerical operations not allowed)

---

<!-- 
#/*=================================================*/
#' # new slide stars here 
#/*=================================================*/
-->

# Basic arithmetic: R as a calculator

```{r }
#--- addition ---#
2 + 3.3   

#--- subtraction ---#
6 - 2.7   

#--- multiplications ---#
6 * 2   

#--- exponentiation ---#
2 ^ 2.4  

```

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

<span style="color:red"> RStudio tip: </span>
+ `command` + `enter` (Mac) runs the code (`Control` + `enter` (Windows))


---

<!-- 
#/*=================================================*/
#' # new slide stars here 
#/*=================================================*/
-->

# Basic arithmetic: R as a calculator (cont.)

```{r }
#--- division ---#
6 / 2   

#--- remainder ---#
6 %% 4   

#--- quotient ---#
6 %/% 4   
   

```

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

<span style="color:red"> RStudio tip: </span>: 
+ `command` (`Control`) + 1 for Mac (Windows) to move the cursor to the source pane
+ `command` (`Control`) + 2 for Mac (Windows) to move the cursor to the source pane

---

<!-- 
#/*=================================================*/
#' # new slide stars here 
#/*=================================================*/
-->

# logical values and operators

```{r }
#--- true or false ---#
5 == 5
5 == 4
5 > 4
5 < 4  
```

---


<!-- 
#/*=================================================*/
#' # new slide stars here 
#/*=================================================*/
-->

# Character

Contents enclosed by double (or single) quotation marks will be recognized as characters.

```{r }
#--- character ---#
"R"

#--- character ---#
" rocks"

```

You cannot do addition using characters

```{r , error = TRUE}
"R" + "rocks"
```

We will learn string manipulations later using the `stringr` package.

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

<span style="color:red"> RStudio tip: </span>: 
+ `Option` + `-` (`Alt` + `-`) for Mac (Windows) to inset the assignment operator (`<-`)

---


<!-- 
#/*=================================================*/
#' # new slide stars here 
#/*=================================================*/
-->


# Assignment 

You can assign contents (numeric numbers, character, boolean, etc) to an object on R and reuse it later using either `<-` or `=`.

```{r eval = F}
object_name <- contents   
```

## Example

```{r }
#--- numeric ---#
a <- 1   
``` 

Notice that these objects are now in the list of objects on the environment tab of RStudio.

## Object evaluation 

Once objects are created, you can evaluate them on the console to see what is inside:

```{r }
a  
```

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

<span style="color:red"> RStudio tip: </span>: 
+ `Command` + `up` (`Control` + `up`) for Mac (Windows) to look at the history of codes you have run (or go to the "History tab" on the right upper pane)

---

# Assignment (cont.)


## Other examples

```{r }
#--- character ---#
b <- "R rocks"

b

#--- logical ---#
d <- 1 == 2

d
```

---

# Assignment (cont.)

You can also use `=` for assignment 

```{r }
#--- character ---#
a = "R rocks"

a
```

It does not really matter which of `<-` or  `=` to use. You should pick whichever makes sense for you. But, it is a good idea to be consistent.

---

# Assignment (cont.)

Several things to remember about assignment: 

+ If you assign contents to an object of the same name, the object that had the same name will be overwritten 

```{r }
a <- 3
a <- 1  
a
```

+ Object names cannot start with a numeric number. Try the following:

```{r numeric_start, eval = FALSE}
1a <- 2  
```

+ You cannot use a reserved word as the name of an object (complete list found [here](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Reserved.html))

```{r eval = FALSE}
#--- try this ---#
if <- 3  
```

---

class: inverse, center, middle
name: objects

# Objects

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---

# Objects

R is an object-oriented programming (OOP), which basically means: 

"Everything is an object and everything has a name."

## Different object types (classes)

+ vector
+ matrix
+ data.frame
+ list
+ function

---

# Vector (one-dimensional array)

A vectors is a class of object that consists of elements of the same kind (it can have only one element). You use `c()` to create a vector.

## Example

```{r }
#--- create a vector of numbers ---#
a <- c(4, 3, 5, 9, 1)   

a

#--- create a vector of characters ---#
b <- c("python", "is", "better", "than", "R")   

b
```

---

# Vector (cont.)

What if we mix elements of different mode  

## Example

```{r }
#--- create a vector of numbers ---#
a_vector <- c(4, 3, "5", 9, 1)   

#--- see its content ---#
a_vector

```

All the numeric values are converted to characters.


---

# List

A `list` is a class of object that consists of elements of mixed types:

## Example

```{r }
#--- create a vector of numbers ---#
a_list <- list(4, 3, "5", 9, 1)   

#--- see its content ---#
a_list

```

A `list` is very flexible. It can hold basically any type of R objects as its elements. 

---

# Matrix (two-dimensional array)

A matrix is a class of object that consists of elements of the same kind (it can have only one element) stored in a two-dimensional array. 

## Example

```{r }
#--- create a matrix of numbers ---#
M_num <- matrix(c(2, 4, 3, 1, 5, 7), nrow = 3)

M_num
```

---

# Matrix (two-dimensional array)

A matrix is a class of object that consists of elements of the same kind (it can have only one element) stored in a two-dimensional array. 

## Example

```{r }
#--- create a matrix of characters ---#
M_char <- matrix(c("a", "b" , "c", "d", "e", "f"), nrow = 3)

M_char
```

---

# data.frame

`data.frame` is like a matrix (or a list of columns)

```{r }
#--- create a data.frame ---#
yield_data <- data.frame(
  nitrogen = c(200, 180, 300),
  yield = c(240, 220, 230),
  state = c("Kansas", "Nebraska", "Iowas")
)

yield_data
```

There are different kinds of objects that are like "data.frame"

+ tibble 
+ data.table

---

# Objects (cont.)

It is critical to recognize the class of the objects:

+ the same function does different things depending on the class of the object to which the function is applied  
+ some functions work on some object classes, but not on others 

Many of the errors you will encounter while working on R has something to do with applying functions that are not applicable to the objects you are working on!

<!-- How is `class` different from atomic mode?

Roughly put, atomic mode refers to the type of the most basic elements that will make up more complex objects like a data.frame or a matrix. 

 -->

---

# Objects (cont.) 

Use `class`, `typeof`, and `str` commands to know more about what kind of objects you are dealing with:

```{r }
#--- check the class ---#
class(yield_data)  

#--- check the "internal" type ---#
typeof(yield_data)  

#--- look into the structure of an object ---#
str(yield_data)  
```

---

# Objects (cont.) 

You could also use `View()` function for visual inspection:

```{r eval = F}
View(yield_data)  
```

<!-- 
#/*=================================================*/
#' # Functions
#/*=================================================*/
-->

---

class: inverse, center, middle
name: function

# Function

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---

# Function 

## What is a function?
A function takes R objects (vector, data.frame, etc), processes them, and returns R objects 

## Example

`min()` takes a vector of values as an argument and returns the minimum of all the values in the vector

```{r }
min(c(1, 2))  
```

---

# Some other useful functions


+ create a sequence of values 

```{r }
v1 <- seq(0, 100, by = 5)
v2 <- seq(0, 100, length = 21)
```

+ repeat values 

```{r }
v3 <- rep(10,5)
```

+ sum values 

```{r }
v1_sum <- sum(v1)
  
```

+ find the length of an vector

```{r }
v1_len <- length(v1)
```

---

# Exercises

+ generate a vector (call it $x$) that starts from 1 and increase by 2 until 99 

+ calculate the sample mean of $x$ 

$\frac{1}{n}\sum_{i=1}^n x_i$

+ calculate the sample variance  of $x$

$\frac{1}{n}\sum_{i=1}^n (x_i-\bar{x})^2$, where $\bar{x}$ is the sample mean

---

# Function (cont.)

+ Functions (both base and user-written) are what makes R compelling to use as major statistical and programming software!
     
+ Indeed, this course is pretty much all about learning useful functions that make your life easier

+ We will learn lots of functions that are made available through user-written packages
 

---

# Packages

## What are packages?

A drawer in your work space (R environment) that has specialized tools (functions) to complete tasks 

## Example packages
+ dplyr (data wrangling)
+ data.table (data wrangling)
+ ggplot2 (data visualization)
+ sf (spatial vector data handling)
+ raster (spatial raster data handling)
+ stars (spatiotemporal data handling)

---

# Packages

+ Before you use tools (functions) in the drawer (package), you need to buy (install) it first. You can install a package using the following syntax:

```{r echo=TRUE, eval=FALSE}
install.packages('package name') 
```

+ For example,

```{r eval = F}
install.packages(ggplot2) 
```

+ You need to bring the drawer (package) to your working space (R environment) by using the `library()` function:

```{r eval = F}
library(ggplot2) 
```

+ Now, you can start using specialized tools (functions) in the drawer (package)!!

---

# Working with R (or any computer programs)

+ You are the architect who has the blueprint of the final product, but does not have an ability to build specific pieces by yourself 

+ You work with one worker (R) who can build specific pieces perfectly without any error <span style="color:red"> if given right tools and instructions </span>

+ This worker is weird. If you do not give right tools or your instruction is wrong, he/she will speak up and tell you there has been an error. He/she will not try to figure out how things could have been done differently by himself/herself.

+ Your job is to provide the right tools and instructions to the worker (R), and correct your instructions when you found out you made a mistake (debugging)

<!-- 
#/*=================================================*/
#' # Vector and matrix
#/*=================================================*/
-->

---

class: inverse, center, middle
name: objects_basics

# A bit more on vector, matrix, list, and data.frame

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---


# Vector arithmetic

Let's define two vectors to work with
```{r }
a <- c(1,2,3,4,5)

b <- a
```

---

# Vector arithmetic

```{r }
#--- addition ---#  
a + b

#--- subtraction ---#  
a - b

#--- multiplication ---#  
a * b

#--- division ---#  
a / b
```

---

# Accessing elements of a vector

To access element(s) of a vector, you use `[]` like below:

```{r }
#--- second element of vector a ---#
a[2]

#--- third element of vector a ---#
b[3]  
```

---

# Accessing elements of a vector

You can access multiple elements of a vector 

```{r }
#--- first and second elements of vector a ---#
a[c(1,2)]

#--- first and fifth elements of vector b ---#
b[c(1,5)] 
```

---

# Matrix

```{r }
#--- define matrix ---#
# A <- matrix(a value, # of rows, # of cols)
A <- matrix(2,3,2)
A
# B <- matrix(vector,# rows)
B <- matrix(c(2,4,3,1,5,7),nrow=3)
B  
```


---

# Matrix arithmetic

```{r }
#--- matrix addition by element ---#
A+B 

#--- matrix multiplication by element ---#
A*B   
   
```

---

# Matrix operations

```{r }
#--- transpose ---#
t(B)

#--- matrix multiplication ---#
A %*% t(B)   
   
```

---

# Accessing elements of a matrix

To access element(s) of a matrix, you use `[]` just like we did for a vector. But, now you have two arguments inside `[]`. 

```{r eval = F}
matrix[indices for rows, indices for columns]  
```

## Examples

```{r }
#--- 1st row, 2nd column ---#
A[1, 2]

#--- 1st and 3rd row, 2nd column ---#
A[c(1, 3), 2]

#--- all the values in the first row ---#
A[1, ]
```

---

# List

As I mentioned earlier, a `list` can hold mixed objects of any type in a single list.

```{r }
a_list <- list(
  a = matrix(2,3,2), 
  b = c("R", "rocks"), 
  c = yield_data
)  

a_list
```

Note that each element has a name.

---

# Accessing elements of a list

To access element(s) of a matrix, you can use `[[]]` or `$` operator for accessing a single element and use `[]` for multiple elements.

## Examples

```{r }
#--- 2nd element ---#
a_list[[2]]

```

---

# Accessing elements of a list

To access element(s) of a matrix, you can use `[[]]` or `$` operator for accessing a single element and use `[]` for multiple elements.

## Examples

```{r }
#--- 1st and 3rd elements ---#
a_list[c(1, 3)]
```

---

# Accessing elements of a list

To access element(s) of a matrix, you can use `[[]]` or `$` operator for accessing a single element and use `[]` for multiple elements.

## Examples

```{r }
#--- an element called b (2nd element) ---#
a_list$b
```

---

# data.frame

`data.frame` (and its relatives) 
+ is the most common object type we use. 
+ is a special kind of `list` of vectors that are of the same length that makes a `matrix`-like structure
+ shares properties of both the matrix and the list


```{r }
yield_data
```

---

# Accessing parts of a `data.frame`

Accessing parts of a `data.frame` works like accessing elements of a matrix or list.

## Examples

```{r }
#--- list-like: access the state column (element) ---#
yield_data$state

#--- matrix-like: 1st row, 2nd column (yield) ---#
yield_data[1, 2]

```

--

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

<span style="color:red"> We will spend lots of time on how to do data wrangling on `data.frame`s using the `tidyverse` package! </span> 

---
class: inverse, center, middle
name: rgis

# Next class: Rmarkdown

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>










