---
title: "Final Paper Expectations and Regulations"
author: "AECN 396/896-002"
output:
  xaringan::moon_reader:
    # css: [default, metropolis, metropolis-fonts] 
    css: xaringan-themer.css 
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: middle

```{r, child = './../setup.Rmd'}
```

```{r, include = F}
#--- load packages ---#
suppressMessages(library(tidyverse))
suppressMessages(library(data.table))
suppressMessages(library(parallel))
````

# Final Project

.content-box-green[**Submission**]

+ both Rmd and html
+ use the submission link I will provide


.content-box-green[**Deadline**]

+ Final project is due Nov 25th.
+ Final project approval is due Oct 16th.

---

# Final Project Structure

.content-box-green[**Introduction**]: 2 points

+ Describe what you are trying to find out (research question)
+ Describe why the research question is worthwhile answering

(one paragraph)

<br>
--

.content-box-green[**Data Description and Processing**]: 25 points

+ Describe what the final dataset would look like
  * observational units (plots in an agronomic experiment, county, gene expression, etc)
  * temporal and spatial scales if applicable 
  * list of variables 
+ Provide summary statistics for each of the raw datasets using tables
+ Generate appropriate figures that help the readers understand the nature of the datasets 
+ Process raw datasets and combine them 

<br>
--

.content-box-green[**Exploratory Analysis**]: 12 points

+ explore visually the relationship of variables you are interested in understanding 
+ run statistical analysis (unless impossible) and explain the results (Note: if you are Ag Econ students, you must run at least one regression)
+ present the results in a table and draw conclusions

<br>
--

.content-box-green[**Conclusion**]: 1 point

+ summarize the report in several sentences

---
class: middle

# Regulations and Bonus Points

.content-box-green[**Regulation**]

+ Use Rmarkdown 
+ Use at least two raw datasets that need to be combined before analysis, and you need to merge the datasets at some point 
+ Annotate each of line of your R codes (for practice) 
+ Use `tidyverse` functions wherever possible (for practice)
+ Use `ggplot2` for all the figures you generate 
+ Use one of `gt`, `flextable`, and `modelsummary` to create tables (Note: if you are in Ag Econ, you need to create at least one regression results table using `modelsummary`)
+ Your project must include spatial operations using at least one of `sf`, `raster`, and `stars` packages if you are in Agronomy and Horticulture.

--

<br>

.content-box-green[**Bonus Points**]

+ Define and use a function(s) in a meaningful way (2 points)
+ Use looping in a meaningful way and parallelize it (2 points)
+ Use more than three types figures in a meaningful way (2 points)

---
class: middle

# Hints

.content-box-green[**Hint 1**]

The fastest and easiest way to find your final project is to simply bring back one of your old projects (whether it was done in R or outside of R) and redo everything (data wrangling, visualization, statistical analysis, reporting) using R, taking advantage of what you have learned in the class.
  
<br>
--

.content-box-green[**Hint 2**]

Assignment 1 is very similar to what your final project should look like

<br>  
--

.content-box-green[**Hint 3**]

This assignment is like writing a journal article except you provide lots of details about how you process datasets with all the R codes displayed.



 