---
title: "Data wrangling with `tidyverse`: Reshaping and Merging"
author: "AECN 396/896-002"
output:
  xaringan::moon_reader:
    # css: [default, metropolis, metropolis-fonts] 
    css: xaringan-themer.css 
    self_contained: true
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: middle

```{r, child = './../setup.Rmd'}
```

# Table of contents

1. [Reshaping data](#reshaping)
2. [Merging datasets](#merging)

---
class: middle

# Learning objectives

The objectives of this chapter is to learn how to use the `tidyverse` package to 

+ reshape a dataset
+ merge multiple datasets 

---
class: inverse, center, middle
name: reshaping

# Reshaping

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---

# Long and wide formats

.left5[

## Long format

A single column representing a single variable 

```{r long-data-gen-2, echo = F}
(
yield_data_long <- data.frame(
  state = c("Kansas", "Nebraska", "Iowa", "Illinois") %>% rep(each = 2),
  year = c(2019, 2020) %>% rep(4),
  yield = c(200, 240, 210, 220, 220, 230, 190, 150)
  )
)
```
]

.right5[

## Wide format

Multiple column representing a single variable 

```{r echo = F}
(
  yield_data_wide <- pivot_wider(yield_data_long, names_from = year, values_from = yield)
)
```

Note: there is nothing in the dataset that tells you what the data values represent in the wide format.

]

---

# Long to wide 

.panelset[ 

<!-- panel ends here -->

.panel[.panel-name[prepare]

Create the following dataset in long format:

```{r long-data-gen-1}
yield_data_long <- data.frame(
  state = c("Kansas", "Nebraska", "Iowa", "Illinois") %>% rep(each = 2),
  year = c(2019, 2020) %>% rep(4),
  yield = c(200, 240, 210, 220, 220, 230, 190, 150)
  )
```

  ]

<!-- panel ends here -->
.panel[.panel-name[how and example]

.left5[

## How

To convert a long-formatted `data.frame` into a wide-formatted `data.frame`, you can use `pivot_wider()` function from the `tidyr` package.

```{r eval = F}
#--- NOT RUN ---#
pivot_wider(
  data, 
  names_from = variable 1, 
  values_from = variable 2
)
```

+ the value of `variable 1` becomes the name the new variables
+ the value of `variable 2` becomes the value of the new variables
]

.right5[

## Example

```{r }
yield_data_wide <- pivot_wider(
  yield_data_long, 
  names_from = year, 
  values_from = yield
)

yield_data_wide
```

+ the value of `year` becomes the name the new variables
+ the value of `yield` becomes the value of the new variables

]

  ]

  <!-- panel ends here -->

.panel[.panel-name[prefix]

You can append a character string to the new variable names. The previous example had `2019` and `2010` as the name of the new variables.

```{r }
yield_data_long %>%
  pivot_wider(
    #--- let R know what the prefix is ---#
    names_prefix = "yield_",
    names_from = "year",
    values_from = "yield"
  )
```

  ] 

  <!-- panel ends here -->

.panel[.panel-name[multiple columns]

.left5[

## Prepare a dataset 

Create the following data in long format;

```{r wide-data-gen}
yield_data_long <- data.frame(
  state = c("Kansas", "Nebraska", "Iowa", "Illinois") %>% rep(each = 2),
  year = c(2019, 2020) %>% rep(4),
  yield = c(200, 240, 210, 220, 220, 230, 190, 150),
  rainfall = c(14, 15, 15, 16, 20, 21, 24, 15)
  ) 
```

```{r echo = F}
yield_data_long
```
]

.right5[

## Example

You can simply supply multiple variables to be made wide like this:

```{r }
yield_data_long %>%  
  pivot_wider(
    names_from = "year",
    values_from = c("yield", "rainfall")
  )  
```
]
  ] 

  <!-- panel ends here -->
]

<!-- panel set ends here -->

---

# Wide to long

.panelset[ 

.panel[.panel-name[how]


.left5[

## How

To convert a long-formatted `data.frame` into a wide-formatted `data.frame`, you can use `pivot_longer()` function from the `tidyr` package.

```{r eval = F}
pivot_longer(
  data, 
  cols = x,
  names_to = y, 
  values_to = z 
)
```

+ `x`: list of the name of the columns to pivot into longer format
+ `y`: what the name of `x` represents
+ `z`: what the values stored in `x` represents
]

.right5[

## Example

```{r }
yield_data_wide %>% 
  pivot_longer(
    - state, 
    names_to = "year", 
    values_to = "yield"
  )
```

+ `x`: all the variables except `state`
+ `y`: "year"
+ `z`: "yield"

]

  ]

<!-- panel ends here -->


<!-- panel ends here -->

.panel[.panel-name[prefix]

You do not want `year_` in front of the year numbers in the new `year` variable? You can use the `names_prefix` option as follows:

```{r }
yield_data_wide %>%  
  pivot_longer(
    - state,
    #--- let R know what the prefix is ---#
    names_prefix = "yield_",
    names_to = "year",
    values_to = "yield"
  )
```

Notice `year` is character. Convert it to numeric using `as.numeric()` if you use is as a numeric variable.

  ] 

  <!-- panel ends here -->

]

<!-- panel set ends here -->

---

# Wide to long: multiple sets of columns

.panelset[ 

.panel[.panel-name[Prep]


.left5[

## Long

Create the following dataset in the long format;

```{r wide-data-gen-2}
yield_data_long <- data.frame(
  state = c("Kansas", "Nebraska", "Iowa", "Illinois") %>% rep(each = 2),
  year = c(2019, 2020) %>% rep(4),
  yield = c(200, 240, 210, 220, 220, 230, 190, 150),
  rainfall = c(14, 15, 15, 16, 20, 21, 24, 15)
  ) 
```

```{r echo = F}
yield_data_long
```

]

.right5[

## Wide

Convert the long dataset into the wide format: 

```{r }
yield_data_wide <- yield_data_long %>% 
  pivot_wider(
    names_from = year, 
    values_from = c(yield, rainfall)
  ) 
```

```{r echo = F}
yield_data_wide 
```

.red[Objective]: We would like to convert the wide data back to the original long data.

]

  ] 

  <!-- panel ends here -->

.panel[.panel-name[one-step?]

You cannot revert this data back to the original long-formatted data in one step. 

```{r }
yield_data_wide %>% 
  pivot_longer(
    - state,
    names_to = "year",
    values_to = "yield"
  )
```

  ]

<!-- panel ends here -->

.panel[.panel-name[separate()]

However, you take advantage of `dplyr::separate()` function, which separate a variable of type character by a user-specified separator into two variables in the dataset.

```{r eval = F}
#--- syntax ---#
separate(data, variable name, the name of variables, separator)  
```

.left5[

Before separation: 

```{r sep-demo-b, eval = F}
yield_data_wide %>% 
  #--- select only state and yield variables ---#
  pivot_longer(
    - state,
    names_to = "type_year",
    values_to = "value"
  ) 
```

.scroll-box-12[
```{r ref.label = "sep-demo-b", echo = F}
```
]

]

.right5[

After separation: 

```{r sep-demo-a, eval = F}
yield_data_wide %>% 
  #--- select only state and yield variables ---#
  pivot_longer(
    - state,
    names_to = "type_year",
    values_to = "value"
  ) %>% separate(type_year, c("type", "year"), sep = "_") #<<
```

.scroll-box-12[
```{r ref.label = "sep-demo-a", echo = F}
```
]
]
  ]

<!-- panel ends here -->

.panel[.panel-name[three-step]

After separating `type_year` to `type` and `year`, all you have to do is to apply `pivot_wider()` to have the desired long-formatted data.

```{r }
yield_data_wide %>% 
  #--- select only state and yield variables ---#
  pivot_longer(
    - state,
    names_to = "type_year",
    values_to = "value"
  ) %>% 
  separate(type_year, c("type", "year"), sep = "_") %>% 
  pivot_wider( #<<
    names_from = "type", #<<
    values_from = "value" #<<
  ) #<<
```

  ] 

  <!-- panel ends here -->

]

<!-- panel set ends here -->


---

# Wide or long?

Suppose you are interested in estimating the following statistical model:

$$corn yield = \beta_0 + \beta_1 R_{May}  + \beta_2 R_{June} + \beta_2 R_{July} + \beta_2 R_{August} + \beta_2 R_{September} + v$$

where $R$ refers to rainfall.

--

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

Then the following dataset

```{r echo = F}
data.frame(
  state = c("Kansas", "Nebraska", "Iowa", "Illinois") %>% rep(each = 2),
  year = c(2019, 2020) %>% rep(4),
  yield = c(200, 240, 210, 220, 220, 230, 190, 150)
  ) %>% 
  .[rep(1:nrow(.), each = 5), ] %>% 
  mutate(
    month = rep(c("May", "June", "July", "August", "September"), 8),
    rainfall = 20*runif(40)
  ) %>% 
  pivot_wider(names_from  = month, values_from = rainfall, names_prefix = "R_")
```

is in a <span style="color:red"> long </span> format.

---

# Wide or long?

This is ".red[too long]" for your analysis. 

```{r echo = F}
data.frame(
  state = c("Kansas", "Nebraska", "Iowa", "Illinois") %>% rep(each = 2),
  year = c(2019, 2020) %>% rep(4),
  yield = c(200, 240, 210, 220, 220, 230, 190, 150)
  ) %>% 
  .[rep(1:nrow(.), each = 5), ] %>% 
  mutate(
    month = rep(c("May", "June", "July", "August", "September"), 8),
    rainfall = 20*runif(40)
  ) %>% 
  head(15)
```

--

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

<span style="color:red"> Key point:</span> Whether a dataset is wide or long is determined based on what you are doing with the dataset.

---

# Exercises 

.panelset[ 


.panel[.panel-name[Preparation]

Run the following code so you can use the `flights` dataset.
```{r }
library(nycflights13)
```
  ]

.panel[.panel-name[Exercise 1]

+ Using `flights` data, calculate the total number of flights by carrier-month, which is in the long format

+ Reshape the data into a wide format so that you have number of flights per month in columns 

Here is the output you are supposed to get if done correctly:

```{r echo = F}
(
num_flights_wide <- flights %>% 
  group_by(carrier, month) %>% 
  summarize(num_obs = n()) %>% 
  pivot_wider(
    names_from = month,
    names_prefix = "month_",
    values_from = num_obs
  ) 
)

```

  ]

.panel[.panel-name[Exercise 2]

+ Reshape the data back into the long format so that a single columns has all the flight number values 

Here is the output you are supposed to get if done correctly:

```{r echo = F}
num_flights_wide %>% 
  pivot_longer(
    starts_with("month_"),
    names_to = "month",
    names_prefix = "month_",
    values_to = "num_flights"
  ) 
```

  ]

]


<!-- 
#/*=================================================*/
#' # Data merge
#/*=================================================*/
-->

---

class: inverse, center, middle
name: merging

# Merging multiple datasets

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---

```{r data_gen, echo = FALSE, cache = TRUE}
N <- 2
T <- 2

set.seed(582374)

price_data <- data.frame(
  price = runif(N * T, min = 2, max = 6),
  county = c(rep('Lancaster',2),rep('Douglas',2)),
  year = c(2015, 2016, 2015, 2016)
  )

yield_data <- data.frame(
  yield = runif(N * T, min = 180, max = 280),
  county = c('Lancaster','Douglas','Lancaster','Douglas'),
  year = c(2015, 2015, 2016, 2016)
  )
```

# Merging multiple datasets

.panelset[ 

.panel[.panel-name[Motivation]

<br>

It is very common that you have data stored in separate files, and you need to combine them before you conduct any statistical analysis. 

For example, if you are interested in how crop price affects the supply of crops, you want to have price and production data in a single dataset. However, it may be that price and production data are stored in two separate files.

  ]

.panel[.panel-name[Match? 1]

<br>

Now suppose, you have collected price and production data for Lancaster and Douglas County from 2015 to 2016. 

Here is what the datasets look like (these are made-up numbers).

.left5[
```{r echo = F, dependson = "data_gen"}  
select(price_data, price)
```
]

.right5[
```{r echo = F, dependson = "data_gen"}  
select(yield_data, yield)
```
]

<br>

Question: Can you merge the two?   

  ]

.panel[.panel-name[Match? 2]

<br>

Let's display one more variable from each of the datasets.

.left5[
```{r echo = F, dependson = "data_gen"}  
select(price_data, price, county)
```
]

.right5[
```{r echo = F, dependson = "data_gen"}  
select(yield_data, yield, county)
```
]

<br>
<br>

Okay, great. At least we know which price and prod belong to which county! In other words, we know which price and prod belong to **who** (or **where**). 

Question: Can you merge the two now?   

  ]

.panel[.panel-name[Match? 3]

<br>

Let's display one more variable from each of the datasets.

.left5[
```{r echo = F, dependson = "data_gen"}
select(price_data, price, county, year)
```
]

.right5[
```{r echo = F, dependson = "data_gen"}
select(yield_data, yield, county, year)
```
]

<br>
<br>

Question: Can you merge the two now?   

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

The variables that let you merge two datasets are called <span style="color:blue"> keys</span>.

What are the keys here?

  ]

.panel[.panel-name[How]

+ You can use the `left_join()` function from the `dplyr` package to merge two datasets.

+ There are different types of join functions:
  * `right_join()` (you never need to use this one)
  * `inner_join()`
  * `full_join()`
  * `semi_join()`
  * `nest_join()`

+ But, most of the time, `left_join()` is sufficient. 

+ Try to learn other functions when you encounter a case where `left_join()` is not sufficient. Do not waster your time until then. 

  ]

  <!-- panel ends here -->

.panel[.panel-name[left_join()]

**Syntax**

```{r eval = F}
#--- Syntax (NOT RUN) ---#
left_join(data_x, data_y, by = keys)  
```

**Rules to remember**

+ .blue[Rule 1]: It returns all rows from `data_x`, and all columns from `data_x` and `data_y`
+ .blue[Rule 2]: Rows in `data_x` with no match in `data_y` will have NA values in the new columns 
+ .blue[Rule 3]: If there are multiple matches between `data_x` and `data_y`, all combinations of the matches are returned

.content-box-red[Note: The order of datasets matter.]

  ]

]

---

# `left_join()` demonstration: 1 to 1

.panelset[ 

<!-- panel ends here -->

.panel[.panel-name[Data prep]

Run the following code to run the merging demonstration later:

```{r data_gen_disp, eval = F}
N <- 2
T <- 2

set.seed(582374)

price_data <- data.frame(
  price = runif(N * T, min = 2, max = 6),
  county = c(rep('Lancaster',2),rep('Douglas',2)),
  year = c(2015, 2016, 2015, 2016)
  )

yield_data <- data.frame(
  yield = runif(N * T, min = 180, max = 280),
  county = c('Lancaster','Douglas','Lancaster','Douglas'),
  year = c(2015, 2015, 2016, 2016)
  )
```

  ] 

  <!-- panel ends here -->

.panel[.panel-name[Demo 1]

The keys are `county` and `year`, so

```{r }
left_join(yield_data, price_data, by = c("county", "year"))  
```

Switching the two?

```{r }
left_join(price_data, yield_data, by = c("county", "year"))  
```

<br>

Note: In this instance, .red[which comes first does not matter] because all the individual rows in `yield_data` (left data) have exactly one match in `price_data` (right data) without fail, and vice versa.  

  ] 

  <!-- panel ends here -->

.panel[.panel-name[Data prep 2]

Let's expand the `yield_data`:

```{r }
yield_data <- data.frame(
  yield = runif(6, min = 180, max = 280),
  county = c("Lancaster", "Douglas", "Chase", "Lancaster", "Douglas", "Chase"),
  year = c(2015, 2015, 2015, 2016, 2016, 2016)
  ) 

yield_data
```

  ] 

  <!-- panel ends here -->

.panel[.panel-name[Demo 2]

**`yield_data` on the right**:

```{r }
left_join(price_data, yield_data, by = c("county", "year"))  
```

**`yield_data` on the left**:

```{r }
left_join(yield_data, price_data, by = c("county", "year"))  
```

.red[Remember?]

+ .blue[Rule 1]: `left_join()` returns all rows from `data_x`, and all columns from `data_x` and `data_y`
+ .blue[Rule 2]: Rows in `data_x` with no match in `data_y` will have NA values in the new columns 
  ] 

  <!-- panel ends here -->

.panel[.panel-name[keeping all rows]

We saw in the previous slide having `price_data` (as `data_x`) and `yield_data` as (`data_y`), `left_join()` discarded rows in `yield_data` (`data_y`).

If you would like to keep unmatched rows in `data_y`, you can use `full_join()`.

```{r }
full_join(price_data, yield_data, by = c("county", "year"))  
```

  ] 

  <!-- panel ends here -->

]
<!-- panel set ends here -->

---

# `left_join` demonstration: 1 to m


.panelset[ 

.panel[.panel-name[Data prep]

Let's create a weather dataset where you have more than one observations per county-year:

```{r }
weather_data <- data.frame(
  rainfall = 20 * runif(12),
  county = rep(c("Lancaster", "Douglas", "Chase"), each = 4),
  year = rep(c(2015, 2016), each = 2) %>% rep(3),
  month = rep(c(4, 5), 6)
  ) 

weather_data
```

  ]

<!-- panel ends here -->

.panel[.panel-name[Demo]

```{r }
left_join(yield_data, weather_data, by = c("county", "year"))  
```

.red[Remember?] 

+ .blue[Rule 3]: If there are multiple matches between `data_x` and `data_y`, all combinations of the matches are returned

  ]

  <!-- panel ends here -->

]

<!-- panel set ends here -->


---

# Exercises

.panelset[ 

.panel[.panel-name[Preparation]

Create the following datasets and take a look at them to understand what's in them:

**Flights in January**:

```{r }
flights_Jan <- flights %>% 
  filter(month == 1) %>% 
  select(month, day, dep_delay, origin)
```

**daily temperature in January**:

```{r }
daily_temp_Jan <- weather %>% 
  filter(month == 1) %>% 
  group_by(origin, month, day) %>% 
  summarize(temp = mean(temp))
```

**hourly temperature in January**:

```{r }
hourly_temp_Jan <- weather %>% 
  filter(month == 1) %>% 
  select(origin, month, day, hour, temp)
```

  ] 

.panel[.panel-name[Exercise 1]

+ You are interested in learning the impact of **daily** temperature on departure delay for the flights in January. To do so, you need to have them in a single dataset.  

+ Is this going to be a 1-to-1 matching or 1-to-m matching?

+ Merge `daily_temp_Jan` to `flights_Jan` using `left_join()`

<br>

Here is the output you are supposed to get if done correctly:

```{r echo = F}
left_join(flights_Jan, daily_temp_Jan, by = c("origin", "month", "day"))
```

  ]

.panel[.panel-name[Exercise 2]

+ You are interested in learning the impact of **hourly** temperature on departure delay for the flights in January. To do so, you need to have them in a single dataset.  

+ Is this going to be a 1-to-1 matching or 1-to-m matching?

+ Merge `hourly_temp_Jan` to `flights_Jan` using `left_join()`

<br>

Here is the output you are supposed to get if done correctly:

```{r echo = F}
left_join(flights_Jan, hourly_temp_Jan, by = c("origin", "month", "day"))
```

Each flight (single row in `flights_Jan`) now has multiple rows. Was this expected? 

  ]

]

